### 2021-02-11 language-derivatives.tex
copy a few starter files

\documentclass[hidelinks]{article}  % fleqn,12pt
\usepackage[margin=1in]{geometry}  % 0.12in, 0.9in, 1in

\usepackage{catchfilebetweentags}
\usepackage[useregional]{datetime2}

%% \usepackage{balance}  % even out final two-column page

\RequirePackage{newunicodechar, amssymb, stmaryrd, unicode-math, setspace}

\input{commands}
\input{unicode}
\input{macros}

\usepackage[links]{agda}% references

%% Switching to \textsf for AgdaFunction messes up vertical alignment.
%% \newcommand{\AgdaFontStyle}[1]{\textsf{#1}}

\renewcommand{\AgdaFontStyle}[1]{\text{#1}}

%% \renewcommand{\AgdaFunction}[1]
%%     {\AgdaNoSpaceMath{\text{\textcolor{AgdaFunction}{\AgdaFormat{#1}{\AgdaLink{#1}}}}}}
%% \renewcommand{\AgdaRecord}[1]
%%     {\AgdaNoSpaceMath{\text{\textcolor{AgdaRecord}{\AgdaFormat{#1}{\AgdaLink{#1}}}}}}

\author{Conal Elliott}

\begin{document}

%% \nc\tit{Working Title}
\nc\tit{Symbolic and Automatic Differentiation of Languages}

\title{\tit}
\date{Early draft of \DTMnow}

\maketitle

\begin{abstract}
Formal languages are usually defined in terms of set theory.
Choosing type theory instead gives us languages as type-level predicates over strings.
Applying a language to a string yields a type whose elements are language membership proofs describing \emph{how} a string parses in the language.
The usual building blocks of languages (including union, contatenation, and Kleene closure) have precise and compelling specifications uncomplicated by operational strategies and are easily generalized to a few general domain-transforming and codomain-transforming operations on predicates.

A simple type isomorphism property captures the essential idea behind language ``differentiation'' and ``integration'' as used for recognizing languages and leads to a collection of lemmas about type-level languages (and indeed functions from lists to any type).
These lemmas form the backbone of two dual implementations of language recognition---(inductive) regular expressions and (coinductive) tries---each containing almost exactly the same code but in dual arrangements.
The language lemmas form most of the implementation in both cases, while the representation and primitive operations trade places.
The regular expression version corresponds to symbolic differentiation, while the trie version corresponds to automatic differentiation.

The relatively easy-to-prove properties of type-level languages transfer almost effortlessly to the decidable implementations.
In particular, despite the inductive and coinductive nature of regular expressions and tries respectively, we need neither inductive nor coinductive/bisumulation arguments to prove algebraic properties.
\end{abstract}

\renewcommand\AgdaTarget[1]{}

%% \input{NewIntroduction}
\input{Introduction}
%% \input{Generation}
\input{Predicate}
%% \input{AltDec}
%% \input{AltCalc}
\input{Predicate/Algebra}
\input{Nullability}
\input{Decidability}
\input{Calculus}
% \input{NonTrie}
\input{Regex}
\input{Trie}
\input{ProvingProperties}

%% \input{BeyondPredicates}
%% \input{Predicate/Examples}
%% \input{Relation}

\workingHere
\note{To do: relate regular expressions and tries to symbolic and automatic differentiation; reflections, related work, future work.}

%% \definecolor{mylinkcolor}{rgb}{0,0.4,0}
%% \hypersetup{
%%   linkcolor  = mylinkcolor,
%%   citecolor  = mylinkcolor,
%%   urlcolor   = mylinkcolor,
%%   colorlinks = true
%% }

\bibliography{bib}

\end{document}

### 2021-02-11 paper.tex
rebuilding paper and talk with shared Agda sources

--- 2021-02-11 language-derivatives.tex
@@
 
-\usepackage[links]{agda}% references
+\usepackage{agda}% references
 
@@
 
-\renewcommand{\AgdaFontStyle}[1]{\text{#1}}
+%% \renewcommand{\AgdaFontStyle}[1]{\text{#1}}
 
@@
 %% \nc\tit{Working Title}
-\nc\tit{Symbolic and Automatic Differentiation of Languages}
+\nc\tit{%Functional Pearl:
+Symbolic and Automatic Differentiation of Languages}
 
@@
 A simple type isomorphism property captures the essential idea behind language ``differentiation'' and ``integration'' as used for recognizing languages and leads to a collection of lemmas about type-level languages (and indeed functions from lists to any type).
-These lemmas form the backbone of two dual implementations of language recognition---(inductive) regular expressions and (coinductive) tries---each containing almost exactly the same code but in dual arrangements.
+These lemmas form the backbone of two dual implementations of language recognition---(inductive) regular expressions and (coinductive) tries---each containing the same code but in dual arrangements.
 The language lemmas form most of the implementation in both cases, while the representation and primitive operations trade places.
@@
 
-%% \input{NewIntroduction}
-\input{Introduction}
-%% \input{Generation}
-\input{Predicate}
-%% \input{AltDec}
-%% \input{AltCalc}
-\input{Predicate/Algebra}
-\input{Nullability}
-\input{Decidability}
-\input{Calculus}
-% \input{NonTrie}
-\input{Regex}
-\input{Trie}
-\input{ProvingProperties}
+\sectionl{Specifying Languages}
 
-%% \input{BeyondPredicates}
-%% \input{Predicate/Examples}
-%% \input{Relation}
+Languages are usually formalized either set-theoretically as a set of strings \needcite{} or operationally as a parser \needcite{}.
+Alternatively, we can use type theory, so that a language is a type-level predicate on ``strings'' (lists of an arbitrary type \AB{A} of ``characters''):
+\Language{Lang}
+\figrefdef{Lang-ops}{Language operations}{\Language{Lang-ops}} gives definitions of the usual language operations, including language union and intersection ({\AB{P} \AF ∪ \AB{Q}} and {\AB{P} \AF ∩ \AB{Q}}) with their identities (the empty language \AF ∅ and universal language \AF 𝒰), language concatenation ({\AB{P} \AF ⋆ \AB{Q}}) and its identity (\AF 𝟏, containing only the empty string), single-character languages ({\AF ` \AB{c}}), Kleene star ({\AB{P} \AF ☆}), and ``scalar multiplication'' (\mbox{\AB{s} \AF · \AB{P}}, which will prove useful later).
 
-\workingHere
-\note{To do: relate regular expressions and tries to symbolic and automatic differentiation; reflections, related work, future work.}
+\note{Explain the types used here and their logical interpretation under the Curry-Howard isomorphism.}
 
-%% \definecolor{mylinkcolor}{rgb}{0,0.4,0}
-%% \hypersetup{
-%%   linkcolor  = mylinkcolor,
-%%   citecolor  = mylinkcolor,
-%%   urlcolor   = mylinkcolor,
-%%   colorlinks = true
-%% }
+Note that for a language \AB{P} and string \AB{w}, {\AB{P} \AB{w}} is the type of \emph{proofs} that {\AB{w} ∈ \AB{P}}, in other words \emph{explanations} of membership, or \emph{parsing}.
+(If the type {\AB{P} \AB{w}} is uninhabited, then {\AB{w} ∉ \AB{P}}.)
+For instance, a proof of {\AB{w} ∈ \AB{P} \AF ∪ \AB{Q}} contains a proof of {\AB{w} ∈ \AB{P}} or of {\AB{w} \AF ∈ \AB{Q}}, \emph{and} the knowledge of which one was chosen.
+Likewise, a proof of {w ∈ \AB{P} \AF ⋆ \AB{Q}} (language concatenation) includes the choice of strings \AB{u} and \AB{v}, a proof that {\AB{u} \AF ⊙ \AB{v} \AD ≡ \AB{w}}, and proofs that {\AB{u} ∈ \AB{P}} and {\AB{v} ∈ \AB{P}}.
+
+The use of type-level predicates makes for simple, direct specification, including the use of existential quantification.
+As described in \secref{Language Algebra}, it's also fairly easy to prove algebraic properties of predicates\out{, including the semimodule and semiring laws for both predicate semirings}.
+These definitions do not, however, give us decidable parsing, because general type inhabitation amounts to theorem proving.
+
+\sectionl{Decomposing Languages}
+
+In order to convert languages into parsers, it will help to decompose languages---and indeed any function from lists---into a regular form.
+Since lists are defined by an algebraic data type, we can decompose languages according to constructor.
 

### 2021-02-12 paper.tex
\source & \agda macros; Reflections
building up the paper

@@
-\documentclass[hidelinks]{article}  % fleqn,12pt
-\usepackage[margin=1in]{geometry}  % 0.12in, 0.9in, 1in
+\documentclass[hidelinks,12pt]{article}  % fleqn,12pt
+\usepackage[margin=0.1in]{geometry}  % 0.12in, 0.9in, 1in
 
@@
 
-\RequirePackage{newunicodechar, amssymb, stmaryrd, unicode-math, setspace}
+\RequirePackage{newunicodechar, amssymb, stmaryrd, unicode-math, setspace, comment}
 
@@
 
-\renewcommand\AgdaTarget[1]{}
+\rnc\AgdaTarget[1]{}
 
@@
 
+\rnc\source{Language}
+
 Languages are usually formalized either set-theoretically as a set of strings \needcite{} or operationally as a parser \needcite{}.
 Alternatively, we can use type theory, so that a language is a type-level predicate on ``strings'' (lists of an arbitrary type \AB{A} of ``characters''):
-\Language{Lang}
-\figrefdef{Lang-ops}{Language operations}{\Language{Lang-ops}} gives definitions of the usual language operations, including language union and intersection ({\AB{P} \AF ∪ \AB{Q}} and {\AB{P} \AF ∩ \AB{Q}}) with their identities (the empty language \AF ∅ and universal language \AF 𝒰), language concatenation ({\AB{P} \AF ⋆ \AB{Q}}) and its identity (\AF 𝟏, containing only the empty string), single-character languages ({\AF ` \AB{c}}), Kleene star ({\AB{P} \AF ☆}), and ``scalar multiplication'' (\mbox{\AB{s} \AF · \AB{P}}, which will prove useful later).
+\agda{Lang}
+\figrefdef{Lang-ops}{Language operations}{\agda{Lang-ops}} gives definitions of the usual language operations, including language union and intersection ({\AB{P} \AF ∪ \AB{Q}} and {\AB{P} \AF ∩ \AB{Q}}) with their identities (the empty language \AF ∅ and universal language \AF 𝒰), language concatenation ({\AB{P} \AF ⋆ \AB{Q}}) and its identity (\AF 𝟏, containing only the empty string), single-character languages ({\AF ` \AB{c}}), Kleene star ({\AB{P} \AF ☆}), and ``scalar multiplication'' (\mbox{\AB{s} \AF · \AB{P}}, which will prove useful later).
 
@@
 
+\rnc\source{Calculus}
+
 In order to convert languages into parsers, it will help to decompose languages---and indeed any function from lists---into a regular form.
-Since lists are defined by an algebraic data type, we can decompose languages according to constructor.
+Since lists are defined by an algebraic data type, we can decompose languages according to constructor, as follows:
+\agda{νδ}
+Each use of \AF{δ} takes us one step closer to reducing general language membership to nullability.
+In other words,\footnote{Repeated application is expressed as a left fold\out{ \stdlibCitep{Data.List}}:
+\agda{foldl}
+}%
+\agda{ν∘foldlδ}
+
+This simple observation is the semantic heart of the syntactic technique of \emph{derivatives of regular expressions} as used for efficient recognition of regular languages \citep{Brzozowski64}, later extended to parsing general context-free languages \citep{Might2010YaccID}.
+Lemma \AF{ν∘foldlδ} liberates this technique from the assumption that languages are represented \emph{symbolically}, by some form of grammar (e.g., regular or context-free), inviting other representations as we will see in \secref{Automatic}.
+
+Lemma \AF{ν∘foldlδ} relates to some other powerful perspectives and principles as well:
+\begin{itemize}
+
+\item 
+\emph{Automata theory}: {\AF{ν} \AF ∘ \AF{foldl} \AF{δ} \AB{P}} is the execution of a state machine.
+Each state is a language, with \AB{P} being the initial state; \AF{δ} is the state transition function; and \AB{ν} is the set of accepting states \needcite{}.
+Lemma \AF{ν∘foldlδ} captures the correctness of this state machine as a recognizer for the language \AB{P}.
+
+\item
+\emph{Incremental computation}: {\AF{ν} \AF ∘ \AF{foldl} \AF{δ} \AB{P}} computes \AB{P} by transforming successive incremental modifications to the empty string (each in the form of {\AB{a} \AIC ∷\_} for a list element \AB{a}) into incremental modifications to the membership question for the related language {\AF{δ} \AB{P} \AB{a}} \needcite{}.
+Lemma \AF{ν∘foldlδ} again captures the correctness of this incremental implementation of \AB{P}.
+
+\item 
+\emph{Calculus}: \AB{δ} is differentiation; {\AF{ν} \AF ∘ \AF{foldl} \AF{δ} \AB{P}} is integration; and lemma \AF{ν∘foldlδ} is the second fundamental theorem of calculus, which says that a function can be recovered by integrating its derivative \needcite{}.
+
+\end{itemize}
+
+Given the definitions of \AF{ν} and \AF{δ} above and of the language operations in \secref{Specifying Languages}, one can prove properties about how they relate, as shown in
+\figrefdef{nu-delta-lemmas}{Properties of \AF{ν} and \AF{δ} for language operations}{\agda{νδ-lemmas}}.
+The correct-by-construction parsing algorithms in \secreftwo{Symbolic Differentiation}{Automatic Differentiation} are corollaries of these properties.
+
+There are three relations involved in \figref{nu-delta-lemmas}: propositional equality (``≡''), type isomorphism (``↔'') and extensional isomorphism (``⟷'').
+Type isomorphism relates types whose inhabitants/proofs are in one-to-one correspondence.
+\emph{Extensional} (or ``pointwise'') isomorphism relates predicates isomorphic on every argument:
+\agda{ext-iso}
+
+As with all lemmas in this paper shown with signatures only, full proofs are in the paper's source code and are formally verified by the Agda compiler.
+The equalities in \figref{nu-delta-lemmas} are all proved automatically by normalization (i.e., their proofs are simply \AIC{refl}), while the other relations require a bit more work.
+
+\sectionl{Decidability}
+
+\rnc\source{Decidability}
+
+For an effective implementation, we will have to bridge the gap between a type (of membership proofs) and its decidable inhabitation.
+Fortunately, there is a convenient and compositional way to do so.
+For any type \AB{X}, the type {\AF{Dec} \AB{X}} contains proof of \AB{X} or a proof of {\AF ¬ \AB{X}} (defined as usual to mean {\AB{X} \AS → \AD{⊥}}):\footnote{This simple \AF{Dec} definition been superceded by a more efficient but more complex version \stdlibCitep{Relation.Nullary}.}
+
+\agda{Dec}
+For compositionality, we will also want higher-order counterparts lifting decidability of types to decidability of constructions on those types:
+\agda{Decs}
 

### 2021-02-12 paper.tex
Decidability

@@
 Fortunately, there is a convenient and compositional way to do so.
-For any type \AB{X}, the type {\AF{Dec} \AB{X}} contains proof of \AB{X} or a proof of {\AF ¬ \AB{X}} (defined as usual to mean {\AB{X} \AS → \AD{⊥}}):\footnote{This simple \AF{Dec} definition been superceded by a more efficient but more complex version \stdlibCitep{Relation.Nullary}.}
-
+For any type \AB{A}, the type {\AF{Dec} \AB{A}} contains proof of \AB{A} or a proof of {\AF ¬ \AB{A}} (defined as usual to mean {\AB{A} \AS → \AD{⊥}}):\footnote{This simple \AF{Dec} definition been superceded by a more efficient but more complex version \stdlibCitep{Relation.Nullary}.}
 \agda{Dec}
@@
 \agda{Decs}
+A \emph{predicate} \AB{P} is decidable when (the type) {\AB{P} \AB{x}} is decidable for all \AB{x} \stdlibCitep{Relation.Unary}:
+\agda{Decidable}
+
+Now we can formulate the problem of decidable language recognition: given a language \AB{P}, construct a term of type {\AF{Decidable} \AB{P}}.
+Since this transformation cannot be fully automated, we will instead look for decidable building blocks that mirror the predicate vocabulary defined in \secref{Specifying Languages}.
+
+Type isomorphisms (such as those in \figref{nu-delta-lemmas}) play an important role in decidability, namely that isomorphic types or predicates are equivalently decidable:\footnote{In fact, logical equivalence suffices for the results in this paper, but isomorphism enables variations such as generating all parses/proofs rather than just one.}
+\agda{isomorphisms}
 

### 2021-02-12 paper.tex
symbolic differentiation

@@
 %%     {\AgdaNoSpaceMath{\text{\textcolor{AgdaRecord}{\AgdaFormat{#1}{\AgdaLink{#1}}}}}}
+
+%% \rnc\AgdaTarget[1]{}
 
@@
 
-\rnc\AgdaTarget[1]{}
-
 \sectionl{Specifying Languages}
-
 \rnc\source{Language}
@@
 The use of type-level predicates makes for simple, direct specification, including the use of existential quantification.
-As described in \secref{Language Algebra}, it's also fairly easy to prove algebraic properties of predicates\out{, including the semimodule and semiring laws for both predicate semirings}.
+As described in \secref{Proving Properties}, it's also fairly easy to prove algebraic properties of predicates\out{, including the semimodule and semiring laws for both predicate semirings}.
 These definitions do not, however, give us decidable parsing, because general type inhabitation amounts to theorem proving.
@@
 \sectionl{Decomposing Languages}
-
 \rnc\source{Calculus}
@@
 \sectionl{Decidability}
-
 \rnc\source{Decidability}
@@
 \agda{Dec}
-For compositionality, we will also want higher-order counterparts lifting decidability of types to decidability of constructions on those types:
-\agda{Decs}
-A \emph{predicate} \AB{P} is decidable when (the type) {\AB{P} \AB{x}} is decidable for all \AB{x} \stdlibCitep{Relation.Unary}:
+For compositionality, we will use higher-order counterparts lifting decidability of types to decidability of constructions on those types, as shown in \figrefdef{compositional-dec}{Compositional decidability}{\agda{compositional-dec}}.
+
+A \emph{language} (or any predicate) \AB{P} is decidable when (the type) {\AB{P} \AB{x}} is decidable for all \AB{x} \stdlibCitep{Relation.Unary}:
 \agda{Decidable}
-
 Now we can formulate the problem of decidable language recognition: given a language \AB{P}, construct a term of type {\AF{Decidable} \AB{P}}.
@@
 
-Type isomorphisms (such as those in \figref{nu-delta-lemmas}) play an important role in decidability, namely that isomorphic types or predicates are equivalently decidable:\footnote{In fact, logical equivalence suffices for the results in this paper, but isomorphism enables variations such as generating all parses/proofs rather than just one.}
+Type isomorphisms (such as those in \figref{nu-delta-lemmas}) play an important role in decidability, namely that isomorphic types or predicates are equivalently decidable:
 \agda{isomorphisms}
+One direction of the isomorphism proves \AB{B} from \AB{A}, while the other proves {\AF ¬ B} from {\AF ¬ A}.
+In fact, logical \emph{equivalence} suffices for the results in this paper, but the stronger condition of isomorphism enables variations such as generating all parses/proofs rather than just one.
+
+\sectionl{Reflections}
+
+The lemmas in \figref{nu-delta-lemmas} tell us how to decompose languages defined in terms of the vocabulary from \figref{Lang-ops}, and the definitions in \figref{compositional-dec} tell us how compute inhabitation of the resulting types, resulting in decidable parsing.
+These lemmas and definitions cannot be applied \emph{automatically} in their present form, however, because languages are functions, as are \AF{ν} and \AF{δ}, and so are not subject to pattern-matching.
+An automatic solution would need some form of \emph{reflection} of language operations or \AF{ν} and \AF{δ} as inspectable data.
+
+This situation is exactly as in differential calculus, since differentiation in that setting is also defined on functions rather than on symbolic representations.
+Fortunately, there are two standard solutions for \emph{computable} differentiation, going by the names of ``symbolic'' and ``automatic'' differentiation.
+In the former, functions are represented symbolically in some suitable vocabulary, and pattern-matching is used to apply rules of differentiation to those representations.
+The latter solution involves re-interpreting the construction vocabulary for functions to construct not just the usual functions, but also their derivatives \needcite{}.
+The latter is often much more efficient than the former, since it easily avoids a good deal of redundant computation due to significant amounts of duplicated work between function ``primals'' and their derivatives.
+
+\secreftwo{Symbolic Differentiation}{Automatic Differentiation} apply the symbolic and automatic strategies respectively to languages.
+In both cases, the algorithms are proved correct with minimal effort as corollaries of the lemmas in \figref{nu-delta-lemmas}.
+
+\sectionl{Symbolic Differentiation}
+\rnc\source{Symbolic}
+
+In the ``symbolic'' style, we reflect the language-building vocabulary of \figref{Lang-ops} into an inductive data type, while re-defining \AF{ν} and \AF{δ} as functions on that data type.
+For convenience, use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the prefix ``{◬.}'' (module-specific names).
+The result is shown in \figrefdef{symbolic-api}{Regular expressions}{\agda{api}} along with a semantic function {\AF{⟦\_⟧}} that converts a symbolic language representation to decidable form.
+
+Decidable recognition (via {\AF{⟦\_⟧}}) relies only on \AF{ν} and \AF{δ}, which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}.
+These definitions are meant to be read in ``column-major'' order, i.e., the left column is one definition (of \AF{ν}), and the right column is another (of \AF{δ}).
+The \AF{Lang} representation is inductive, and definitions are by pattern matching.
+Correctness is guaranteed by the types of \AF{ν} and \AF{δ}, so the correctness proof here is accomplished by type-checking.
+
+\sectionl{Automatic Differentiation}
+
+\sectionl{Proving Properties}
 

### 2021-02-13 paper.tex
symbolic and automatic differentiation

@@
 \documentclass[hidelinks,12pt]{article}  % fleqn,12pt
-\usepackage[margin=0.1in]{geometry}  % 0.12in, 0.9in, 1in
+\usepackage[margin=1in]{geometry}  % 0.12in, 0.9in, 1in
 
@@
 
-In the ``symbolic'' style, we reflect the language-building vocabulary of \figref{Lang-ops} into an inductive data type, while re-defining \AF{ν} and \AF{δ} as functions on that data type.
+In the ``symbolic differentiation'' style, we reflect the language-building vocabulary of \figref{Lang-ops} into an inductive data type (of modestly extended regular expressions), while re-defining \AF{ν} and \AF{δ} as functions on that data type.
 For convenience, use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the prefix ``{◬.}'' (module-specific names).
-The result is shown in \figrefdef{symbolic-api}{Regular expressions}{\agda{api}} along with a semantic function {\AF{⟦\_⟧}} that converts a symbolic language representation to decidable form.
+The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\agda{api}} along with a semantic function {\AF{⟦\_⟧}} that converts a symbolic language representation to decidable form.
 
 Decidable recognition (via {\AF{⟦\_⟧}}) relies only on \AF{ν} and \AF{δ}, which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}.
-These definitions are meant to be read in ``column-major'' order, i.e., the left column is one definition (of \AF{ν}), and the right column is another (of \AF{δ}).
+These definitions are meant to be read in ``column-major'' order, i.e., each column is one function definition.
 The \AF{Lang} representation is inductive, and definitions are by pattern matching.
@@
 \sectionl{Automatic Differentiation}
+\rnc\source{Automatic}
+
+\agda{rules}
+
+\secref{Symbolic Differentiation} embodies one choice of reflecting functions into a data representation.
+Now consider the dual strategy of ``automatic differentiation''.
+This time, we reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type (of tries \needcite{}), while redefining the language-building vocabulary as functions on that data type.
+Again, we will use the same names as in \secref{Specifying Languages}, referring to the original versions via the prefix ``{◬.}''.
+The result is shown in \figrefdef{automatic-api}{Tries (coinductive)}{\agda{api}}.
+
+Decidable recognition again relies only on \AF{ν} and \AF{δ}, which are defined in \figrefdef{automatic-defs}{Automatic differentiation (row-major/copatterns)}{\agda{defs}}.
+These definitions are meant to be read in ``row-major'' order, i.e., each row is one function definition.
+The \AF{Lang} representation is coinductive, and definitions are by \emph{copattern} matching \citep{AbelPientka2016}.
+Note that the definitions are syntactically identical to those in \figref{symbolic-defs} but are organized dually.
+(The compiler-generated syntax coloring differs to reflect the changed interpretation of the definitions.)
+
+The correctness proof is \emph{almost} accomplished by type-checking, but there is a technical problem.
+The \AF{δ} clause for {\AB{p} \AF ⋆ \AB{q}} does not satisfy Agda's termination checker, which cannot see that the {\AF{δ} \AB{p} \AB{a}} argument in the recursive use of \AF{\_⋆\_} is in some sense smaller than \AB{p}.
+Fortunately, this issue was already solved by Andreas \citet{Abel2016}, also in the setting of trie-based language recognition, by using sized types \citep{Abel2008, AbelPientka2016}.
+\rnc\source{SizedAutomatic}
+This solution needs only give \AF{Lang} (now tries) an index {\AB{i} \AK : \APT{Size}} corresponding to the maximum depth to which a trie can be searched, or equivalently the longest string that can be matched\out{ (via {\AF{⟦\_⟧}})}.
+In practice, we will work with arbitrarily deep tries, i.e., ones having index \APo{∞}, as in the type of {\AF{⟦\_⟧}}.
+The modified representation is shown in \figrefdef{sized automatic-api}{Sized tries (coinductive)}{\agda{api}}.
+
+Decidable recognition is defined exactly as with unsized tries (\figref{automatic-defs}), but this time the compiler successfully proves \emph{total} correctness (including termination).
 

### 2021-02-13 paper.tex
fonts

@@
 \documentclass[hidelinks,12pt]{article}  % fleqn,12pt
-\usepackage[margin=1in]{geometry}  % 0.12in, 0.9in, 1in
+\usepackage[margin=0.2in]{geometry}  % 0.12in, 0.9in, 1in
 
@@
 
+\usepackage{libertine}  %% [tt=false]
 \usepackage{agda}% references
@@
 
-A simple type isomorphism property captures the essential idea behind language ``differentiation'' and ``integration'' as used for recognizing languages and leads to a collection of lemmas about type-level languages (and indeed functions from lists to any type).
+A simple characterization of languages (and indeed functions from lists to any type) captures the essential idea behind language ``differentiation'' and ``integration'' as used for recognizing languages and leads to a collection of lemmas about type-level languages.
 These lemmas form the backbone of two dual implementations of language recognition---(inductive) regular expressions and (coinductive) tries---each containing the same code but in dual arrangements.
@@
 Each use of \AF{δ} takes us one step closer to reducing general language membership to nullability.
-In other words,\footnote{Repeated application is expressed as a left fold\out{ \stdlibCitep{Data.List}}:
+In other words,\footnote{Repeated application is expressed as a standard left fold\out{ \stdlibCitep{Data.List}}:
 \agda{foldl}
@@
 This simple observation is the semantic heart of the syntactic technique of \emph{derivatives of regular expressions} as used for efficient recognition of regular languages \citep{Brzozowski64}, later extended to parsing general context-free languages \citep{Might2010YaccID}.
-Lemma \AF{ν∘foldlδ} liberates this technique from the assumption that languages are represented \emph{symbolically}, by some form of grammar (e.g., regular or context-free), inviting other representations as we will see in \secref{Automatic}.
+Lemma \AF{ν∘foldlδ} liberates this technique from the assumption that languages are represented \emph{symbolically}, by some form of grammar (e.g., regular or context-free), inviting other representations as we will see in \secref{Automatic Differentiation}.
 
@@
 
-There are three relations involved in \figref{nu-delta-lemmas}: propositional equality (``≡''), type isomorphism (``↔'') and extensional isomorphism (``⟷'').
-Type isomorphism relates types whose inhabitants/proofs are in one-to-one correspondence.
+There are three relations involved in \figref{nu-delta-lemmas}: propositional equality (``≡''), (type) isomorphism (``↔'') and extensional isomorphism (``⟷'').
+Isomorphism relates types (propositions) whose inhabitants (proofs) are in one-to-one correspondence.
 \emph{Extensional} (or ``pointwise'') isomorphism relates predicates isomorphic on every argument:
-\agda{ext-iso}
+\ExecuteMetaData[Inverses.tex]{ext-iso}
 
@@
 \agda{Dec}
-For compositionality, we will use higher-order counterparts lifting decidability of types to decidability of constructions on those types, as shown in \figrefdef{compositional-dec}{Compositional decidability}{\agda{compositional-dec}}.
+For compositionality, we will use a few operations that lift decidability of types to decidability of constructions on those types, as shown in \figrefdef{compositional-dec}{Compositional decidability}{\agda{compositional-dec}}.\notefoot{Explanatory footnote goes here.}
 
-A \emph{language} (or any predicate) \AB{P} is decidable when (the type) {\AB{P} \AB{x}} is decidable for all \AB{x} \stdlibCitep{Relation.Unary}:
+A \emph{language} (or any predicate) \AB{P} is ``decidable'' when (the type) {\AB{P} \AB{x}} is decidable for all \AB{x}\out{ \stdlibCitep{Relation.Unary}}:
 \agda{Decidable}
-Now we can formulate the problem of decidable language recognition: given a language \AB{P}, construct a term of type {\AF{Decidable} \AB{P}}.
+With these definitions, we can formulate the problem of decidable language recognition: given a language \AB{P}, construct a term of type {\AF{Decidable} \AB{P}}.
 Since this transformation cannot be fully automated, we will instead look for decidable building blocks that mirror the predicate vocabulary defined in \secref{Specifying Languages}.
@@
 
-The lemmas in \figref{nu-delta-lemmas} tell us how to decompose languages defined in terms of the vocabulary from \figref{Lang-ops}, and the definitions in \figref{compositional-dec} tell us how compute inhabitation of the resulting types, resulting in decidable parsing.
+The lemmas in \figref{nu-delta-lemmas} tell us how to decompose languages defined in terms of the vocabulary from \figref{Lang-ops}, while the definitions in \figref{compositional-dec} tell us how compute inhabitation of the resulting types, resulting in decidable parsing.
 These lemmas and definitions cannot be applied \emph{automatically} in their present form, however, because languages are functions, as are \AF{ν} and \AF{δ}, and so are not subject to pattern-matching.
@@
 This situation is exactly as in differential calculus, since differentiation in that setting is also defined on functions rather than on symbolic representations.
-Fortunately, there are two standard solutions for \emph{computable} differentiation, going by the names of ``symbolic'' and ``automatic'' differentiation.
+Fortunately, there are two standard solutions for \emph{computable} differentiation, going by the names of ``symbolic'' and ``automatic'' differentiation.\out{\footnote{There is also an incorrect (approximate) variation often referred to as ``numeric'' differentiation.}}
 In the former, functions are represented symbolically in some suitable vocabulary, and pattern-matching is used to apply rules of differentiation to those representations.
-The latter solution involves re-interpreting the construction vocabulary for functions to construct not just the usual functions, but also their derivatives \needcite{}.
+The latter solution involves re-interpreting the vocabulary of functions to construct not just the usual functions, but also their derivatives \needcite{}.
 The latter is often much more efficient than the former, since it easily avoids a good deal of redundant computation due to significant amounts of duplicated work between function ``primals'' and their derivatives.
@@
 
-In the ``symbolic differentiation'' style, we reflect the language-building vocabulary of \figref{Lang-ops} into an inductive data type (of modestly extended regular expressions), while re-defining \AF{ν} and \AF{δ} as functions on that data type.
+The ``symbolic differentiation'' style reflects language-building vocabulary (\figref{Lang-ops}) into an inductive data type (of modestly extended regular expressions), while re-defining \AF{ν} and \AF{δ} as functions that pattern-match on that data type.
 For convenience, use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the prefix ``{◬.}'' (module-specific names).
@@
 
-Decidable recognition (via {\AF{⟦\_⟧}}) relies only on \AF{ν} and \AF{δ}, which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}.
+Decidable recognition (via {\AF{⟦\_⟧}}) relies only on \AF{ν} and \AF{δ}, which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\notefoot{Describe this derivation.}
 These definitions are meant to be read in ``column-major'' order, i.e., each column is one function definition.
-The \AF{Lang} representation is inductive, and definitions are by pattern matching.
-Correctness is guaranteed by the types of \AF{ν} and \AF{δ}, so the correctness proof here is accomplished by type-checking.
+Correctness is guaranteed by the types of \AF{ν} and \AF{δ} and so is proved automatically by type-checking \figref{symbolic-defs}.
 
@@
 Decidable recognition again relies only on \AF{ν} and \AF{δ}, which are defined in \figrefdef{automatic-defs}{Automatic differentiation (row-major/copatterns)}{\agda{defs}}.
-These definitions are meant to be read in ``row-major'' order, i.e., each row is one function definition.
+These definitions are meant to be read in ``row-major'' order, i.e., each row is one definition.
 The \AF{Lang} representation is coinductive, and definitions are by \emph{copattern} matching \citep{AbelPientka2016}.
@@
 The correctness proof is \emph{almost} accomplished by type-checking, but there is a technical problem.
-The \AF{δ} clause for {\AB{p} \AF ⋆ \AB{q}} does not satisfy Agda's termination checker, which cannot see that the {\AF{δ} \AB{p} \AB{a}} argument in the recursive use of \AF{\_⋆\_} is in some sense smaller than \AB{p}.
+The \AF{δ} clause for {\AB{p} \AF ⋆ \AB{q}} does not satisfy Agda's termination checker, which cannot see that the argument {\AF{δ} \AB{p} \AB{a}} in the recursive use of \AF{\_⋆\_} is in some sense smaller than \AB{p}.
 Fortunately, this issue was already solved by Andreas \citet{Abel2016}, also in the setting of trie-based language recognition, by using sized types \citep{Abel2008, AbelPientka2016}.
 \rnc\source{SizedAutomatic}
-This solution needs only give \AF{Lang} (now tries) an index {\AB{i} \AK : \APT{Size}} corresponding to the maximum depth to which a trie can be searched, or equivalently the longest string that can be matched\out{ (via {\AF{⟦\_⟧}})}.
+This solution needs only to give \AF{Lang} (now tries) an index {\AB{i} \AK : \APT{Size}} corresponding to the maximum depth to which a trie can be searched, or equivalently the longest string that can be matched\out{ (via {\AF{⟦\_⟧}})}.
 In practice, we will work with arbitrarily deep tries, i.e., ones having index \APo{∞}, as in the type of {\AF{⟦\_⟧}}.

### 2021-02-13 paper.tex
wording

@@
 The ``symbolic differentiation'' style reflects language-building vocabulary (\figref{Lang-ops}) into an inductive data type (of modestly extended regular expressions), while re-defining \AF{ν} and \AF{δ} as functions that pattern-match on that data type.
-For convenience, use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the prefix ``{◬.}'' (module-specific names).
+For convenience, use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the module prefix ``{◬.}\hspace{0.05em}''.
 The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\agda{api}} along with a semantic function {\AF{⟦\_⟧}} that converts a symbolic language representation to decidable form.
@@
 Now consider the dual strategy of ``automatic differentiation''.
-This time, we reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type (of tries \needcite{}), while redefining the language-building vocabulary as functions on that data type.
-Again, we will use the same names as in \secref{Specifying Languages}, referring to the original versions via the prefix ``{◬.}''.
+This time, reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type (of tries \needcite{}), while redefining the language-building vocabulary as functions on that data type defined by \emph{copatterns} \citep{AbelPientka2016}.
+Again, we will use the same names as in \secref{Specifying Languages}, referring to the original versions via the module prefix ``{◬.}\hspace{0.05em}''.
 The result is shown in \figrefdef{automatic-api}{Tries (coinductive)}{\agda{api}}.
@@
 These definitions are meant to be read in ``row-major'' order, i.e., each row is one definition.
-The \AF{Lang} representation is coinductive, and definitions are by \emph{copattern} matching \citep{AbelPientka2016}.
 Note that the definitions are syntactically identical to those in \figref{symbolic-defs} but are organized dually.
@@
 The \AF{δ} clause for {\AB{p} \AF ⋆ \AB{q}} does not satisfy Agda's termination checker, which cannot see that the argument {\AF{δ} \AB{p} \AB{a}} in the recursive use of \AF{\_⋆\_} is in some sense smaller than \AB{p}.
-Fortunately, this issue was already solved by Andreas \citet{Abel2016}, also in the setting of trie-based language recognition, by using sized types \citep{Abel2008, AbelPientka2016}.
+Fortunately, this issue was already identified and solved by Andreas \citet{Abel2016}---also in the setting of trie-based language recognition---by using \emph{sized types} \citep{Abel2008, AbelPientka2016}.
 \rnc\source{SizedAutomatic}
@@
 
-Decidable recognition is defined exactly as with unsized tries (\figref{automatic-defs}), but this time the compiler successfully proves \emph{total} correctness (including termination).
+Decidable recognition is defined exactly as with unsized tries (\figref{automatic-defs}), and this time the compiler successfully proves \emph{total} correctness (including termination).
 

### 2021-02-13 paper.tex
--safe

@@
 The use of type-level predicates makes for simple, direct specification, including the use of existential quantification.
-As described in \secref{Proving Properties}, it's also fairly easy to prove algebraic properties of predicates\out{, including the semimodule and semiring laws for both predicate semirings}.
+As described in \secref{Properties}, it's also fairly easy to prove algebraic properties of predicates\out{, including the semimodule and semiring laws for both predicate semirings}.
 These definitions do not, however, give us decidable parsing, because general type inhabitation amounts to theorem proving.
@@
 
-\sectionl{Proving Properties}
+\sectionl{Properties}
+
+\subsectionl{Predicate Algebra}
+
+The basic building blocks of type-level predicates---and languages in particular---form the vocabulary of a \emph{closed semiring} in two different ways\out{, as reflected in the structure of regular expressions \needcite{}}.
+The semiring abstraction has three aspects: (a) a commutative monoid providing ``zero'' and ``addition'', (b) a (possibly non-commutative) monoid providing ``one'' and ``multiplication'', and (c) the relationship between them, namely that multiplication distributes over addition and zero.\footnote{Distributing of multiplication over zero is also known as ``annihilation''.}
+In the first predicate semiring, which is \emph{commutative} (i.e., multiplication commutes), zero and addition are \AF ∅ and \AF{\_∪\_}, while one and multiplication are \AF 𝒰 and \AF{\_∩\_}.
+
+Closure adds the star operation.
+Following existing conventions \stdlibCitep{Algebra.Structures}, extend the semiring operations and laws with star and its law, parametrized over an arbitrary type \AB{A} and equivalence relation \AB{\_≈\_}:\notefoot{To do: add (right-associated) \AF{starʳ}.}
+\ExecuteMetaData[Closed/Structures.tex]{closed}
+The \AF{ClosedSemiring} algebraic structure is similarly extended to \AF{ClosedCommutativeSemiring}.
+
+Conveniently, booleans and types form commutative semirings with all necessary definitions already in the standard library.
+(The equivalence relation used for types is isomorphism rather than equality.)
+Both are also closed.
+For booleans, closure maps both \AIC{false} and \AIC{true} to \AIC{true}, with the \AF{starˡ} law holding definitionally.
+For types, the closure of \AB{A} is {\AB{A} ✶} (the usual inductive list type) with a simple, non-inductive proof of \AF{starˡ}.
+
+\begin{code}[hide]
+open import Closed ; open Closed.Types {ℓ}
+\end{code}
+
+This first closed semiring for predicates follows from a much more general pattern.
+Given any two types \AB{A} and \AB{B}, if \AB{B} is a monoid then {\AB A \AS → \AB B} is as well.
+The monoid operation {\AB{\_∙\_} \AS : \AF{Op₂} \AB{B}} lifts to the function-level binary operation {\AS λ (\AB{f} \AB{g} : \AB{A} \AS → \AB{B}) \AS → \AS λ \AB a \AS → \AB f \AB a \AB ∙ \AB g \AB a}.
+The monoid identity {\AB ε \AS : \AB{B}} lifts to the identity {\AS λ \AB a \AS → \AB ε}.
+All of the laws transfer from \AB{B} to {\AB A \AS → \AB B}.
+Likewise for other algebraic structures.
+
+Looking more closely, additional algebraic structure emerges on predicates: (full) \emph{semimodules} generalize vector spaces by relaxing the associated types of ``scalars'' from fields to commutative semirings, i.e., dropping the requirements of multiplicative and additive inverses \needcite{}.
+Left and right semimodules further generalize scalars to arbitrary semirings, dropping the commutativity requirement for scalar multiplication.
+Again, nothing is needed from the codomain type \AB{B} beyond the properties of a semiring, so again predicates get these algebraic structures for free (already paid for by types).
+For every {\AB{P} \AS : \AF{Pred} \AB{A}}, \AB{P} is a vector in the semimodule {\AF{Pred} \AB{A}}, including the special case of ``languages'', for which \AB{A} is a list type.
+The dimension of the semimodule is the cardinality of the domain type \AB{A} (typically infinite), with ``basis vectors'' having the form {\AF{pureⱽ} \AB{a}} for {\AB{a} \AS : \AB{A}}.
+A general vector (predicate) is a linear combination (often infinite) of these basis vectors, with addition being union and scaling conjoining membership proofs.
+
+\begin{code}[hide]
+⋆-∪-isClosedSemiring : IsClosedSemiring _⟷_ _∪_ _⋆_ ∅ 𝟏  _☆
+⋆-∪-isClosedSemiring = record { isSemiring = ⋆-∪-isSemiring ; starˡ = ☆-starˡ }
+
+⋆-∪-isClosedCommutativeSemiring :
+  Commutative _≡_ _∙_ → IsClosedCommutativeSemiring _⟷_ _∪_ _⋆_ ∅ 𝟏 _☆
+⋆-∪-isClosedCommutativeSemiring ∙-comm = record
+  { isCommutativeSemiring = ⋆-∪-isCommutativeSemiring ∙-comm
+  ; starˡ = ☆-starˡ
+  }
+
+⋆-∪-ClosedSemiring : ClosedSemiring _ _
+⋆-∪-ClosedSemiring = record { isClosedSemiring =  ⋆-∪-isClosedSemiring }
+
+⋆-∪-ClosedCommutativeSemiring : Commutative _≡_ _∙_ → ClosedCommutativeSemiring _ _
+⋆-∪-ClosedCommutativeSemiring ∙-comm =
+  record { isClosedCommutativeSemiring = ⋆-∪-isClosedCommutativeSemiring ∙-comm }
+\end{code}
+
+There is still more algebraic structure to be found and exploited.
+When our \AB{domain} type \AB{A} is a monoid (as with languages, infinite streams and grids, and functions of continuous space and time), predicates over \AB{A} form a second semiring, known as ``the monoid semiring'' \needcite{}, in which zero and addition are as in the first predicate commutative semiring and semimodule, while one and multiplication are \AF{𝟏} and \AF{\_⋆\_}, defined in \secref{Languages, Predicates, and Types}.
+In this setting, language concatenation is subsumed by a very general form of \emph{convolution} \needcite{}.
+The semimodule structure of predicates provides the additive aspect and is built entirely from the \emph{codomain}-transforming predicate operations defined in \secref{Languages, Predicates, and Types}.
+The multiplicative aspect (convolution) comes entirely from the \emph{domain}-transforming operations, applied to any domain monoid.
+The two needed distributive properties hold for \emph{any} domain-transforming operation, whether associative or not.
+The proofs of these properties, while not terribly complex, are beyond the scope of this paper but are available in the paper's Agda source code.\notefoot{There will be short and long forms of this paper.
+Perhaps the long form will contain narrated parts of the monoid semiring proof.}
+
+Not only do types form a closed semiring, the \emph{only} properties needed from types in the monoid semiring come from the laws of closed semirings.
+We can thus apply the ideas in this paper to many other useful semirings as well, including natural numbers (e.g., number of parses), probability distributions, and the tropical semirings (max/plus and min/plus) for optimization.
+\out{We will explore this generality in \secref{Beyond Predicates}.}
+As a bonus, when the domain monoid commutes (multiplicatively)---e.g., for functions of space and time rather than languages---the monoid semiring does as well .
+
+\subsectionl{Transferring properties}
 

### 2021-02-15 paper.tex
predicates
misc

@@
 For compositionality, we will use a few operations that lift decidability of types to decidability of constructions on those types, as shown in \figrefdef{compositional-dec}{Compositional decidability}{\agda{compositional-dec}}.\notefoot{Explanatory footnote goes here.}
-
-A \emph{language} (or any predicate) \AB{P} is ``decidable'' when (the type) {\AB{P} \AB{x}} is decidable for all \AB{x}\out{ \stdlibCitep{Relation.Unary}}:
+Moreover, decidability lifts naturally from types to predicates:
 \agda{Decidable}
@@
 
+\nc\api{\setstretch{1.1}\agda{api}}
+
 The ``symbolic differentiation'' style reflects language-building vocabulary (\figref{Lang-ops}) into an inductive data type (of modestly extended regular expressions), while re-defining \AF{ν} and \AF{δ} as functions that pattern-match on that data type.
 For convenience, use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the module prefix ``{◬.}\hspace{0.05em}''.
-The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\agda{api}} along with a semantic function {\AF{⟦\_⟧}} that converts a symbolic language representation to decidable form.
+The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\api} along with a semantic function {\AF{⟦\_⟧}} that converts a symbolic language representation to decidable form.
 
@@
 Again, we will use the same names as in \secref{Specifying Languages}, referring to the original versions via the module prefix ``{◬.}\hspace{0.05em}''.
-The result is shown in \figrefdef{automatic-api}{Tries (coinductive)}{\agda{api}}.
+The result is shown in \figrefdef{automatic-api}{Tries (coinductive)}{\api}.
 
@@
 In practice, we will work with arbitrarily deep tries, i.e., ones having index \APo{∞}, as in the type of {\AF{⟦\_⟧}}.
-The modified representation is shown in \figrefdef{sized automatic-api}{Sized tries (coinductive)}{\agda{api}}.
+The modified representation is shown in \figrefdef{sized automatic-api}{Sized tries (coinductive)}{\api}.
 
 Decidable recognition is defined exactly as with unsized tries (\figref{automatic-defs}), and this time the compiler successfully proves \emph{total} correctness (including termination).
+
+\begin{comment}
 
@@
 
+\end{comment}
+
 \bibliography{bib}

### 2021-02-16 paper.tex
deactivate starʳ for now ; predicate generalization

@@
 
-\begin{comment}
+\sectionl{Generalizing from languages to predicates}
+
+\rnc\source{Predicate}
+
+The language-building vocabulary defined in \secref{Specifying Languages} can be simplified and generalized.
+First note that there are two categories of operations.
+One category transforms the codomain (types): \AF{∅}, \AF 𝒰, \AF{\_∪\_}, \AF{\_∩\_}.
+The other transforms the domain (lists): \AF{𝟏}, \AF{\_⋆\_}, \AF{\_☆}, \AF{`}.
+The first category applies to predicates over all types (not just over lists).
+Within the second category, \AF{𝟏}, \AF{\_⋆\_}, and \AF{\_☆} apply to all monoids, while \AF{`} is specific to lists.
+
+\figrefdef{predicates}{Predicate operations}{
+\begin{center}
+\mathindent-12ex
+\agda{Pred}
+\end{center}
+\vspace{-2ex}
+\hfill
+\mathindent0ex
+\begin{minipage}[b]{17em}
+%% Transform codomain: %\small
+\agda{codomain-transformers}
+\end{minipage}
+\hfill
+\begin{minipage}[b]{20em}
+%% Transform domain: %\small
+\agda{domain-transformers}
+\end{minipage}
+\hfill\;
+} shows two parallel collections of predicate operations, each covariantly transforming the codomain (left) or domain (right).
+These generalized operations then specialize to language operations as in \figrefdef{lang-via-pred}{Languages via predicate operations}{
+\mathindent0ex
+\begin{center}
+\mathindent-12ex
+\agda{Lang}
+\end{center}
+\vspace{-2ex}
+\hfill
+\begin{minipage}[c]{11em}
+\setstretch{1.7}
+%% (any domain type)
+\agda{codomain-ops}
+\end{minipage}
+\hfill
+\begin{minipage}[c]{15em}
+%% (any monoid)
+\agda{domain-ops}
+%% (lists)
+\agda{list-ops}
+\end{minipage}
+\hfill\;
+}, which also includes language complement (\AF ∁).
+\rnc\source{Calculus}
+The language representations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation} adapt easily to the generalized codomain operations (\AF{pureᵀ}, \AF{mapᵀ}, and \AF{mapᵀ₂}) resulting in broader coverage with less code.
+The relevant \AF{ν} and \AF{δ} lemmas in \figref{nu-delta-lemmas} are easily generalized as shown in \figrefdef{nu-delta-codomain-lemmas}{Properties of \AF{ν} and \AF{δ} for predicate codomain operations}{\agda{νδ-codomain}} and are all proved automatically by normalization.
 
@@
 In the first predicate semiring, which is \emph{commutative} (i.e., multiplication commutes), zero and addition are \AF ∅ and \AF{\_∪\_}, while one and multiplication are \AF 𝒰 and \AF{\_∩\_}.
-
-Closure adds the star operation.
-Following existing conventions \stdlibCitep{Algebra.Structures}, extend the semiring operations and laws with star and its law, parametrized over an arbitrary type \AB{A} and equivalence relation \AB{\_≈\_}:\notefoot{To do: add (right-associated) \AF{starʳ}.}
-\ExecuteMetaData[Closed/Structures.tex]{closed}
-The \AF{ClosedSemiring} algebraic structure is similarly extended to \AF{ClosedCommutativeSemiring}.
+Closure adds a star/closure operation \AF{\_✯} with \AF{starˡ} law: {\AB x \AF ✯ \AF ≈ 1 \AF + \AB x \AF ✲ \AB x \AF ✯}.
 
@@
 
-\begin{code}[hide]
-open import Closed ; open Closed.Types {ℓ}
-\end{code}
-
 This first closed semiring for predicates follows from a much more general pattern.
 Given any two types \AB{A} and \AB{B}, if \AB{B} is a monoid then {\AB A \AS → \AB B} is as well.
-The monoid operation {\AB{\_∙\_} \AS : \AF{Op₂} \AB{B}} lifts to the function-level binary operation {\AS λ (\AB{f} \AB{g} : \AB{A} \AS → \AB{B}) \AS → \AS λ \AB a \AS → \AB f \AB a \AB ∙ \AB g \AB a}.
+The monoid operation \AB{\_∙\_} lifts to the function-level binary operation {\AS λ (\AB{f} \AB{g} : \AB{A} \AS → \AB{B}) \AS → \AS λ \AB a \AS → \AB f \AB a \AB ∙ \AB g \AB a}.
 The monoid identity {\AB ε \AS : \AB{B}} lifts to the identity {\AS λ \AB a \AS → \AB ε}.
@@
 A general vector (predicate) is a linear combination (often infinite) of these basis vectors, with addition being union and scaling conjoining membership proofs.
-
-\begin{code}[hide]
-⋆-∪-isClosedSemiring : IsClosedSemiring _⟷_ _∪_ _⋆_ ∅ 𝟏  _☆
-⋆-∪-isClosedSemiring = record { isSemiring = ⋆-∪-isSemiring ; starˡ = ☆-starˡ }
-
-⋆-∪-isClosedCommutativeSemiring :
-  Commutative _≡_ _∙_ → IsClosedCommutativeSemiring _⟷_ _∪_ _⋆_ ∅ 𝟏 _☆
-⋆-∪-isClosedCommutativeSemiring ∙-comm = record
-  { isCommutativeSemiring = ⋆-∪-isCommutativeSemiring ∙-comm
-  ; starˡ = ☆-starˡ
-  }
-
-⋆-∪-ClosedSemiring : ClosedSemiring _ _
-⋆-∪-ClosedSemiring = record { isClosedSemiring =  ⋆-∪-isClosedSemiring }
-
-⋆-∪-ClosedCommutativeSemiring : Commutative _≡_ _∙_ → ClosedCommutativeSemiring _ _
-⋆-∪-ClosedCommutativeSemiring ∙-comm =
-  record { isClosedCommutativeSemiring = ⋆-∪-isClosedCommutativeSemiring ∙-comm }
-\end{code}
 
@@
 \out{We will explore this generality in \secref{Beyond Predicates}.}
-As a bonus, when the domain monoid commutes (multiplicatively)---e.g., for functions of space and time rather than languages---the monoid semiring does as well .
+As a bonus, when the domain monoid commutes (multiplicatively)---e.g., for functions of space and time rather than languages---the monoid semiring does as well.
 
-\subsectionl{Transferring properties}
-
-\end{comment}
+\subsectionl{Transporting algebraic properties}
 

### 2021-02-17 paper.tex
proving and transporting properties

@@
 In practice, we will work with arbitrarily deep tries, i.e., ones having index \APo{∞}, as in the type of {\AF{⟦\_⟧}}.
-The modified representation is shown in \figrefdef{sized automatic-api}{Sized tries (coinductive)}{\api}.
+The modified representation is shown in \figrefdef{sized-automatic-api}{Sized tries (coinductive)}{\api}.
 
@@
 \rnc\source{Calculus}
-The language representations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation} adapt easily to the generalized codomain operations (\AF{pureᵀ}, \AF{mapᵀ}, and \AF{mapᵀ₂}) resulting in broader coverage with less code.
-The relevant \AF{ν} and \AF{δ} lemmas in \figref{nu-delta-lemmas} are easily generalized as shown in \figrefdef{nu-delta-codomain-lemmas}{Properties of \AF{ν} and \AF{δ} for predicate codomain operations}{\agda{νδ-codomain}} and are all proved automatically by normalization.
+The \AF{ν} and \AF{δ} lemmas for codomain operations in \figref{nu-delta-lemmas} are easily generalized as shown in \figrefdef{nu-delta-codomain-lemmas}{Properties of \AF{ν} and \AF{δ} for predicate codomain operations}{\agda{νδ-codomain}} and are all proved automatically by normalization.
+(In contrast, the \AF{ν} and \AF{δ} lemmas do not easily generalize for domain operations.)
+Since the decidable language implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation} are corollaries of \AF{ν} and \AF{δ} lemmas, those implementations adapt easily to the generalized codomain operations (\AF{pureᵀ}, \AF{mapᵀ}, and \AF{mapᵀ₂}), resulting in a somewhat smaller implementation with broader coverage code.
 
@@
 
-\subsectionl{Predicate Algebra}
+\sectionl{Predicate Algebra}
 
@@
 The two needed distributive properties hold for \emph{any} domain-transforming operation, whether associative or not.
-The proofs of these properties, while not terribly complex, are beyond the scope of this paper but are available in the paper's Agda source code.\notefoot{There will be short and long forms of this paper.
-Perhaps the long form will contain narrated parts of the monoid semiring proof.}
+The proofs of these properties, while reasonably straightforward, are beyond the scope of this paper but are available in the paper's Agda source code.\out{\notefoot{There will be short and long forms of this paper.
+Perhaps the long form will contain narrated parts of the monoid semiring proof.}}
 
 Not only do types form a closed semiring, the \emph{only} properties needed from types in the monoid semiring come from the laws of closed semirings.
-We can thus apply the ideas in this paper to many other useful semirings as well, including natural numbers (e.g., number of parses), probability distributions, and the tropical semirings (max/plus and min/plus) for optimization.
+One can thus apply the ideas in this paper to many other useful semirings as well, including natural numbers (e.g., number of parses), probability distributions, and the tropical semirings (max/plus and min/plus) for optimization.
 \out{We will explore this generality in \secref{Beyond Predicates}.}
@@
 
-\subsectionl{Transporting algebraic properties}
+\subsectionl{Transporting Algebraic Properties}
+
+\rnc\source{Transport}
+
+Given that type-level languages have useful and interesting algebraic properties and that the decidable implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation} are tightly connected to languages, do the algebraic properties transfer to those implementations?
+This question immediately raises a technical obstacle.
+Algebraic abstractions and their properties are simply typed.
+For instance, each instance of the \ARe{Monoid} abstraction involves a single type \AB τ with a binary operation {\AB{\_∙\_} \AK : \AB τ \AK → \AB τ \AK → \AB τ} and identity value {\AB ε \AK : \AB τ}.
+This recipe accommodates the language operations in \figref{Lang-ops} and their generalizations in \figref{predicates} but not the dependently typed, \emph{indexed} versions of languages and their operations in \figreftwo{symbolic-api}{sized-automatic-api}, nor decidable types and their vocabulary in \figref{compositional-dec}.
+
+Fortunately, there is a simple way to encapsulate indexed types into simple (non-indexed) types, namely existential quantification.
+For instance, {\AF{∃} \AD{Dec}} and {\AF{∃} \AD{(\ARe{Lang} \APo ∞)}} are simple types.
+For a type family {\AB f \AK : \AB A \AK → \APT Set⇃}, values of type {\AF ∃ \AB f} are pairs {(\AB a , \AB b)} with {\AB a \AK : \AB A} and {\AB b \AK : \AB f \AB a}.
+We can also encapsulate dependently typed \emph{operations} on indexed types into simply typed operations on the (simply typed) existentially wrapped versions of those types.
+For instance, consider language union from \figref{symbolic-api}:
+\begin{code}
+% Copied from latex/Symbolic.tex
+\>[2]\AgdaOperator{\AgdaInductiveConstructor{\AgdaUnderscore{}∪\AgdaUnderscore{}}}%
+\>[7]\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaDatatype{Lang}%
+\>[15]\AgdaGeneralizable{P}%
+\>[18]\AgdaSymbol{→}\AgdaSpace{}%
+\AgdaDatatype{Lang}\AgdaSpace{}%
+\AgdaGeneralizable{Q}%
+\>[28]\AgdaSymbol{→}\AgdaSpace{}%
+\AgdaDatatype{Lang}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaGeneralizable{P}%
+\>[39]\AgdaOperator{\AgdaFunction{◬.∪}}%
+\>[44]\AgdaGeneralizable{Q}\AgdaSymbol{)}\<%
+\end{code}
+Referring to the type and operations of \figref{symbolic-api} via the module prefix ``{◬.}\hspace{0.05em}'', define \agda{wrapped-Lang}
+Then wrap union as follows:
+\agda{wrapped-union}
+Note that the type-level language components of the arguments and result are left to be inferred by the compiler.
+\rnc\source{Existential}
+Indexed operations of all arities for a type family \AB{f} can be systematically wrapped in this style, as shown in
+\figrefdef{inj}{Existential wrappers}{\agda{inj}}, this time with the inferable components filled in for clarity.
+Then wrapped union can be defined simply as {\AF{\_∪\_} \AS = \AF{inj₂} \AIC{◬.\_∪\_}} and likewise for the other operations.
+
+In addition to wrapping types and operations, we also need to transport \emph{properties}.
+A crucial question is the choice of equivalence relation for the new properties.
+Choosing wrapped equivalence to be equivalence of \emph{first} projections (type \emph{indices}) results in all algebraic properties of those indices easily transporting to the existentially wrapped versions.
+This choice may at first seem like cheating, since the entire operational aspect of the representation is ignored.
+The correctness of that aspect, however, is guaranteed by its dependent typing, which anchors its meaning (no matter how cleverly implemented) to the non-operational type index.
+For language implementations of \secreftwo{Symbolic Differentiation}{Automatic Differentiation}, this anchoring is guaranteed by the type of \AF{⟦\_⟧} in \figreftwo{symbolic-api}{sized-automatic-api}.
+Given this choice of equivalence, properties of index types are easily lifted to properties of existential types, as in \figrefdef{prop}{Liftings of index properties to existential types}{\agda{prop}}.
+These definitions ease specification of algebraic instances for existentially wrapped types from corresponding instances for their indices.
+% As an example, \figrefdef{wrap-semiring}{Lifting semiring instances}{\agda{Semiring}} 
+As examples, we can lift the commutative semiring of types to wrappings of the decidable counterparts from \figref{compositional-dec} in \secref{Decidability}, and lift the \AF{\_∪\_}/\AF{\_⋆\_} language semiring of \secref{Specifying Languages} to the decidable implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation}, as follows (where \AF{\_≟\_} is decidable equality on list elements):\notefoot{Lift the \AF{\_∪\_}/\AF{\_∩\_} predicate semiring as well.}
+\rnc\source{Transport}
+\agda{examples}
 

### 2021-02-17 paper.tex
tidy examples

@@
 % As an example, \figrefdef{wrap-semiring}{Lifting semiring instances}{\agda{Semiring}} 
-As examples, we can lift the commutative semiring of types to wrappings of the decidable counterparts from \figref{compositional-dec} in \secref{Decidability}, and lift the \AF{\_∪\_}/\AF{\_⋆\_} language semiring of \secref{Specifying Languages} to the decidable implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation}, as follows (where \AF{\_≟\_} is decidable equality on list elements):\notefoot{Lift the \AF{\_∪\_}/\AF{\_∩\_} predicate semiring as well.}
+As examples, we can lift the commutative semiring of types to wrappings of the decidable counterparts from \figref{compositional-dec} in \secref{Decidability}, and lift both language semirings to the decidable implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation}, as follows\out{ (where \AF{\_≟\_} is decidable equality on list elements)}:\notefoot{What is the first closure for predicates?}
 \rnc\source{Transport}

### 2021-02-17 paper.tex
shorten subsection title

@@
 
-\subsectionl{Transporting Algebraic Properties}
+\subsectionl{Transporting Properties}
 
@@
 For instance, {\AF{∃} \AD{Dec}} and {\AF{∃} \AD{(\ARe{Lang} \APo ∞)}} are simple types.
-For a type family {\AB f \AK : \AB A \AK → \APT Set⇃}, values of type {\AF ∃ \AB f} are pairs {(\AB a , \AB b)} with {\AB a \AK : \AB A} and {\AB b \AK : \AB f \AB a}.
+For a type family {\AB F \AK : \AB A \AK → \APT Set⇃}, values of type {\AF ∃ \AB F} are pairs {(\AB a , \AB b)} with {\AB a \AK : \AB A} and {\AB b \AK : \AB F \AB a}.
 We can also encapsulate dependently typed \emph{operations} on indexed types into simply typed operations on the (simply typed) existentially wrapped versions of those types.
@@
 \rnc\source{Existential}
-Indexed operations of all arities for a type family \AB{f} can be systematically wrapped in this style, as shown in
+Indexed operations of all arities for a type family \AB{F} can be systematically wrapped in this style, as shown in
 \figrefdef{inj}{Existential wrappers}{\agda{inj}}, this time with the inferable components filled in for clarity.

### 2021-02-17 paper.tex
fix section references; mathindent

@@
 There is still more algebraic structure to be found and exploited.
-When our \AB{domain} type \AB{A} is a monoid (as with languages, infinite streams and grids, and functions of continuous space and time), predicates over \AB{A} form a second semiring, known as ``the monoid semiring'' \needcite{}, in which zero and addition are as in the first predicate commutative semiring and semimodule, while one and multiplication are \AF{𝟏} and \AF{\_⋆\_}, defined in \secref{Languages, Predicates, and Types}.
+When our \AB{domain} type \AB{A} is a monoid (as with languages, infinite streams and grids, and functions of continuous space and time), predicates over \AB{A} form a second semiring, known as ``the monoid semiring'' \needcite{}, in which zero and addition are as in the first predicate commutative semiring and semimodule, while one and multiplication are \AF{𝟏} and \AF{\_⋆\_}, defined in \secref{Specifying Languages}.
 In this setting, language concatenation is subsumed by a very general form of \emph{convolution} \needcite{}.
-The semimodule structure of predicates provides the additive aspect and is built entirely from the \emph{codomain}-transforming predicate operations defined in \secref{Languages, Predicates, and Types}.
+The semimodule structure of predicates provides the additive aspect and is built entirely from the \emph{codomain}-transforming predicate operations defined in \secref{Specifying Languages}.
 The multiplicative aspect (convolution) comes entirely from the \emph{domain}-transforming operations, applied to any domain monoid.
@@
 Indexed operations of all arities for a type family \AB{F} can be systematically wrapped in this style, as shown in
-\figrefdef{inj}{Existential wrappers}{\agda{inj}}, this time with the inferable components filled in for clarity.
+\figrefdef{inj}{Existential wrappers}{\mathindent0ex\agda{inj}}, this time with the inferable components filled in for clarity.
 Then wrapped union can be defined simply as {\AF{\_∪\_} \AS = \AF{inj₂} \AIC{◬.\_∪\_}} and likewise for the other operations.
@@
 For language implementations of \secreftwo{Symbolic Differentiation}{Automatic Differentiation}, this anchoring is guaranteed by the type of \AF{⟦\_⟧} in \figreftwo{symbolic-api}{sized-automatic-api}.
-Given this choice of equivalence, properties of index types are easily lifted to properties of existential types, as in \figrefdef{prop}{Liftings of index properties to existential types}{\agda{prop}}.
+Given this choice of equivalence, properties of index types are easily lifted to properties of existential types, as in \figrefdef{prop}{Liftings of index properties to existential types}{\mathindent0ex\agda{prop}}.
 These definitions ease specification of algebraic instances for existentially wrapped types from corresponding instances for their indices.

### 2021-02-17 paper.tex
footnotes

@@
-\documentclass[hidelinks,12pt]{article}  % fleqn,12pt
-\usepackage[margin=0.2in]{geometry}  % 0.12in, 0.9in, 1in
+\documentclass[hidelinks]{article}  % fleqn,12pt
+\usepackage[margin=1in]{geometry}  % 0.12in, 0.9in, 1in
 
@@
 \figrefdef{Lang-ops}{Language operations}{\agda{Lang-ops}} gives definitions of the usual language operations, including language union and intersection ({\AB{P} \AF ∪ \AB{Q}} and {\AB{P} \AF ∩ \AB{Q}}) with their identities (the empty language \AF ∅ and universal language \AF 𝒰), language concatenation ({\AB{P} \AF ⋆ \AB{Q}}) and its identity (\AF 𝟏, containing only the empty string), single-character languages ({\AF ` \AB{c}}), Kleene star ({\AB{P} \AF ☆}), and ``scalar multiplication'' (\mbox{\AB{s} \AF · \AB{P}}, which will prove useful later).
-
 \note{Explain the types used here and their logical interpretation under the Curry-Howard isomorphism.}
@@
 \hfill\;
-} shows two parallel collections of predicate operations, each covariantly transforming the codomain (left) or domain (right).
+} shows two parallel collections of predicate operations, each covariantly transforming the codomain (left) or domain (right).\footnote{The types of operations on the left can be further relaxed from \AF{Set⇂} to any codomain type, while keeping \AB{A} fixed, resulting in the usual covariant applicative functor of functions \emph{from} a fixed type \citep{McBride2008APE}.
+Dually, the operations on the right (in their current generality) form a covariant applicative functor of functions \emph{to} the type \AF{Set⇂}.}
 These generalized operations then specialize to language operations as in \figrefdef{lang-via-pred}{Languages via predicate operations}{
@@
 }, which also includes language complement (\AF ∁).
+
 \rnc\source{Calculus}
@@
 These definitions ease specification of algebraic instances for existentially wrapped types from corresponding instances for their indices.
-% As an example, \figrefdef{wrap-semiring}{Lifting semiring instances}{\agda{Semiring}} 
-As examples, we can lift the commutative semiring of types to wrappings of the decidable counterparts from \figref{compositional-dec} in \secref{Decidability}, and lift both language semirings to the decidable implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation}, as follows\out{ (where \AF{\_≟\_} is decidable equality on list elements)}:\notefoot{What is the first closure for predicates?}
 \rnc\source{Transport}
+As examples, we can lift the commutative semiring of types to wrappings of the decidable counterparts from \figref{compositional-dec} in \secref{Decidability}, and lift both language semirings to the decidable implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation}, as in \figrefdef{transport-examples}{Examples of algebraic instances transported from indices to existential wrappings}{
+\mathindent0ex
 \agda{examples}
+}\out{ (where \AF{\_≟\_} is decidable equality on list elements)}:\notefoot{What is the first closure for predicates?}
 

### 2021-02-17 paper.tex
conditionally use acmart
transport examples line wrap
ruleSep

@@
+\newif\ifacm
+
+\acmtrue
+%% \acmfalse
+
+\ifacm
+\documentclass[acmsmall,screen]{acmart}
+\else
 \documentclass[hidelinks]{article}  % fleqn,12pt
 \usepackage[margin=1in]{geometry}  % 0.12in, 0.9in, 1in
+\usepackage[useregional]{datetime2}
+\fi
 
 \usepackage{catchfilebetweentags}
-\usepackage[useregional]{datetime2}
 
@@
 
-\RequirePackage{newunicodechar, amssymb, stmaryrd, unicode-math, setspace, comment}
+\ifacm\else
+%% Already in acmart
+\RequirePackage{amssymb, unicode-math}
+\fi
+\RequirePackage{newunicodechar, stmaryrd, setspace, comment}
 
@@
 \input{macros}
+
+\ifacm
+\bibliographystyle{ACM-Reference-Format}
+\else
+\usepackage[round]{natbib} % [square]
+\bibliographystyle{plainnat}
+\fi
+
+%% I prefer ∅ in XITSMath-Regular, but I don't know how to make setmathfont
+%% work with acmart.
+\ifacm
+\newunicodechar{◬}{\raisebox{0.02ex}{\ensuremath{_{\diamond}\hspace{-0.323em}}}}
+\else
+\setmathfont{XITSMath-Regular.otf}
+\newunicodechar{◬}{\raisebox{0.23ex}{\ensuremath{_{\diamond}\hspace{-0.352em}}}}
+\newunicodechar{∷}{\ensuremath{\mathbin{\hspace{-0.2em}:\hspace{-0.27em}:\hspace{-0.15em}}}}
+\fi
 
@@
 
+\ifacm\else
 \maketitle
+\fi
 
@@
 Applying a language to a string yields a type whose elements are language membership proofs describing \emph{how} a string parses in the language.
-The usual building blocks of languages (including union, contatenation, and Kleene closure) have precise and compelling specifications uncomplicated by operational strategies and are easily generalized to a few general domain-transforming and codomain-transforming operations on predicates.
+The usual building blocks of languages (including union, concatenation, and Kleene closure) have precise and compelling specifications uncomplicated by operational strategies and are easily generalized to a few general domain-transforming and codomain-transforming operations on predicates.
 
@@
 \end{abstract}
+
+\ifacm
+\maketitle
+\mathindent2em
+\fi
 
@@
 \agda{rules}
+\rnc\rulesSep{-3ex}
 
@@
 \figrefdef{predicates}{Predicate operations}{
+\ifacm
+\small
+\fi
 \begin{center}
@@
 \agda{examples}
-}\out{ (where \AF{\_≟\_} is decidable equality on list elements)}:\notefoot{What is the first closure for predicates?}
-
+}
 \bibliography{bib}

### 2021-02-17 paper.tex
box-dot

@@
 \fi
-\RequirePackage{newunicodechar, stmaryrd, setspace, comment}
+\RequirePackage{newunicodechar, stmaryrd, setspace, comment, scalerel}
 
@@
 \newunicodechar{◬}{\raisebox{0.02ex}{\ensuremath{_{\diamond}\hspace{-0.323em}}}}
+\newunicodechar{▣}{\raisebox{-0.12ex}{\ensuremath{\scaleobj{0.4}{\Box}}\hspace{-0.238em}}}
 \else
@@
 \newunicodechar{◬}{\raisebox{0.23ex}{\ensuremath{_{\diamond}\hspace{-0.352em}}}}
+\newunicodechar{▣}{\raisebox{-0.13ex}{\ensuremath{\scaleobj{0.3}{\Box}}\hspace{-0.23em}}}
 \newunicodechar{∷}{\ensuremath{\mathbin{\hspace{-0.2em}:\hspace{-0.27em}:\hspace{-0.15em}}}}
@@
 
-A simple characterization of languages (and indeed functions from lists to any type) captures the essential idea behind language ``differentiation'' and ``integration'' as used for recognizing languages and leads to a collection of lemmas about type-level languages.
+A simple characterization of languages (and indeed functions from lists to any type) captures the essential idea behind language ``differentiation'' and ``integration'' as used for recognizing languages and leads to a collection of lemmas about type-level predicates.
 These lemmas form the backbone of two dual implementations of language recognition---(inductive) regular expressions and (coinductive) tries---each containing the same code but in dual arrangements.
@@
 
-\sectionl{Predicate Algebra}
+\subsectionl{Predicate Algebra}
 
@@
 \end{code}
-Referring to the type and operations of \figref{symbolic-api} via the module prefix ``{◬.}\hspace{0.05em}'', define \agda{wrapped-Lang}
+Referring to the type and operations of \secref{Specifying Languages} and \figref{symbolic-api} via the module prefix ``{◬.}\hspace{0.05em}'' and ``{▣.}\hspace{0.05em}'' respectively, define \agda{wrapped-Lang}
 Then wrap union as follows:
 \agda{wrapped-union}
-Note that the type-level language components of the arguments and result are left to be inferred by the compiler.
+The type-level language components of the arguments and result are given explicitly here but could instead be inferred by the compiler.
 \rnc\source{Existential}
 Indexed operations of all arities for a type family \AB{F} can be systematically wrapped in this style, as shown in
-\figrefdef{inj}{Existential wrappers}{\mathindent0ex\agda{inj}}, this time with the inferable components filled in for clarity.
+\figrefdef{inj}{Existential wrappers}{\mathindent0ex\agda{inj}}, again with inferable components given explicitly for clarity.
 Then wrapped union can be defined simply as {\AF{\_∪\_} \AS = \AF{inj₂} \AIC{◬.\_∪\_}} and likewise for the other operations.

### 2021-02-17 paper.tex
indent

@@
 
-Languages are usually formalized either set-theoretically as a set of strings \needcite{} or operationally as a parser \needcite{}.
-Alternatively, we can use type theory, so that a language is a type-level predicate on ``strings'' (lists of an arbitrary type \AB{A} of ``characters''):
+Languages are usually formalized either set-theoretically as sets of strings or operationally as parsers.
+Alternatively, one can use type theory, so that a language is a type-level predicate on ``strings'' (lists of an arbitrary type \AB{A} of ``characters''):
 \agda{Lang}
-\figrefdef{Lang-ops}{Language operations}{\agda{Lang-ops}} gives definitions of the usual language operations, including language union and intersection ({\AB{P} \AF ∪ \AB{Q}} and {\AB{P} \AF ∩ \AB{Q}}) with their identities (the empty language \AF ∅ and universal language \AF 𝒰), language concatenation ({\AB{P} \AF ⋆ \AB{Q}}) and its identity (\AF 𝟏, containing only the empty string), single-character languages ({\AF ` \AB{c}}), Kleene star ({\AB{P} \AF ☆}), and ``scalar multiplication'' (\mbox{\AB{s} \AF · \AB{P}}, which will prove useful later).
+\figrefdef{Lang-ops}{Languages as type-level predicates}{\agda{Lang-ops}} gives definitions of the usual language operations, including language union and intersection ({\AB{P} \AF ∪ \AB{Q}} and {\AB{P} \AF ∩ \AB{Q}}) with their identities (the empty language \AF ∅ and universal language \AF 𝒰), language concatenation ({\AB{P} \AF ⋆ \AB{Q}}) and its identity (\AF 𝟏, containing only the empty string), single-character languages ({\AF ` \AB{c}}), Kleene star ({\AB{P} \AF ☆}), and ``scalar multiplication'' (\mbox{\AB{s} \AF · \AB{P}}, which will prove useful later).
 \note{Explain the types used here and their logical interpretation under the Curry-Howard isomorphism.}
@@
 For language implementations of \secreftwo{Symbolic Differentiation}{Automatic Differentiation}, this anchoring is guaranteed by the type of \AF{⟦\_⟧} in \figreftwo{symbolic-api}{sized-automatic-api}.
+(Similarly, for the decidable type constructors in \figref{compositional-dec}, correctness is guaranteed by the types of \AIC{yes} and \AIC{no} in the definition of ‌\ARe{Dec} in \secref{Decidability}.)
 Given this choice of equivalence, properties of index types are easily lifted to properties of existential types, as in \figrefdef{prop}{Liftings of index properties to existential types}{\mathindent0ex\agda{prop}}.
@@
 \rnc\source{Transport}
-As examples, we can lift the commutative semiring of types to wrappings of the decidable counterparts from \figref{compositional-dec} in \secref{Decidability}, and lift both language semirings to the decidable implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation}, as in \figrefdef{transport-examples}{Examples of algebraic instances transported from indices to existential wrappings}{
-\mathindent0ex
-\agda{examples}
-}
+As examples, we can lift the commutative semiring of types to wrappings of the decidable counterparts from \figref{compositional-dec} in \secref{Decidability}, and lift both language semirings to the decidable implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation}, as in \figrefdef{transport-examples}{Examples of algebraic instances transported from indices to existential wrappings}{\agda{examples}}.
 \bibliography{bib}

### 2021-02-18 paper.tex
small corrections; acm options

@@
 
-\acmtrue
-%% \acmfalse
+%% \acmtrue
+\acmfalse
 
 \ifacm
-\documentclass[acmsmall,screen]{acmart}
+\documentclass[acmsmall,screen,anonymous,timestamp]{acmart}
 \else
@@
 The basic building blocks of type-level predicates---and languages in particular---form the vocabulary of a \emph{closed semiring} in two different ways\out{, as reflected in the structure of regular expressions \needcite{}}.
-The semiring abstraction has three aspects: (a) a commutative monoid providing ``zero'' and ``addition'', (b) a (possibly non-commutative) monoid providing ``one'' and ``multiplication'', and (c) the relationship between them, namely that multiplication distributes over addition and zero.\footnote{Distributing of multiplication over zero is also known as ``annihilation''.}
+The semiring abstraction has three aspects: (a) a commutative monoid providing ``zero'' and ``addition'', (b) a (possibly non-commutative) monoid providing ``one'' and ``multiplication'', and (c) the relationship between them, namely that multiplication distributes over addition and zero.\footnote{Distribution of multiplication over zero is also known as ``annihilation''.}
 In the first predicate semiring, which is \emph{commutative} (i.e., multiplication commutes), zero and addition are \AF ∅ and \AF{\_∪\_}, while one and multiplication are \AF 𝒰 and \AF{\_∩\_}.
@@
 \end{code}
-Referring to the type and operations of \secref{Specifying Languages} and \figref{symbolic-api} via the module prefix ``{◬.}\hspace{0.05em}'' and ``{▣.}\hspace{0.05em}'' respectively, define \agda{wrapped-Lang}
+Referring to the type and operations of \secref{Specifying Languages} and \figref{symbolic-api} via the module prefixes ``{◬.}\hspace{0.05em}'' and ``{▣.}\hspace{0.05em}'' respectively, define \agda{wrapped-Lang}
 Then wrap union as follows:

### 2021-02-19 paper.tex
date GMT-8
related work

@@
 \title{\tit}
-\date{Early draft of \DTMnow}
+\date{Early draft of \DTMnow{} GMT-8}
 
@@
 As examples, we can lift the commutative semiring of types to wrappings of the decidable counterparts from \figref{compositional-dec} in \secref{Decidability}, and lift both language semirings to the decidable implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation}, as in \figrefdef{transport-examples}{Examples of algebraic instances transported from indices to existential wrappings}{\agda{examples}}.
+
+\sectionl{Related Work}
+
+Most work on formal languages in functional programming has been in parsing, particularly via parsing combinators \needcite{}.
+Such work typically concentrates on usability, composability, and sometimes efficiency, rather than on simple, precise specification and verifiable correctness.
+
+There has also been some work using type theory to capture languages as type-level predicates much like in \secref{Specifying Languages}:
+\begin{itemize}
+
+\item \citet{KorkutEtAl2016} defined a non-indexed algebraic data type of regular expressions together with an indexed inductive type of proofs of membership of strings in the language denoted by the regular expressions and a corresponding function that matches an implicitly concatenated stack of regular expressions.
+Special attention is paid to ensuring and proving termination, particularly in regard to languages that contain the empty string.
+Addressing both involves defunctionalization and translating between general regular expressions and a restricted ``standard'' form.
+
+\item \citet{AgularMannaa2009} also defined a non-indexed algebraic type of regular expressions.
+They also \emph{defined} indexed types for \AF{ν} and \AF{δ} on regular expressions rather than defining the meanings of those operations in terms of languages and then proving properties of them.
+They do prove a consistency property about \AF{δ}, but apparently not about \AF{ν}, and they note their method's ``inability to prove the non-membership of some string in a given language''.
+
+\item \citet{DoczkalEtAl2013} formalized regular languages constructively in Coq, also in terms of a type-level membership predicate.
+They then prove several classic results about finite automata, including minimization and relationship to Nerode and Myhill partitions.
+
+\item \citet{FirsovUustalu2013} used Agda to formalize regular expressions with a corresponding inductive type of language membership proofs and then related regular expressions to nondeterministic finite automata (NFAs) represented as boolean matrices and operations on that representation that correspond to those on regular expressions.
+They then proved soundness and completeness of the NFA representation with respect to regular language membership.
+
+\item \citet{Traytel2017} defined formal languages as tries and explored operations on tries defined via primitive corecursion and reasoned about coinductively in the Isabelle/HOL proof assistant.
+There appears to be no specification higher level (less operational) than tries, leading to some rather complex definitions (which, as \emph{definitions}, cannot be proved correct).
+The definition of language concatenation is particularly delicate.
+For the same reason, proofs of even basic properties involve coinduction.
+The authors point out that, however, that the coinductive trie formulation enabled better proof automation than with sets of strings.
+
+\item \citet{Abel2016} also formulated languages via tries and in particular noted the termination issue mentioned in \secref{Automatic Differentiation} and provided the sized-types solution adopted in this paper.
+Relationships to automata theory and coalgebraic notions were also explored in depth.
+Trie look-up yields boolean values rather than propositions, and there was no simpler specification of languages with respect to which the trie implementation could be specified and proved consistent, so every operation is \emph{defined} coinductively.
+The concatenation and closure (star) cases are especially delicate.
+Equality on tries is defined by strong bisimilarity, and thus properties require corresponding machinery to state and prove.
+
+%% requiring more careful construction and reading than the specification given in \figref{Lang-ops} and used in the implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation}
+
+\end{itemize}
+
 \bibliography{bib}

### 2021-02-19 paper.tex
related work

@@
 
-A simple characterization of languages (and indeed functions from lists to any type) captures the essential idea behind language ``differentiation'' and ``integration'' as used for recognizing languages and leads to a collection of lemmas about type-level predicates.
+A simple characterization of languages (and indeed functions from lists to any type) captures the essential idea behind language ``differentiation'' (and integration) as used for recognizing languages and leads to a collection of lemmas about type-level predicates.
 These lemmas form the backbone of two dual implementations of language recognition---(inductive) regular expressions and (coinductive) tries---each containing the same code but in dual arrangements.
@@
 \agda{νδ}
+In terms of languages, {\AF{δ} \AB{P} \AB{a}} comprises all strings \AB{w} such that {\AB{a} \AIC ∷ \AB{w}} is in \AB{P}, i.e., the tails of strings in \AB{P} that start with \AB{a}.
+
 Each use of \AF{δ} takes us one step closer to reducing general language membership to nullability.
@@
 \agda{ν∘foldlδ}
-
 This simple observation is the semantic heart of the syntactic technique of \emph{derivatives of regular expressions} as used for efficient recognition of regular languages \citep{Brzozowski64}, later extended to parsing general context-free languages \citep{Might2010YaccID}.
@@
 First note that there are two categories of operations.
-One category transforms the codomain (types): \AF{∅}, \AF 𝒰, \AF{\_∪\_}, \AF{\_∩\_}.
-The other transforms the domain (lists): \AF{𝟏}, \AF{\_⋆\_}, \AF{\_☆}, \AF{`}.
+One category transforms the codomain (types): \AF{∅}, \AF 𝒰, \AF{\_∪\_}, and \AF{\_∩\_}.
+The other transforms the domain (lists): \AF{𝟏}, \AF{\_⋆\_}, \AF{\_☆}, and \AF{`}.
 The first category applies to predicates over all types (not just over lists).
@@
 
-\item \citet{KorkutEtAl2016} defined a non-indexed algebraic data type of regular expressions together with an indexed inductive type of proofs of membership of strings in the language denoted by the regular expressions and a corresponding function that matches an implicitly concatenated stack of regular expressions.
-Special attention is paid to ensuring and proving termination, particularly in regard to languages that contain the empty string.
-Addressing both involves defunctionalization and translating between general regular expressions and a restricted ``standard'' form.
-
 \item \citet{AgularMannaa2009} also defined a non-indexed algebraic type of regular expressions.
 They also \emph{defined} indexed types for \AF{ν} and \AF{δ} on regular expressions rather than defining the meanings of those operations in terms of languages and then proving properties of them.
-They do prove a consistency property about \AF{δ}, but apparently not about \AF{ν}, and they note their method's ``inability to prove the non-membership of some string in a given language''.
+They prove a consistency property about \AF{δ}, but apparently not about \AF{ν}, and they note their method's ``inability to prove the non-membership of some string in a given language''.
 
@@
 They then proved soundness and completeness of the NFA representation with respect to regular language membership.
+
+\item \citet{KorkutEtAl2016} defined a non-indexed algebraic data type of regular expressions together with an indexed inductive type of proofs of membership of strings in the language denoted by the regular expressions and a corresponding function that matches an implicitly concatenated stack of regular expressions.
+Special attention is paid to ensuring and proving termination, particularly in regard to languages that contain the empty string.
+Addressing both involves defunctionalization and translating between general regular expressions and a restricted ``standard'' form.
 

### 2021-02-19 paper.tex
contributions / Brzozowski

@@
 As described in \secref{Properties}, it's also fairly easy to prove algebraic properties of predicates\out{, including the semimodule and semiring laws for both predicate semirings}.
-These definitions do not, however, give us decidable parsing, because general type inhabitation amounts to theorem proving.
+These definitions do not give us decidable parsing, because general type inhabitation amounts to theorem proving.
+As shown in the rest of this paper, however, type-level predicates serve as a simple, precise, and non-operational basis, with compelling and correct implementations following easily and systematically as corollaries.
+
+This paper makes the following specific contributions:
+\begin{itemize}
+\item
+  A collection of lemmas are stated (and proved) that capture the semantic essence of Brzozowski's syntactic technique of derivatives of regular expressions, in terms of functions over lists.
+\item
+  A duality is revealed between regular expressions and tries in the context of language differentiation.
+  Each implementation contains exactly the same code but in transposed forms.
+  Both are corollaries of the semantic lemmas and are automatically proved correct by type-checking.
+\item
+  The theory and algorithms are specified in terms of proof isomorphism rather than mere logical equivalence.
+  As such, the differentiation lemmas for language concatenation and Kleene closure are more informative than in previous work, capturing \emph{all} parsings, including closure of languages containing the empty string.
+\item
+  The vocabulary and reasoning of languages is generalized to predicates over arbitrary types, and then re-specialized to the usual vocabulary.
+  This generalization consists of two generally useful covariant applicative functors on functions (the usual one on functions-from-X and another on functions-to-Set).
+  Properties are proved in terms of this more general vocabulary.
+\item
+  Properties are defined first at the level of predicates, making both easy to understand and prove, thanks to elimination of operational details.
+  These properties are then transported effortlessly to the computable parsing implementations.
+\end{itemize}
+The source code for this paper is available and contains fully verified Agda code.
+All type signatures and code in the paper (including all lemmas and algorithms whether shown in the paper or elided) are generated by the Agda compiler with no postulates.
 
@@
 
+Of the previous work involving language derivatives, none appear to use the essential, non-inductive, nearly-trivial definition in terms of functions of lists from \secref{Decomposing Languages}, although \citet{Abel2016} did indeed mention this definition informally as motivation for the more complex, coinductively defined operations on tries.
+\citet{Brzozowski64} also made clear mention of the underlying meaning on sets of strings in his original work on regular expression derivatives.
+
 \end{itemize}

### 2021-02-19 paper.tex
more rfcs

@@
 \figrefdef{Lang-ops}{Languages as type-level predicates}{\agda{Lang-ops}} gives definitions of the usual language operations, including language union and intersection ({\AB{P} \AF ∪ \AB{Q}} and {\AB{P} \AF ∩ \AB{Q}}) with their identities (the empty language \AF ∅ and universal language \AF 𝒰), language concatenation ({\AB{P} \AF ⋆ \AB{Q}}) and its identity (\AF 𝟏, containing only the empty string), single-character languages ({\AF ` \AB{c}}), Kleene star ({\AB{P} \AF ☆}), and ``scalar multiplication'' (\mbox{\AB{s} \AF · \AB{P}}, which will prove useful later).
-\note{Explain the types used here and their logical interpretation under the Curry-Howard isomorphism.}
+
+The definitions in \figref{Lang-ops} reflect the logical interpretation of types under the Curry-Howard (``propositions as types'') isomorphism, in which types represent propositions and elements/values of a type represent proofs of the corresponding proposition \citep{Wadler2015propositions}.
+The specific embodiment of Curry-Howard in \figref{Lang-ops} and throughout this paper is the language Agda \citep{Norell2009,BoveEtAl2009Agda}, which is both programming language and proof assistant founded on Martin-Löf's intuitionistic type theory \citep{MartinLöf1984intuitionistic}.
+Specifically,
+\begin{itemize}
+
+\item The types \AD ⊥ and \ARe ⊤ represent falsity and truth respectively, with \AD ⊥ having no elements (i.e., uninhabited) and \ARe ⊤ having a single element \AFi{tt}.
+
+\item The types {\AB A \AD ⊎ \AB B} and {\AB A \ARe × \AB B} represent disjunction and conjunction respectively as non-dependent sum and products.
+
+\item For a function {\AB F \AS : \AB A \AS → \APT{Set}}, the types {\AF ∃ \AB F}  and {\AF ∀ \AB x \AS → \AB F \AB x} represent existential and universal quantification as \emph{dependent} product and function types.
+
+\item The type {\AB x \AD ≡ \AB y} represents \emph{propositional} equality (rather than computable equality) as a data type with the single constructor {\AIC{refl} \AS : \AS ∀ \AB x \AS → \AB x \AD ≡ \AB x}.
+
+\end{itemize}
 
@@
 \end{itemize}
-The source code for this paper is available and contains fully verified Agda code.
+The source code for this paper is freely available and contains fully verified Agda code.
 All type signatures and code in the paper (including all lemmas and algorithms whether shown in the paper or elided) are generated by the Agda compiler with no postulates.
@@
 
-In order to convert languages into parsers, it will help to decompose languages---and indeed any function from lists---into a regular form.
-Since lists are defined by an algebraic data type, we can decompose languages according to constructor, as follows:
+In order to convert languages into parsers, it will help to decompose languages into a regular form.
+Since lists are defined by an algebraic data type, we can decompose languages---and indeed any function from lists---according to constructor, as follows:
 \agda{νδ}
@@
 Each use of \AF{δ} takes us one step closer to reducing general language membership to nullability.
-In other words,\footnote{Repeated application is expressed as a standard left fold\out{ \stdlibCitep{Data.List}}:
-\agda{foldl}
+In other words,\footnote{For functions \AB{f} and \AB{g}, {\AB{f} \AF{≗} \AB{g}} is extensional equality, i.e., {\AS ∀ \AB x \AS → \AB{f} \AB{x} \AD{≡} \AB{g} \AB{x}}.
+Repeated application is expressed as a standard left fold\out{ \stdlibCitep{Data.List}}:
+\agda{foldl}\vspace{-2ex}
 }%

### 2021-02-19 paper.tex
Fill in missing references and explanations flagged in the text.

@@
 \item
-  A collection of lemmas are stated (and proved) that capture the semantic essence of Brzozowski's syntactic technique of derivatives of regular expressions, in terms of functions over lists.
+  A collection of lemmas are stated (and proved) about functions over lists, capturing the semantic essence of Brzozowski's syntactic technique of derivatives of regular expressions.
 \item
@@
 \emph{Automata theory}: {\AF{ν} \AF ∘ \AF{foldl} \AF{δ} \AB{P}} is the execution of a state machine.
-Each state is a language, with \AB{P} being the initial state; \AF{δ} is the state transition function; and \AB{ν} is the set of accepting states \needcite{}.
+Each state is a language, with \AB{P} being the initial state; \AF{δ} is the state transition function; and \AB{ν} is the set of accepting states\out{ \needcite{}}.
 Lemma \AF{ν∘foldlδ} captures the correctness of this state machine as a recognizer for the language \AB{P}.
@@
 \item
-\emph{Incremental computation}: {\AF{ν} \AF ∘ \AF{foldl} \AF{δ} \AB{P}} computes \AB{P} by transforming successive incremental modifications to the empty string (each in the form of {\AB{a} \AIC ∷\_} for a list element \AB{a}) into incremental modifications to the membership question for the related language {\AF{δ} \AB{P} \AB{a}} \needcite{}.
+\emph{Incremental computation}: {\AF{ν} \AF ∘ \AF{foldl} \AF{δ} \AB{P}} computes \AB{P} by transforming successive incremental modifications to the empty string (each in the form of {\AB{a} \AIC ∷\_} for a list element \AB{a}) into incremental modifications to the membership question for the related language {\AF{δ} \AB{P} \AB{a}}\out{ \needcite{}}.
 Lemma \AF{ν∘foldlδ} again captures the correctness of this incremental implementation of \AB{P}.
@@
 \item 
-\emph{Calculus}: \AB{δ} is differentiation; {\AF{ν} \AF ∘ \AF{foldl} \AF{δ} \AB{P}} is integration; and lemma \AF{ν∘foldlδ} is the second fundamental theorem of calculus, which says that a function can be recovered by integrating its derivative \needcite{}.
+\emph{Calculus}: \AB{δ} is differentiation; {\AF{ν} \AF ∘ \AF{foldl} \AF{δ} \AB{P}} is integration; and lemma \AF{ν∘foldlδ} is the second fundamental theorem of calculus, which says that a function can be recovered by integrating its derivative\out{ \needcite{}}.
 
@@
 \agda{Dec}
-For compositionality, we will use a few operations that lift decidability of types to decidability of constructions on those types, as shown in \figrefdef{compositional-dec}{Compositional decidability}{\agda{compositional-dec}}.\notefoot{Explanatory footnote goes here.}
+For compositionality, we will use a few operations that lift decidability of types to decidability of constructions on those types, as shown in \figrefdef{compositional-dec}{Compositional decidability}{\agda{compositional-dec}}.\footnote{%
+A few explanatory remarks:
+\begin{itemize}
+
+\item {\AS λ \AS{()}} is the unique function from \AD ⊥ to any type (here also {\AD ⊥}).
+
+\item \AF{inj₁} and \AF{inj₂} are left and right injections into a sum type.
+
+\item \AF{proj₁} and \AF{proj₂} are left and right projections out of a product type.
+
+\item Given two functions {\AB f \AS : \AB A \AS → \AB C} and {\AB g \AS : \AB B \AS → C}, the function {\AF [ \AB f \AF , \AB g \AF ] \AS : \AB A \AF ⊎ \AB B \AS → \AB C} maps {\AIC{inj₁} \AB a} to {\AB f \AB a} and {\AIC{inj₂} \AB b} to {\AB g \AB b}.
+In this use, {\AB C} is \AD ⊥, since we are proving a negation.
+
+\item The \AF{\_✶‽} definition reflects the fact that the empty list exists for every element type (even \AF ⊥).
+
+\end{itemize}
+
+}
 Moreover, decidability lifts naturally from types to predicates:
@@
 In the former, functions are represented symbolically in some suitable vocabulary, and pattern-matching is used to apply rules of differentiation to those representations.
-The latter solution involves re-interpreting the vocabulary of functions to construct not just the usual functions, but also their derivatives \needcite{}.
+The latter solution involves re-interpreting the vocabulary of functions to construct not just the usual functions, but also their derivatives \citep{Griewank89onAD, GriewankWalther2008EvalDerivs, Elliott-2018-ad-icfp}.
 The latter is often much more efficient than the former, since it easily avoids a good deal of redundant computation due to significant amounts of duplicated work between function ``primals'' and their derivatives.
@@
 
-Decidable recognition (via {\AF{⟦\_⟧}}) relies only on \AF{ν} and \AF{δ}, which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\notefoot{Describe this derivation.}
+Decidable recognition (via {\AF{⟦\_⟧}}) relies only on \AF{ν} and \AF{δ}, which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\out{\notefoot{Describe this derivation.}}
 These definitions are meant to be read in ``column-major'' order, i.e., each column is one function definition.
@@
 Now consider the dual strategy of ``automatic differentiation''.
-This time, reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type (of tries \needcite{}), while redefining the language-building vocabulary as functions on that data type defined by \emph{copatterns} \citep{AbelPientka2016}.
+This time, reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type of tries, while redefining the language-building vocabulary as functions on that data type defined by \emph{copatterns} \citep{AbelPientka2016}.\footnote{The classic trie (``prefix tree'') data structure was introduced by Thue in 1912 to represent sets of strings \citep[Section 6.3]{Knuth1998ACP3}, later generalized to arbitrary (regular) algebraic data types \citep{Connelly1995GenTrie}, and then from sets to functions \citep{Hinze2000GGT}.}
 Again, we will use the same names as in \secref{Specifying Languages}, referring to the original versions via the module prefix ``{◬.}\hspace{0.05em}''.
@@
 
-Looking more closely, additional algebraic structure emerges on predicates: (full) \emph{semimodules} generalize vector spaces by relaxing the associated types of ``scalars'' from fields to commutative semirings, i.e., dropping the requirements of multiplicative and additive inverses \needcite{}.
+Looking more closely, additional algebraic structure emerges on predicates: (full) \emph{semimodules} generalize vector spaces by relaxing the associated types of ``scalars'' from fields to commutative semirings, i.e., dropping the requirements of multiplicative and additive inverses.
 Left and right semimodules further generalize scalars to arbitrary semirings, dropping the commutativity requirement for scalar multiplication.
@@
 There is still more algebraic structure to be found and exploited.
-When our \AB{domain} type \AB{A} is a monoid (as with languages, infinite streams and grids, and functions of continuous space and time), predicates over \AB{A} form a second semiring, known as ``the monoid semiring'' \needcite{}, in which zero and addition are as in the first predicate commutative semiring and semimodule, while one and multiplication are \AF{𝟏} and \AF{\_⋆\_}, defined in \secref{Specifying Languages}.
-In this setting, language concatenation is subsumed by a very general form of \emph{convolution} \needcite{}.
+When our \AB{domain} type \AB{A} is a monoid (as with languages, infinite streams and grids, and functions of continuous space and time), predicates over \AB{A} form a second semiring, known as ``the monoid semiring'' \citep{Golan2005RecentSemi}, in which zero and addition are as in the first predicate commutative semiring and semimodule, while one and multiplication are \AF{𝟏} and \AF{\_⋆\_}, defined in \secref{Specifying Languages}.
+In this setting, language concatenation is subsumed by a very general form of \emph{convolution} \cite{Golan2005RecentSemi, Elliott2019-convolution}.
 The semimodule structure of predicates provides the additive aspect and is built entirely from the \emph{codomain}-transforming predicate operations defined in \secref{Specifying Languages}.
@@
 
-Most work on formal languages in functional programming has been in parsing, particularly via parsing combinators \needcite{}.
+Most work on formal languages in functional programming has been in parsing, particularly via parsing combinators \citep{Hutton1996monadic, Leijen2001parsec, Swierstra2008combinator}.
 Such work typically concentrates on usability, composability, and sometimes efficiency, rather than on simple, precise specification and verifiable correctness.

### 2021-02-19 paper.tex
wording in the abstract

@@
 
-A simple characterization of languages (and indeed functions from lists to any type) captures the essential idea behind language ``differentiation'' (and integration) as used for recognizing languages and leads to a collection of lemmas about type-level predicates.
-These lemmas form the backbone of two dual implementations of language recognition---(inductive) regular expressions and (coinductive) tries---each containing the same code but in dual arrangements.
+A simple characterization of languages (and indeed functions from lists to any type) captures the essential idea behind language ``differentiation'' (and integration) as used for recognizing languages, leading to a collection of lemmas about type-level predicates.
+These lemmas are the heart of two dual implementations of language recognition---(inductive) regular expressions and (coinductive) tries---each containing the same code but in dual arrangements.
 The language lemmas form most of the implementation in both cases, while the representation and primitive operations trade places.

### 2021-02-20 paper.tex
explicit universe levels

@@
 Languages are usually formalized either set-theoretically as sets of strings or operationally as parsers.
-Alternatively, one can use type theory, so that a language is a type-level predicate on ``strings'' (lists of an arbitrary type \AB{A} of ``characters''):
-\agda{Lang}
+Alternatively, one can use type theory, so that a language is a type-level predicate on ``strings'' (lists of an arbitrary type \AB{A} of ``characters'').
 \figrefdef{Lang-ops}{Languages as type-level predicates}{\agda{Lang-ops}} gives definitions of the usual language operations, including language union and intersection ({\AB{P} \AF ∪ \AB{Q}} and {\AB{P} \AF ∩ \AB{Q}}) with their identities (the empty language \AF ∅ and universal language \AF 𝒰), language concatenation ({\AB{P} \AF ⋆ \AB{Q}}) and its identity (\AF 𝟏, containing only the empty string), single-character languages ({\AF ` \AB{c}}), Kleene star ({\AB{P} \AF ☆}), and ``scalar multiplication'' (\mbox{\AB{s} \AF · \AB{P}}, which will prove useful later).
@@
 The definitions in \figref{Lang-ops} reflect the logical interpretation of types under the Curry-Howard (``propositions as types'') isomorphism, in which types represent propositions and elements/values of a type represent proofs of the corresponding proposition \citep{Wadler2015propositions}.
-The specific embodiment of Curry-Howard in \figref{Lang-ops} and throughout this paper is the language Agda \citep{Norell2009,BoveEtAl2009Agda}, which is both programming language and proof assistant founded on Martin-Löf's intuitionistic type theory \citep{MartinLöf1984intuitionistic}.
+The specific embodiment of Curry-Howard in \figref{Lang-ops} and throughout this paper is the language Agda \citep{Norell2008AFP, BoveEtAl2009Agda}, which is both programming language and proof assistant founded on Martin-Löf's intuitionistic type theory \citep{MartinLöf1984intuitionistic}.
 Specifically,
 \begin{itemize}
+
+\item Types have type {\APT{Set} \AB ℓ} for some universe level \AB ℓ.
+These levels avoid logical inconsistencies resulting from {\APT{Set} \AS : \APT{Set}} but can be safely ignored in this paper (which is parametrized over \AB ℓ).
 
@@
 For instance, {\AF{∃} \AD{Dec}} and {\AF{∃} \AD{(\ARe{Lang} \APo ∞)}} are simple types.
-For a type family {\AB F \AK : \AB A \AK → \APT Set⇃}, values of type {\AF ∃ \AB F} are pairs {(\AB a , \AB b)} with {\AB a \AK : \AB A} and {\AB b \AK : \AB F \AB a}.
+For a type family {\AB F \AK : \AB A \AK → \APT Set \AB ℓ}, values of type {\AF ∃ \AB F} are pairs {(\AB a , \AB b)} with {\AB a \AK : \AB A} and {\AB b \AK : \AB F \AB a}.
 We can also encapsulate dependently typed \emph{operations} on indexed types into simply typed operations on the (simply typed) existentially wrapped versions of those types.

### 2021-02-20 paper.tex
predicate operations spacing
proofreading pass

@@
 \setmathfont{XITSMath-Regular.otf}
-\newunicodechar{◬}{\raisebox{0.23ex}{\ensuremath{_{\diamond}\hspace{-0.352em}}}}
+\newunicodechar{◬}{\raisebox{0.23ex}{\ensuremath{_{\diamond}\hspace{-0.358em}}}}
 \newunicodechar{▣}{\raisebox{-0.13ex}{\ensuremath{\scaleobj{0.3}{\Box}}\hspace{-0.23em}}}
@@
 Alternatively, one can use type theory, so that a language is a type-level predicate on ``strings'' (lists of an arbitrary type \AB{A} of ``characters'').
-\figrefdef{Lang-ops}{Languages as type-level predicates}{\agda{Lang-ops}} gives definitions of the usual language operations, including language union and intersection ({\AB{P} \AF ∪ \AB{Q}} and {\AB{P} \AF ∩ \AB{Q}}) with their identities (the empty language \AF ∅ and universal language \AF 𝒰), language concatenation ({\AB{P} \AF ⋆ \AB{Q}}) and its identity (\AF 𝟏, containing only the empty string), single-character languages ({\AF ` \AB{c}}), Kleene star ({\AB{P} \AF ☆}), and ``scalar multiplication'' (\mbox{\AB{s} \AF · \AB{P}}, which will prove useful later).
+The usual language operations are defined in \figrefdef{Lang-ops}{Languages as type-level predicates}{\agda{Lang-ops}}, including language union and intersection ({\AB{P} \AF ∪ \AB{Q}} and {\AB{P} \AF ∩ \AB{Q}}) with their identities (the empty language \AF ∅ and universal language \AF 𝒰), language concatenation ({\AB{P} \AF ⋆ \AB{Q}}) and its identity (\AF 𝟏, containing only the empty string), single-character languages ({\AF ` \AB{c}}), Kleene star ({\AB{P} \AF ☆}), and ``scalar multiplication'' (\mbox{\AB{s} \AF · \AB{P}}, which will prove useful later).
 
-The definitions in \figref{Lang-ops} reflect the logical interpretation of types under the Curry-Howard (``propositions as types'') isomorphism, in which types represent propositions and elements/values of a type represent proofs of the corresponding proposition \citep{Wadler2015propositions}.
+The definitions in \figref{Lang-ops} reflect the logical interpretation of types under the Curry-Howard (``propositions as types'') isomorphism, in which types represent propositions and element (values) of a type represent proofs of the corresponding proposition \citep{Wadler2015propositions}.
 The specific embodiment of Curry-Howard in \figref{Lang-ops} and throughout this paper is the language Agda \citep{Norell2008AFP, BoveEtAl2009Agda}, which is both programming language and proof assistant founded on Martin-Löf's intuitionistic type theory \citep{MartinLöf1984intuitionistic}.
@@
 
-\item For a function {\AB F \AS : \AB A \AS → \APT{Set}}, the types {\AF ∃ \AB F}  and {\AF ∀ \AB x \AS → \AB F \AB x} represent existential and universal quantification as \emph{dependent} product and function types.
+\item For a function {\AB F \AS : \AB A \AS → \APT{Set} \AB{ℓ}}, the types {\AF ∃ \AB F}  and {\AF ∀ \AB x \AS → \AB F \AB x} represent existential and universal quantification as \emph{dependent} product and function types.
 
@@
 
-Note that for a language \AB{P} and string \AB{w}, {\AB{P} \AB{w}} is the type of \emph{proofs} that {\AB{w} ∈ \AB{P}}, in other words \emph{explanations} of membership, or \emph{parsing}.
+Note that for a language \AB{P} and string \AB{w}, {\AB{P} \AB{w}} is the type of \emph{proofs} that {\AB{w} ∈ \AB{P}}, in other words \emph{explanations} of membership, or \emph{parsings}.
 (If the type {\AB{P} \AB{w}} is uninhabited, then {\AB{w} ∉ \AB{P}}.)
@@
 These definitions do not give us decidable parsing, because general type inhabitation amounts to theorem proving.
-As shown in the rest of this paper, however, type-level predicates serve as a simple, precise, and non-operational basis, with compelling and correct implementations following easily and systematically as corollaries.
+As shown in the rest of this paper, however, type-level predicates serve as a simple, precise, and non-operational basis for specification and reasoning, with correct and computable parsing implementations following systematically as corollaries.
 
@@
 \item
-  A collection of lemmas are stated (and proved) about functions over lists, capturing the semantic essence of Brzozowski's syntactic technique of derivatives of regular expressions.
+  A collection of lemmas are stated and proved about functions over lists, capturing the semantic essence of Brzozowski's syntactic technique of derivatives of regular expressions (\secref{Decomposing Languages}).
 \item
-  A duality is revealed between regular expressions and tries in the context of language differentiation.
+  A duality is revealed between regular expressions and tries (prefix trees) in the context of language differentiation (\secreftwo{Symbolic Differentiation}{Automatic Differentiation}).
   Each implementation contains exactly the same code but in transposed forms.
@@
 \item
-  The vocabulary and reasoning of languages is generalized to predicates over arbitrary types, and then re-specialized to the usual vocabulary.
-  This generalization consists of two generally useful covariant applicative functors on functions (the usual one on functions-from-X and another on functions-to-Set).
-  Properties are proved in terms of this more general vocabulary.
+  The vocabulary of languages is generalized to predicates over arbitrary types, and then re-specialized to the usual vocabulary (\secref{Predicate Algebra}).
+  This generalization consists of two useful covariant applicative functors on functions (the usual one on functions-from-\AF{X} and another on functions-to-\APT{Set}).
+  % Properties are proved in terms of this more general vocabulary.
 \item
-  Properties are defined first at the level of predicates, making both easy to understand and prove, thanks to elimination of operational details.
-  These properties are then transported effortlessly to the computable parsing implementations.
+  Properties are stated first at the level of predicates, making them easy to understand and prove, thanks to elimination of operational details.
+  These properties are then transported effortlessly to the computable parsing implementations (\secref{Transporting Properties}).
 \end{itemize}
-The source code for this paper is freely available and contains fully verified Agda code.
-All type signatures and code in the paper (including all lemmas and algorithms whether shown in the paper or elided) are generated by the Agda compiler with no postulates.
+Although many proofs are elided below, the source code for this paper is freely available and contains fully verified Agda code with no postulates.
+All type signatures and definitions in the paper are generated by the Agda compiler.
 
@@
 \agda{νδ}
-In terms of languages, {\AF{δ} \AB{P} \AB{a}} comprises all strings \AB{w} such that {\AB{a} \AIC ∷ \AB{w}} is in \AB{P}, i.e., the tails of strings in \AB{P} that start with \AB{a}.
+In terms of languages, {\AF{δ} \AB{P} \AB{a}} comprises all strings \AB{w} such that {\AB{a} \AIC ∷ \AB{w}} is in \AB{P}\out{, i.e., the tails of strings in \AB{P} that start with \AB{a}}.
 
@@
 
+The equalities in \figref{nu-delta-lemmas} are all proved automatically by normalization (i.e., their proofs are simply \AIC{refl}), while the other relations require a bit more work.
 As with all lemmas in this paper shown with signatures only, full proofs are in the paper's source code and are formally verified by the Agda compiler.
-The equalities in \figref{nu-delta-lemmas} are all proved automatically by normalization (i.e., their proofs are simply \AIC{refl}), while the other relations require a bit more work.
 
@@
 
-For an effective implementation, we will have to bridge the gap between a type (of membership proofs) and its decidable inhabitation.
+For an effective implementation, we must bridge the gap between a type (of membership proofs) and its decidable inhabitation.
 Fortunately, there is a convenient and compositional way to do so.
@@
 
-\item {\AS λ \AS{()}} is the unique function from \AD ⊥ to any type (here also {\AD ⊥}).
+\item {\AS λ \AS{()}} is the unique function from \AD ⊥ to any type (here also {\AD ⊥}),  sometimes called ``\AF{absurd}'' or ``\AF{⊥-elim}''.
 
@@
 
-\item Given two functions {\AB f \AS : \AB A \AS → \AB C} and {\AB g \AS : \AB B \AS → C}, the function {\AF [ \AB f \AF , \AB g \AF ] \AS : \AB A \AF ⊎ \AB B \AS → \AB C} maps {\AIC{inj₁} \AB a} to {\AB f \AB a} and {\AIC{inj₂} \AB b} to {\AB g \AB b}.
-In this use, {\AB C} is \AD ⊥, since we are proving a negation.
+\item For functions {\AB f \AS : \AB A \AS → \AB C} and {\AB g \AS : \AB B \AS → C}, the function {\AF [ \AB f \AF , \AB g \AF ] \AS : \AB A \AF ⊎ \AB B \AS → \AB C} maps {\AIC{inj₁} \AB a} to {\AB f \AB a} and {\AIC{inj₂} \AB b} to {\AB g \AB b}.
+% In this use, {\AB C} is \AD ⊥, since we are proving a negation.
 
-\item The \AF{\_✶‽} definition reflects the fact that the empty list exists for every element type (even \AF ⊥).
+\item \AF{\_✶‽} reflects the fact that the empty list exists for every element type (even \AF ⊥).
 
@@
 These lemmas and definitions cannot be applied \emph{automatically} in their present form, however, because languages are functions, as are \AF{ν} and \AF{δ}, and so are not subject to pattern-matching.
-An automatic solution would need some form of \emph{reflection} of language operations or \AF{ν} and \AF{δ} as inspectable data.
+An automatic solution would need some form of \emph{reflection} of language operations or of \AF{ν} and \AF{δ} as inspectable data.
 
 This situation is exactly as in differential calculus, since differentiation in that setting is also defined on functions rather than on symbolic representations.
-Fortunately, there are two standard solutions for \emph{computable} differentiation, going by the names of ``symbolic'' and ``automatic'' differentiation.\out{\footnote{There is also an incorrect (approximate) variation often referred to as ``numeric'' differentiation.}}
-In the former, functions are represented symbolically in some suitable vocabulary, and pattern-matching is used to apply rules of differentiation to those representations.
+Fortunately, there are two standard solutions for \emph{computable} differentiation, commonly referred to as ``symbolic'' and ``automatic'' differentiation.\out{\footnote{There is also an incorrect (approximate) variation often referred to as ``numeric'' differentiation.}}
+In the former, functions are represented symbolically in some suitable vocabulary, and pattern-matching is used to apply rules of differentiation to those symbolic representations.
 The latter solution involves re-interpreting the vocabulary of functions to construct not just the usual functions, but also their derivatives \citep{Griewank89onAD, GriewankWalther2008EvalDerivs, Elliott-2018-ad-icfp}.
-The latter is often much more efficient than the former, since it easily avoids a good deal of redundant computation due to significant amounts of duplicated work between function ``primals'' and their derivatives.
+This latter option is often much more efficient than the former, since it easily avoids a good deal of redundant computation due to significant amounts of duplicated work between function ``primals'' and their derivatives.
 
@@
 
-The ``symbolic differentiation'' style reflects language-building vocabulary (\figref{Lang-ops}) into an inductive data type (of modestly extended regular expressions), while re-defining \AF{ν} and \AF{δ} as functions that pattern-match on that data type.
-For convenience, use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the module prefix ``{◬.}\hspace{0.05em}''.
-The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\api} along with a semantic function {\AF{⟦\_⟧}} that converts a symbolic language representation to decidable form.
+The ``symbolic differentiation'' style reflects language-building vocabulary (\figref{Lang-ops}) into an inductive data type (of modestly extended regular expressions), while \AF{ν} and \AF{δ} become functions defined by pattern-matching on that data type.
+For convenience, we can use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the module prefix ``{◬.}\hspace{0.05em}''.
+The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\api} along with a function {\AF{⟦\_⟧}} that converts a symbolic language representation to a computable parser.
 
-Decidable recognition (via {\AF{⟦\_⟧}}) relies only on \AF{ν} and \AF{δ}, which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\out{\notefoot{Describe this derivation.}}
+Decidable parsing (via {\AF{⟦\_⟧}}) relies only on \AF{ν} and \AF{δ}, which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\out{\notefoot{Describe this derivation.}}
 These definitions are meant to be read in ``column-major'' order, i.e., each column is one function definition.
 Correctness is guaranteed by the types of \AF{ν} and \AF{δ} and so is proved automatically by type-checking \figref{symbolic-defs}.
+(We could thus use \emph{any} definitions of \AF{ν} and \AF{δ} that type- and termination-check, although not many definitions would.)
+
+In \figref{symbolic-defs}, note the role of the isomorphisms from \figref{nu-delta-lemmas}.
+Since those isomorphisms relate \AF{ν} and \AF{δ} on the language-building operations to types and languages expressed in that \emph{same} vocabulary, we end up with a recursive algorithm.
 
@@
 Now consider the dual strategy of ``automatic differentiation''.
-This time, reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type of tries, while redefining the language-building vocabulary as functions on that data type defined by \emph{copatterns} \citep{AbelPientka2016}.\footnote{The classic trie (``prefix tree'') data structure was introduced by Thue in 1912 to represent sets of strings \citep[Section 6.3]{Knuth1998ACP3}, later generalized to arbitrary (regular) algebraic data types \citep{Connelly1995GenTrie}, and then from sets to functions \citep{Hinze2000GGT}.}
+This time, reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type of tries\footnote{The classic trie (``prefix tree'') data structure was introduced by \citet{Thue1912Gegenseitige} to represent sets of strings \citep[Section 6.3]{Knuth1998ACP3}, later generalized to arbitrary non-nested algebraic data types \citep{Connelly1995GenTrie}, and then from sets to functions \citep{Hinze2000GGT}.}, while redefining the language-building vocabulary as functions on that data type defined by \emph{copatterns} \citep{AbelPientka2016}.
 Again, we will use the same names as in \secref{Specifying Languages}, referring to the original versions via the module prefix ``{◬.}\hspace{0.05em}''.
@@
 
-The correctness proof is \emph{almost} accomplished by type-checking, but there is a technical problem.
-The \AF{δ} clause for {\AB{p} \AF ⋆ \AB{q}} does not satisfy Agda's termination checker, which cannot see that the argument {\AF{δ} \AB{p} \AB{a}} in the recursive use of \AF{\_⋆\_} is in some sense smaller than \AB{p}.
+The correctness proof is \emph{almost} accomplished by type- and termination-checking, but a technical problem arises.
+The {\AF{δ} (\AB{p} \AF ⋆ \AB{q})} clause does not satisfy Agda's termination checker, which cannot see that the argument {\AF{δ} \AB{p} \AB{a}} in the recursive use of \AF{\_⋆\_} is in some sense smaller than \AB{p}.
+(\figref{automatic-defs} compiles only due to suppressing termination checking for the problematic clause with a compiler pragma.)
 Fortunately, this issue was already identified and solved by Andreas \citet{Abel2016}---also in the setting of trie-based language recognition---by using \emph{sized types} \citep{Abel2008, AbelPientka2016}.
 \rnc\source{SizedAutomatic}
-This solution needs only to give \AF{Lang} (now tries) an index {\AB{i} \AK : \APT{Size}} corresponding to the maximum depth to which a trie can be searched, or equivalently the longest string that can be matched\out{ (via {\AF{⟦\_⟧}})}.
+This solution only requires giving \AF{Lang} (now tries) an index {\AB{i} \AK : \APT{Size}} corresponding to the maximum depth to which a trie can be searched, or equivalently the longest string that can be matched\out{ (via {\AF{⟦\_⟧}})}.
 In practice, we will work with arbitrarily deep tries, i.e., ones having index \APo{∞}, as in the type of {\AF{⟦\_⟧}}.
 The modified representation is shown in \figrefdef{sized-automatic-api}{Sized tries (coinductive)}{\api}.
-
-Decidable recognition is defined exactly as with unsized tries (\figref{automatic-defs}), and this time the compiler successfully proves \emph{total} correctness (including termination).
+Decidable recognition is defined exactly as with unsized tries (\figref{automatic-defs}), with the sole exception of removing the compiler pragma that suppressed termination checking.
+This time the compiler successfully proves \emph{total} correctness (including termination).
 
@@
 The first category applies to predicates over all types (not just over lists).
-Within the second category, \AF{𝟏}, \AF{\_⋆\_}, and \AF{\_☆} apply to all monoids, while \AF{`} is specific to lists.
+Within the second category, \AF{𝟏}, \AF{\_⋆\_}, and \AF{\_☆} apply to all monoids\footnote{The \emph{definitions} of those operations do not require the monoid properties, but their properties will (\secref{Predicate Algebra}).}, while \AF{`} is specific to lists.
 
@@
 \mathindent0ex
-\begin{minipage}[b]{17em}
-%% Transform codomain: %\small
+\begin{minipage}[b]{19em}
 \agda{codomain-transformers}
@@
 \hfill
-\begin{minipage}[b]{20em}
-%% Transform domain: %\small
+\begin{minipage}[b]{19em}
 \agda{domain-transformers}
@@
 \hfill\;
-}, which also includes language complement (\AF ∁).
+}, which adds language complement (\AF ∁).
 
 \rnc\source{Calculus}
-The \AF{ν} and \AF{δ} lemmas for codomain operations in \figref{nu-delta-lemmas} are easily generalized as shown in \figrefdef{nu-delta-codomain-lemmas}{Properties of \AF{ν} and \AF{δ} for predicate codomain operations}{\agda{νδ-codomain}} and are all proved automatically by normalization.
-(In contrast, the \AF{ν} and \AF{δ} lemmas do not easily generalize for domain operations.)
-Since the decidable language implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation} are corollaries of \AF{ν} and \AF{δ} lemmas, those implementations adapt easily to the generalized codomain operations (\AF{pureᵀ}, \AF{mapᵀ}, and \AF{mapᵀ₂}), resulting in a somewhat smaller implementation with broader coverage code.
+The \AF{ν} and \AF{δ} lemmas for codomain (but not domain) operations in \figref{nu-delta-lemmas} are easily generalized as shown in \figrefdef{nu-delta-codomain-lemmas}{Properties of \AF{ν} and \AF{δ} for predicate codomain operations}{\agda{νδ-codomain}} and are all proved automatically by normalization.
+Since the decidable language implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation} are corollaries of \AF{ν} and \AF{δ} lemmas, those implementations adapt easily to the generalized codomain operations (\AF{pureᵀ}, \AF{mapᵀ}, and \AF{mapᵀ₂}), resulting in a somewhat smaller implementation and broader coverage.
 
@@
 
-Conveniently, booleans and types form commutative semirings with all necessary definitions already in the standard library.
+Conveniently, booleans and types form commutative semirings with all necessary proofs already in the standard library.
 (The equivalence relation used for types is isomorphism rather than equality.)
@@
 Likewise for other algebraic structures.
+(As always, proofs are in this paper's source code.)
 
@@
 The dimension of the semimodule is the cardinality of the domain type \AB{A} (typically infinite), with ``basis vectors'' having the form {\AF{pureⱽ} \AB{a}} for {\AB{a} \AS : \AB{A}}.
-A general vector (predicate) is a linear combination (often infinite) of these basis vectors, with addition being union and scaling conjoining membership proofs.
+A general vector (predicate) is a linear combination (often infinite) of these basis vectors, with addition being union and scaling defined by conjoining membership proofs (as in \figref{lang-via-pred}).
 
 There is still more algebraic structure to be found and exploited.
-When our \AB{domain} type \AB{A} is a monoid (as with languages, infinite streams and grids, and functions of continuous space and time), predicates over \AB{A} form a second semiring, known as ``the monoid semiring'' \citep{Golan2005RecentSemi}, in which zero and addition are as in the first predicate commutative semiring and semimodule, while one and multiplication are \AF{𝟏} and \AF{\_⋆\_}, defined in \secref{Specifying Languages}.
-In this setting, language concatenation is subsumed by a very general form of \emph{convolution} \cite{Golan2005RecentSemi, Elliott2019-convolution}.
-The semimodule structure of predicates provides the additive aspect and is built entirely from the \emph{codomain}-transforming predicate operations defined in \secref{Specifying Languages}.
-The multiplicative aspect (convolution) comes entirely from the \emph{domain}-transforming operations, applied to any domain monoid.
+When our \AB{domain} type \AB{A} is a monoid (as with languages, infinite streams and grids, and functions of continuous space and time), predicates over \AB{A} form a second semiring, known as ``the monoid semiring'' \citep{Golan2005RecentSemi}, in which zero and addition are as in the first predicate commutative semiring and semimodule, while one and multiplication are \AF{𝟏} and \AF{\_⋆\_}, defined in \figreftwo{Lang-ops}{lang-via-pred}.
+In this setting, language concatenation is subsumed by a very general form of \emph{convolution} \cite{Golan2005RecentSemi, Dongol2016CUC, Elliott2019-convolution}.
+The semimodule structure of predicates provides the additive aspect and is built entirely from the \emph{codomain}-transforming predicate operations defined in \figref{Lang-ops} and redefined in \figref{lang-via-pred}.
+The multiplicative aspect (convolution) comes entirely from the \emph{domain}-transforming operations applied to any domain monoid.
 The two needed distributive properties hold for \emph{any} domain-transforming operation, whether associative or not.
@@
 As a bonus, when the domain monoid commutes (multiplicatively)---e.g., for functions of space and time rather than languages---the monoid semiring does as well.
+Applications include power series (univariate and multivariate) and image processing \citep{Elliott2019-convolution}.
+The role of semirings in parsing began with \citet{Chomsky1959CFL} and was further explored by \citet{Goodman1998PIO, Goodman1999SP} and by \citet{Liu2004}.
 
@@
 
-Given that type-level languages have useful and interesting algebraic properties and that the decidable implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation} are tightly connected to languages, do the algebraic properties transfer to those implementations?
+We have seen that type-level languages have useful and illuminating algebraic properties and that the decidable implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation} are tightly connected to languages.
+Must we prove these properties again for each implementation---here there are more operational details to manage---or do the algebraic properties somehow transfer to those implementations?
 This question immediately raises a technical obstacle.
@@
 For instance, each instance of the \ARe{Monoid} abstraction involves a single type \AB τ with a binary operation {\AB{\_∙\_} \AK : \AB τ \AK → \AB τ \AK → \AB τ} and identity value {\AB ε \AK : \AB τ}.
-This recipe accommodates the language operations in \figref{Lang-ops} and their generalizations in \figref{predicates} but not the dependently typed, \emph{indexed} versions of languages and their operations in \figreftwo{symbolic-api}{sized-automatic-api}, nor decidable types and their vocabulary in \figref{compositional-dec}.
+This recipe accommodates the language operations in \figreftwo{Lang-ops}{lang-via-pred} and their generalizations in \figref{predicates} but not the dependently typed, \emph{indexed} versions of languages and their operations in \figreftwo{symbolic-api}{sized-automatic-api}, nor decidable types and their vocabulary in \figref{compositional-dec}.
 
-Fortunately, there is a simple way to encapsulate indexed types into simple (non-indexed) types, namely existential quantification.
-For instance, {\AF{∃} \AD{Dec}} and {\AF{∃} \AD{(\ARe{Lang} \APo ∞)}} are simple types.
-For a type family {\AB F \AK : \AB A \AK → \APT Set \AB ℓ}, values of type {\AF ∃ \AB F} are pairs {(\AB a , \AB b)} with {\AB a \AK : \AB A} and {\AB b \AK : \AB F \AB a}.
+Fortunately, there is a simple way to encapsulate indexed types into non-indexed types, namely existential quantification.
+For instance, {\AF{∃} \AD{Dec}} and {\AF{∃} \AD{(\ARe{Lang} \APo ∞)}} are non-indexed types.
+For a type family {\AB F \AK : \AB A \AK → \APT{Set} \AB ℓ}, values of type {\AF ∃ \AB F} are pairs {(\AB a , \AB b)} with {\AB a \AK : \AB A} and {\AB b \AK : \AB F \AB a}.
 We can also encapsulate dependently typed \emph{operations} on indexed types into simply typed operations on the (simply typed) existentially wrapped versions of those types.
@@
 \end{code}
-Referring to the type and operations of \secref{Specifying Languages} and \figref{symbolic-api} via the module prefixes ``{◬.}\hspace{0.05em}'' and ``{▣.}\hspace{0.05em}'' respectively, define \agda{wrapped-Lang}
+Referring to the type and operations of \figreftwo{Lang-ops}{symbolic-api} via the module prefixes ``{◬.}\hspace{0.05em}'' and ``{▣.}\hspace{0.05em}'' respectively, define \agda{wrapped-Lang}
 Then wrap union as follows:
@@
 Indexed operations of all arities for a type family \AB{F} can be systematically wrapped in this style, as shown in
-\figrefdef{inj}{Existential wrappers}{\mathindent0ex\agda{inj}}, again with inferable components given explicitly for clarity.
-Then wrapped union can be defined simply as {\AF{\_∪\_} \AS = \AF{inj₂} \AIC{◬.\_∪\_}} and likewise for the other operations.
+\figrefdef{inj}{Existential wrappers}{\mathindent0ex\agda{inj}}, again with automatically inferable components given explicitly for clarity.
+Then wrapped union can be defined simply as {\AF{\_∪\_} \AS = \AF{inj₂} \AIC{▣.\_∪\_}} and likewise for the other operations.
 
-In addition to wrapping types and operations, we also need to transport \emph{properties}.
+In addition to transporting types and operations as just described, we also need to transport \emph{properties}.
 A crucial question is the choice of equivalence relation for the new properties.
-Choosing wrapped equivalence to be equivalence of \emph{first} projections (type \emph{indices}) results in all algebraic properties of those indices easily transporting to the existentially wrapped versions.
+If we choose equivalence on these dependent pairs to be equivalence of \emph{first} projections (type \emph{indices}), then all algebraic properties of those indices will hold for the existentially wrapped versions as well.
 This choice may at first seem like cheating, since the entire operational aspect of the representation is ignored.
 The correctness of that aspect, however, is guaranteed by its dependent typing, which anchors its meaning (no matter how cleverly implemented) to the non-operational type index.
-For language implementations of \secreftwo{Symbolic Differentiation}{Automatic Differentiation}, this anchoring is guaranteed by the type of \AF{⟦\_⟧} in \figreftwo{symbolic-api}{sized-automatic-api}.
-(Similarly, for the decidable type constructors in \figref{compositional-dec}, correctness is guaranteed by the types of \AIC{yes} and \AIC{no} in the definition of ‌\ARe{Dec} in \secref{Decidability}.)
-Given this choice of equivalence, properties of index types are easily lifted to properties of existential types, as in \figrefdef{prop}{Liftings of index properties to existential types}{\mathindent0ex\agda{prop}}.
+For the language implementations of \secreftwo{Symbolic Differentiation}{Automatic Differentiation}, this anchoring is guaranteed by the type of \AF{⟦\_⟧} in \figreftwo{symbolic-api}{sized-automatic-api}.
+(Similarly, for the decidable type constructors in \figref{compositional-dec}, correctness is guaranteed by the types of the \AIC{yes} and \AIC{no} constructors in the definition of ‌\ARe{Dec} in \secref{Decidability}.)
+Given this choice of equivalence, properties of index types easily lift to properties of existential types, as shown in \figrefdef{prop}{Liftings of index properties to existential types}{\mathindent0ex\agda{prop}}.
+(These definitions are partially inferable as well, e.g., from \agda{prop₃′}.)
 These definitions ease specification of algebraic instances for existentially wrapped types from corresponding instances for their indices.
@@
 Most work on formal languages in functional programming has been in parsing, particularly via parsing combinators \citep{Hutton1996monadic, Leijen2001parsec, Swierstra2008combinator}.
-Such work typically concentrates on usability, composability, and sometimes efficiency, rather than on simple, precise specification and verifiable correctness.
+Such work typically concentrates on usability, composability, and sometimes efficiency, rather than on simple semantic specification and verifiable correctness.
 
@@
 They also \emph{defined} indexed types for \AF{ν} and \AF{δ} on regular expressions rather than defining the meanings of those operations in terms of languages and then proving properties of them.
-They prove a consistency property about \AF{δ}, but apparently not about \AF{ν}, and they note their method's ``inability to prove the non-membership of some string in a given language''.
+They proved a consistency property about \AF{δ}, but apparently not about \AF{ν}, and they note their method's ``inability to prove the non-membership of some string in a given language''.
 
 \item \citet{DoczkalEtAl2013} formalized regular languages constructively in Coq, also in terms of a type-level membership predicate.
-They then prove several classic results about finite automata, including minimization and relationship to Nerode and Myhill partitions.
+They then proved several classic results about finite automata, including minimization and relationship to Nerode and Myhill partitions.
 
-\item \citet{FirsovUustalu2013} used Agda to formalize regular expressions with a corresponding inductive type of language membership proofs and then related regular expressions to nondeterministic finite automata (NFAs) represented as boolean matrices and operations on that representation that correspond to those on regular expressions.
+\item \citet{FirsovUustalu2013} used Agda to formalize regular expressions with a corresponding inductive type of language membership proofs and then related regular expressions to nondeterministic finite automata (NFAs) represented as boolean matrices, as well as operations on that representation that correspond to those on regular expressions.
 They then proved soundness and completeness of the NFA representation with respect to regular language membership.
@@
 \item \citet{KorkutEtAl2016} defined a non-indexed algebraic data type of regular expressions together with an indexed inductive type of proofs of membership of strings in the language denoted by the regular expressions and a corresponding function that matches an implicitly concatenated stack of regular expressions.
-Special attention is paid to ensuring and proving termination, particularly in regard to languages that contain the empty string.
-Addressing both involves defunctionalization and translating between general regular expressions and a restricted ``standard'' form.
+Special attention was paid to ensuring and proving termination, particularly in regard to languages that contain the empty string.
+Addressing both involved defunctionalization and translating between general regular expressions and a restricted ``standard'' form.
 
-\item \citet{Traytel2017} defined formal languages as tries and explored operations on tries defined via primitive corecursion and reasoned about coinductively in the Isabelle/HOL proof assistant.
-There appears to be no specification higher level (less operational) than tries, leading to some rather complex definitions (which, as \emph{definitions}, cannot be proved correct).
-The definition of language concatenation is particularly delicate.
-For the same reason, proofs of even basic properties involve coinduction.
-The authors point out that, however, that the coinductive trie formulation enabled better proof automation than with sets of strings.
+\item \citet{Traytel2017} defined formal languages as tries, with operations via primitive corecursion and reasoning via coinduction, all in the Isabelle/HOL proof assistant.
+There appears to have been no specification higher level (less operational) than tries, leading to some rather complex definitions (which, as \emph{definitions}, cannot be proved correct).
+Language concatenation was particularly delicate.
+For the same reason, proofs of even basic properties involved coinduction.
+The authors pointed out, however, that the coinductive trie formulation enabled better proof automation than with sets of strings.
 
@@
 Relationships to automata theory and coalgebraic notions were also explored in depth.
-Trie look-up yields boolean values rather than propositions, and there was no simpler specification of languages with respect to which the trie implementation could be specified and proved consistent, so every operation is \emph{defined} coinductively.
-The concatenation and closure (star) cases are especially delicate.
-Equality on tries is defined by strong bisimilarity, and thus properties require corresponding machinery to state and prove.
+Trie look-up yielded boolean values rather than propositions, and there was no simpler specification of languages with respect to which the trie implementation could be specified and proved consistent, so every operation was \emph{defined} coinductively.
+The concatenation and closure (star) cases were especially delicate.
+Equality on tries was defined by strong bisimilarity, and thus properties required corresponding machinery to state and prove.
 
@@
 
-Of the previous work involving language derivatives, none appear to use the essential, non-inductive, nearly-trivial definition in terms of functions of lists from \secref{Decomposing Languages}, although \citet{Abel2016} did indeed mention this definition informally as motivation for the more complex, coinductively defined operations on tries.
+\end{itemize}
+
+Of the previous work involving language derivatives, none appear to have based their formal specification and proofs on the essential, non-inductive, nearly-trivial definition in terms of functions of lists from \secref{Decomposing Languages}, although \citet{Abel2016} did indeed mention this definition informally as motivation for the more complex, coinductively defined operations on tries.
 \citet{Brzozowski64} also made clear mention of the underlying meaning on sets of strings in his original work on regular expression derivatives.
-
-\end{itemize}
 

### 2021-02-20 paper.tex
Make three sections into subsections of new section "From Languages to Parsing"

@@
 \item
-  A duality is revealed between regular expressions and tries (prefix trees) in the context of language differentiation (\secreftwo{Symbolic Differentiation}{Automatic Differentiation}).
+  A duality is revealed between regular expressions and tries (prefix trees) in the context of language differentiation (\secref{From Languages to Parsing}).
   Each implementation contains exactly the same code but in transposed forms.
@@
 \figrefdef{nu-delta-lemmas}{Properties of \AF{ν} and \AF{δ} for language operations}{\agda{νδ-lemmas}}.
-The correct-by-construction parsing algorithms in \secreftwo{Symbolic Differentiation}{Automatic Differentiation} are corollaries of these properties.
+The correct-by-construction parsing algorithms in \secref{From Languages to Parsing} are corollaries of these properties.
 
@@
 
-\sectionl{Reflections}
+\sectionl{From Languages to Parsing}
+
+\subsectionl{Reflections}
 
@@
 
-\secreftwo{Symbolic Differentiation}{Automatic Differentiation} apply the symbolic and automatic strategies respectively to languages.
-In both cases, the algorithms are proved correct with minimal effort as corollaries of the lemmas in \figref{nu-delta-lemmas}.
+\secref{From Languages to Parsing} applies the symbolic and automatic strategies respectively to languages.
+In both cases, the algorithms are automatically proved correct by type checking as corollaries of the lemmas in \figref{nu-delta-lemmas}.
 
-\sectionl{Symbolic Differentiation}
+\subsectionl{Symbolic Differentiation}
 \rnc\source{Symbolic}
@@
 
-\sectionl{Automatic Differentiation}
+\subsectionl{Automatic Differentiation}
 \rnc\source{Automatic}
@@
 The \AF{ν} and \AF{δ} lemmas for codomain (but not domain) operations in \figref{nu-delta-lemmas} are easily generalized as shown in \figrefdef{nu-delta-codomain-lemmas}{Properties of \AF{ν} and \AF{δ} for predicate codomain operations}{\agda{νδ-codomain}} and are all proved automatically by normalization.
-Since the decidable language implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation} are corollaries of \AF{ν} and \AF{δ} lemmas, those implementations adapt easily to the generalized codomain operations (\AF{pureᵀ}, \AF{mapᵀ}, and \AF{mapᵀ₂}), resulting in a somewhat smaller implementation and broader coverage.
+Since the decidable language implementations in \secref{From Languages to Parsing} are corollaries of \AF{ν} and \AF{δ} lemmas, those implementations adapt easily to the generalized codomain operations (\AF{pureᵀ}, \AF{mapᵀ}, and \AF{mapᵀ₂}), resulting in a somewhat smaller implementation and broader coverage.
 
@@
 
-We have seen that type-level languages have useful and illuminating algebraic properties and that the decidable implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation} are tightly connected to languages.
+We have seen that type-level languages have useful and illuminating algebraic properties and that the decidable implementations in \secref{From Languages to Parsing} are tightly connected to languages.
 Must we prove these properties again for each implementation---here there are more operational details to manage---or do the algebraic properties somehow transfer to those implementations?
@@
 The correctness of that aspect, however, is guaranteed by its dependent typing, which anchors its meaning (no matter how cleverly implemented) to the non-operational type index.
-For the language implementations of \secreftwo{Symbolic Differentiation}{Automatic Differentiation}, this anchoring is guaranteed by the type of \AF{⟦\_⟧} in \figreftwo{symbolic-api}{sized-automatic-api}.
+For the language implementations of \secref{From Languages to Parsing}, this anchoring is guaranteed by the type of \AF{⟦\_⟧} in \figreftwo{symbolic-api}{sized-automatic-api}.
 (Similarly, for the decidable type constructors in \figref{compositional-dec}, correctness is guaranteed by the types of the \AIC{yes} and \AIC{no} constructors in the definition of ‌\ARe{Dec} in \secref{Decidability}.)
@@
 \rnc\source{Transport}
-As examples, we can lift the commutative semiring of types to wrappings of the decidable counterparts from \figref{compositional-dec} in \secref{Decidability}, and lift both language semirings to the decidable implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation}, as in \figrefdef{transport-examples}{Examples of algebraic instances transported from indices to existential wrappings}{\agda{examples}}.
+As examples, we can lift the commutative semiring of types to wrappings of the decidable counterparts from \figref{compositional-dec} in \secref{Decidability}, and lift both language semirings to the decidable implementations in \secref{From Languages to Parsing}, as in \figrefdef{transport-examples}{Examples of algebraic instances transported from indices to existential wrappings}{\agda{examples}}.
 
@@
 
-%% requiring more careful construction and reading than the specification given in \figref{Lang-ops} and used in the implementations in \secreftwo{Symbolic Differentiation}{Automatic Differentiation}
+%% requiring more careful construction and reading than the specification given in \figref{Lang-ops} and used in the implementations in \secref{From Languages to Parsing}
 

### 2021-02-20 paper.tex
spacing
"early draft" ⟶ "draft"
abstract wording

@@
 %% \acmtrue
-\acmfalse
 
@@
 \title{\tit}
-\date{Early draft of \DTMnow{} GMT-8}
+\date{Draft of \DTMnow{} GMT-8}
 
@@
 A simple characterization of languages (and indeed functions from lists to any type) captures the essential idea behind language ``differentiation'' (and integration) as used for recognizing languages, leading to a collection of lemmas about type-level predicates.
-These lemmas are the heart of two dual implementations of language recognition---(inductive) regular expressions and (coinductive) tries---each containing the same code but in dual arrangements.
-The language lemmas form most of the implementation in both cases, while the representation and primitive operations trade places.
+These lemmas are the heart of two dual parsing implementations---using (inductive) regular expressions and (coinductive) tries---each containing the same code but in dual arrangements (with representation and primitive operations trading places).
 The regular expression version corresponds to symbolic differentiation, while the trie version corresponds to automatic differentiation.
@@
 \end{center}
+\ifacm
+\vspace{1.5ex}
+\else
 \vspace{-2ex}
+\fi
 \hfill
@@
 \end{center}
+\ifacm
+\vspace{1ex}
+\else
 \vspace{-2ex}
+\fi
 \hfill

### 2021-02-21 paper.tex
Typo correction. Fixes #2.

@@
 \input{macros}
+
+\definecolor{mylinkcolor}{rgb}{0,0.2,0}
+\hypersetup{
+  linkcolor  = mylinkcolor,
+  citecolor  = mylinkcolor,
+  urlcolor   = mylinkcolor,
+  colorlinks = true, % false
+}
 
@@
 For instance, a proof of {\AB{w} ∈ \AB{P} \AF ∪ \AB{Q}} contains a proof of {\AB{w} ∈ \AB{P}} or of {\AB{w} \AF ∈ \AB{Q}}, \emph{and} the knowledge of which one was chosen.
-Likewise, a proof of {w ∈ \AB{P} \AF ⋆ \AB{Q}} (language concatenation) includes the choice of strings \AB{u} and \AB{v}, a proof that {\AB{u} \AF ⊙ \AB{v} \AD ≡ \AB{w}}, and proofs that {\AB{u} ∈ \AB{P}} and {\AB{v} ∈ \AB{P}}.
+Likewise, a proof of {w ∈ \AB{P} \AF ⋆ \AB{Q}} (language concatenation) includes the choice of strings \AB{u} and \AB{v}, a proof that {\AB{u} \AF ⊙ \AB{v} \AD ≡ \AB{w}}, and proofs that {\AB{u} ∈ \AB{P}} and {\AB{v} ∈ \AB{Q}}.
 

### 2021-02-21 paper.tex
"bisumulation" ⟶ "bisimulation". Fixes #3.

@@
 The relatively easy-to-prove properties of type-level languages transfer almost effortlessly to the decidable implementations.
-In particular, despite the inductive and coinductive nature of regular expressions and tries respectively, we need neither inductive nor coinductive/bisumulation arguments to prove algebraic properties.
+In particular, despite the inductive and coinductive nature of regular expressions and tries respectively, we need neither inductive nor coinductive/bisimulation arguments to prove algebraic properties.
 \end{abstract}

### 2021-02-22 paper.tex
Use ◇ for diamond

@@
 \ifacm
-\newunicodechar{◬}{\raisebox{0.02ex}{\ensuremath{_{\diamond}\hspace{-0.323em}}}}
+\newunicodechar{◇}{\raisebox{0.02ex}{\ensuremath{_{\diamond}\hspace{-0.323em}}}}
 \newunicodechar{▣}{\raisebox{-0.12ex}{\ensuremath{\scaleobj{0.4}{\Box}}\hspace{-0.238em}}}
@@
 \setmathfont{XITSMath-Regular.otf}
-\newunicodechar{◬}{\raisebox{0.23ex}{\ensuremath{_{\diamond}\hspace{-0.358em}}}}
+\newunicodechar{◇}{\raisebox{0.23ex}{\ensuremath{_{\diamond}\hspace{-0.358em}}}}
 \newunicodechar{▣}{\raisebox{-0.13ex}{\ensuremath{\scaleobj{0.3}{\Box}}\hspace{-0.23em}}}
@@
 The ``symbolic differentiation'' style reflects language-building vocabulary (\figref{Lang-ops}) into an inductive data type (of modestly extended regular expressions), while \AF{ν} and \AF{δ} become functions defined by pattern-matching on that data type.
-For convenience, we can use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the module prefix ``{◬.}\hspace{0.05em}''.
+For convenience, we can use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the module prefix ``{◇.}\hspace{0.05em}''.
 The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\api} along with a function {\AF{⟦\_⟧}} that converts a symbolic language representation to a computable parser.
@@
 This time, reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type of tries\footnote{The classic trie (``prefix tree'') data structure was introduced by \citet{Thue1912Gegenseitige} to represent sets of strings \citep[Section 6.3]{Knuth1998ACP3}, later generalized to arbitrary non-nested algebraic data types \citep{Connelly1995GenTrie}, and then from sets to functions \citep{Hinze2000GGT}.}, while redefining the language-building vocabulary as functions on that data type defined by \emph{copatterns} \citep{AbelPientka2016}.
-Again, we will use the same names as in \secref{Specifying Languages}, referring to the original versions via the module prefix ``{◬.}\hspace{0.05em}''.
+Again, we will use the same names as in \secref{Specifying Languages}, referring to the original versions via the module prefix ``{◇.}\hspace{0.05em}''.
 The result is shown in \figrefdef{automatic-api}{Tries (coinductive)}{\api}.
@@
 \AgdaSymbol{(}\AgdaGeneralizable{P}%
-\>[39]\AgdaOperator{\AgdaFunction{◬.∪}}%
+\>[39]\AgdaOperator{\AgdaFunction{◇.∪}}%
 \>[44]\AgdaGeneralizable{Q}\AgdaSymbol{)}\<%
 \end{code}
-Referring to the type and operations of \figreftwo{Lang-ops}{symbolic-api} via the module prefixes ``{◬.}\hspace{0.05em}'' and ``{▣.}\hspace{0.05em}'' respectively, define \agda{wrapped-Lang}
+Referring to the type and operations of \figreftwo{Lang-ops}{symbolic-api} via the module prefixes ``{◇.}\hspace{0.05em}'' and ``{▣.}\hspace{0.05em}'' respectively, define \agda{wrapped-Lang}
 Then wrap union as follows:

### 2021-02-22 paper.tex
OstermannJabs2018

@@
 This time, reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type of tries\footnote{The classic trie (``prefix tree'') data structure was introduced by \citet{Thue1912Gegenseitige} to represent sets of strings \citep[Section 6.3]{Knuth1998ACP3}, later generalized to arbitrary non-nested algebraic data types \citep{Connelly1995GenTrie}, and then from sets to functions \citep{Hinze2000GGT}.}, while redefining the language-building vocabulary as functions on that data type defined by \emph{copatterns} \citep{AbelPientka2016}.
+(This program duality pattern has been noted and investigated by \citet{OstermannJabs2018}.)
 Again, we will use the same names as in \secref{Specifying Languages}, referring to the original versions via the module prefix ``{◇.}\hspace{0.05em}''.

### 2021-02-22 paper.tex
box module

@@
 \newunicodechar{◇}{\raisebox{0.02ex}{\ensuremath{_{\diamond}\hspace{-0.323em}}}}
-\newunicodechar{▣}{\raisebox{-0.12ex}{\ensuremath{\scaleobj{0.4}{\Box}}\hspace{-0.238em}}}
+\newunicodechar{□}{\raisebox{-0.12ex}{\ensuremath{\scaleobj{0.4}{\Box}}\hspace{-0.238em}}}
 \else
@@
 \newunicodechar{◇}{\raisebox{0.23ex}{\ensuremath{_{\diamond}\hspace{-0.358em}}}}
-\newunicodechar{▣}{\raisebox{-0.13ex}{\ensuremath{\scaleobj{0.3}{\Box}}\hspace{-0.23em}}}
+\newunicodechar{□}{\raisebox{-0.13ex}{\ensuremath{\scaleobj{0.3}{\Box}}\hspace{-0.23em}}}
 \newunicodechar{∷}{\ensuremath{\mathbin{\hspace{-0.2em}:\hspace{-0.27em}:\hspace{-0.15em}}}}
@@
 \end{code}
-Referring to the type and operations of \figreftwo{Lang-ops}{symbolic-api} via the module prefixes ``{◇.}\hspace{0.05em}'' and ``{▣.}\hspace{0.05em}'' respectively, define \agda{wrapped-Lang}
+Referring to the type and operations of \figreftwo{Lang-ops}{symbolic-api} via the module prefixes ``{◇.}\hspace{0.05em}'' and ``{□.}\hspace{0.05em}'' respectively, define \agda{wrapped-Lang}
 Then wrap union as follows:
@@
 \figrefdef{inj}{Existential wrappers}{\mathindent0ex\agda{inj}}, again with automatically inferable components given explicitly for clarity.
-Then wrapped union can be defined simply as {\AF{\_∪\_} \AS = \AF{inj₂} \AIC{▣.\_∪\_}} and likewise for the other operations.
+Then wrapped union can be defined simply as {\AF{\_∪\_} \AS = \AF{inj₂} \AIC{□.\_∪\_}} and likewise for the other operations.
 

### 2021-02-26 paper.tex
Typo fix. Closes #4.

@@
 
-The definitions in \figref{Lang-ops} reflect the logical interpretation of types under the Curry-Howard (``propositions as types'') isomorphism, in which types represent propositions and element (values) of a type represent proofs of the corresponding proposition \citep{Wadler2015propositions}.
+The definitions in \figref{Lang-ops} reflect the logical interpretation of types under the Curry-Howard (``propositions as types'') isomorphism, in which types represent propositions and elements (values) of a type represent proofs of the corresponding proposition \citep{Wadler2015propositions}.
 The specific embodiment of Curry-Howard in \figref{Lang-ops} and throughout this paper is the language Agda \citep{Norell2008AFP, BoveEtAl2009Agda}, which is both programming language and proof assistant founded on Martin-Löf's intuitionistic type theory \citep{MartinLöf1984intuitionistic}.

### 2021-02-26 paper.tex
Consistent tagging of `_∈_` as a regular function. Closes #6.

@@
 
-Note that for a language \AB{P} and string \AB{w}, {\AB{P} \AB{w}} is the type of \emph{proofs} that {\AB{w} ∈ \AB{P}}, in other words \emph{explanations} of membership, or \emph{parsings}.
+Note that for a language \AB{P} and string \AB{w}, {\AB{P} \AB{w}} is the type of \emph{proofs} that {\AB{w} \AF ∈ \AB{P}}, in other words \emph{explanations} of membership, or \emph{parsings}.
 (If the type {\AB{P} \AB{w}} is uninhabited, then {\AB{w} ∉ \AB{P}}.)
-For instance, a proof of {\AB{w} ∈ \AB{P} \AF ∪ \AB{Q}} contains a proof of {\AB{w} ∈ \AB{P}} or of {\AB{w} \AF ∈ \AB{Q}}, \emph{and} the knowledge of which one was chosen.
-Likewise, a proof of {w ∈ \AB{P} \AF ⋆ \AB{Q}} (language concatenation) includes the choice of strings \AB{u} and \AB{v}, a proof that {\AB{u} \AF ⊙ \AB{v} \AD ≡ \AB{w}}, and proofs that {\AB{u} ∈ \AB{P}} and {\AB{v} ∈ \AB{Q}}.
+For instance, a proof of {\AB{w} \AF ∈ \AB{P} \AF ∪ \AB{Q}} contains a proof of {\AB{w} \AF ∈ \AB{P}} or of {\AB{w} \AF ∈ \AB{Q}}, \emph{and} the knowledge of which one was chosen.
+Likewise, a proof of {w \AF ∈ \AB{P} \AF ⋆ \AB{Q}} (language concatenation) includes the choice of strings \AB{u} and \AB{v}, a proof that {\AB{u} \AF ⊙ \AB{v} \AD ≡ \AB{w}}, and proofs that {\AB{u} \AF ∈ \AB{P}} and {\AB{v} \AF ∈ \AB{Q}}.
 

### 2021-02-26 paper.tex
Clarify wording. Fixes #7 (I hope).

@@
 (If the type {\AB{P} \AB{w}} is uninhabited, then {\AB{w} ∉ \AB{P}}.)
-For instance, a proof of {\AB{w} \AF ∈ \AB{P} \AF ∪ \AB{Q}} contains a proof of {\AB{w} \AF ∈ \AB{P}} or of {\AB{w} \AF ∈ \AB{Q}}, \emph{and} the knowledge of which one was chosen.
+For instance, every proof of {\AB{w} \AF ∈ \AB{P} \AF ∪ \AB{Q}} contains a proof of {\AB{w} \AF ∈ \AB{P}} or of {\AB{w} \AF ∈ \AB{Q}}, \emph{and} the knowledge of which one was chosen (with different proofs reflecting different choices).
 Likewise, a proof of {w \AF ∈ \AB{P} \AF ⋆ \AB{Q}} (language concatenation) includes the choice of strings \AB{u} and \AB{v}, a proof that {\AB{u} \AF ⊙ \AB{v} \AD ≡ \AB{w}}, and proofs that {\AB{u} \AF ∈ \AB{P}} and {\AB{v} \AF ∈ \AB{Q}}.

### 2021-02-26 paper.tex
BaanenSwierstra2020

@@
 
+\item \citet{BaanenSwierstra2020} also explored correctness of regular expression matching in Agda, but from the perspective of \emph{effects} and their predicate transformer semantics.
+The authors shared a common starting point with \citet{KorkutEtAl2016} and addressed differentiation on regular expressions, while separating termination from partial correctness.
+Their regular expression matching function targets a free monad (later to be interpreted with suitable effect semantics) rather than propositions.
+
 \end{itemize}

### 2021-02-27 paper.tex
non-idempotence

@@
 
-%% \acmtrue
+\acmtrue
 
 \ifacm
-\documentclass[acmsmall,screen,anonymous,timestamp]{acmart}
+\documentclass[acmsmall,screen,anonymous,timestamp,review]{acmart}
 \else
@@
 \bibliographystyle{ACM-Reference-Format}
+\citestyle{acmauthoryear}
 \else
@@
 \hfill\;
-} shows two parallel collections of predicate operations, each covariantly transforming the codomain (left) or domain (right).\footnote{The types of operations on the left can be further relaxed from \AF{Set⇂} to any codomain type, while keeping \AB{A} fixed, resulting in the usual covariant applicative functor of functions \emph{from} a fixed type \citep{McBride2008APE}.
-Dually, the operations on the right (in their current generality) form a covariant applicative functor of functions \emph{to} the type \AF{Set⇂}.}
+} shows two parallel collections of predicate operations, each covariantly transforming the codomain (left) or domain (right).\footnote{The types of operations on the left can be further relaxed from \AF{Set ℓ} to any codomain type, while keeping \AB{A} fixed, resulting in the usual covariant applicative functor of functions \emph{from} a fixed type \citep{McBride2008APE}.
+Dually, the operations on the right (in their current generality) form a covariant applicative functor of functions \emph{to} types.}
 These generalized operations then specialize to language operations as in \figrefdef{lang-via-pred}{Languages via predicate operations}{
@@
 
-Conveniently, booleans and types form commutative semirings with all necessary proofs already in the standard library.
-(The equivalence relation used for types is isomorphism rather than equality.)
-Both are also closed.
+Conveniently, booleans and types form commutative semirings with all necessary proofs already in the standard library.\footnote{The equivalence relation used for types is isomorphism rather than equality.
+The boolean semiring is idempotent, while the type semiring is not (in order to distinguish multiple parsings in ambiguous languages).
+The latter non-idempotence accounts for the difference between the \AF ν and \AF δ lemmas for {\AB P ⋆ \AB Q} and {\AB P ☆} in \figref{nu-delta-lemmas} and the corresponding rules in previous work, as mentioned in \secref{Related Work}.}
+They are both closed as well.
 For booleans, closure maps both \AIC{false} and \AIC{true} to \AIC{true}, with the \AF{starˡ} law holding definitionally.
@@
 \citet{Brzozowski64} also made clear mention of the underlying meaning on sets of strings in his original work on regular expression derivatives.
+Moreover, none of these previous investigations appear to have addressed proof isomorphism (distinguishing multiple parsings/explanations), and correspondingly all use transformations based the less precise laws originally discovered by \citet{Brzozowski64}.
+Those laws were based on the assumption that union (``addition'' in both language semirings of \secref{Predicate Algebra}) is idempotent, which is not true for type-level language predicates related by proof isomorphism (rather than mere logical equivalence).
+The lack of idempotence is crucial for going beyond recognizers to parsers, in which we might want to extract more than one proof (parsing).
+The difference is mainly visible in the \AF ν and \AF δ lemmas for {\AB P ⋆ \AB Q} and {\AB P ☆} in \figref{nu-delta-lemmas}.
 

### 2021-02-27 paper.tex
wordsmithing

@@
 \item Types have type {\APT{Set} \AB ℓ} for some universe level \AB ℓ.
-These levels avoid logical inconsistencies resulting from {\APT{Set} \AS : \APT{Set}} but can be safely ignored in this paper (which is parametrized over \AB ℓ).
+These levels avoid logical inconsistencies\out{ resulting from {\APT{Set} \AS : \APT{Set}}} but can be safely ignored in this paper (which is parametrized over \AB ℓ).
 
@@
 (If the type {\AB{P} \AB{w}} is uninhabited, then {\AB{w} ∉ \AB{P}}.)
-For instance, every proof of {\AB{w} \AF ∈ \AB{P} \AF ∪ \AB{Q}} contains a proof of {\AB{w} \AF ∈ \AB{P}} or of {\AB{w} \AF ∈ \AB{Q}}, \emph{and} the knowledge of which one was chosen (with different proofs reflecting different choices).
-Likewise, a proof of {w \AF ∈ \AB{P} \AF ⋆ \AB{Q}} (language concatenation) includes the choice of strings \AB{u} and \AB{v}, a proof that {\AB{u} \AF ⊙ \AB{v} \AD ≡ \AB{w}}, and proofs that {\AB{u} \AF ∈ \AB{P}} and {\AB{v} \AF ∈ \AB{Q}}.
+For instance, every proof of {\AB{w} \AF ∈ \AB{P} \AF ∪ \AB{Q}} contains a proof of {\AB{w} \AF ∈ \AB{P}} or of {\AB{w} \AF ∈ \AB{Q}} \emph{and} the knowledge of which one was chosen (with different proofs reflecting different choices).
+Likewise, a proof of {w \AF ∈ \AB{P} \AF ⋆ \AB{Q}}\out{ (language concatenation)} includes the choice of strings \AB{u} and \AB{v}, a proof that {\AB{u} \AF ⊙ \AB{v} \AD ≡ \AB{w}}, and proofs that {\AB{u} \AF ∈ \AB{P}} and {\AB{v} \AF ∈ \AB{Q}}.
 
@@
 Not only do types form a closed semiring, the \emph{only} properties needed from types in the monoid semiring come from the laws of closed semirings.
-One can thus apply the ideas in this paper to many other useful semirings as well, including natural numbers (e.g., number of parses), probability distributions, and the tropical semirings (max/plus and min/plus) for optimization.
+One can thus apply the ideas in this paper to many other useful semirings, including natural numbers (e.g., number of parses), probability distributions, and the tropical semirings (max/plus and min/plus) for optimization.
 \out{We will explore this generality in \secref{Beyond Predicates}.}
-As a bonus, when the domain monoid commutes (multiplicatively)---e.g., for functions of space and time rather than languages---the monoid semiring does as well.
+When the domain monoid commutes (multiplicatively)---e.g., for functions of space and time rather than languages---the monoid semiring does as well.
 Applications include power series (univariate and multivariate) and image processing \citep{Elliott2019-convolution}.
-The role of semirings in parsing began with \citet{Chomsky1959CFL} and was further explored by \citet{Goodman1998PIO, Goodman1999SP} and by \citet{Liu2004}.
+Research on semirings in parsing began with \citet{Chomsky1959CFL} and was further explored by \citet{Goodman1998PIO, Goodman1999SP} and by \citet{Liu2004}.
 

### 2021-02-27 paper.tex
spacing ; AD

@@
 
-\secref{From Languages to Parsing} applies the symbolic and automatic strategies respectively to languages.
+\secreftwo{Symbolic Differentiation}{Automatic Differentiation} applies the symbolic and automatic strategies respectively to languages.
 In both cases, the algorithms are automatically proved correct by type checking as corollaries of the lemmas in \figref{nu-delta-lemmas}.
@@
 This time the compiler successfully proves \emph{total} correctness (including termination).
+
+%% To more clearly see the parallel with automatic differentiation, note that
 
@@
 \rnc\source{Calculus}
-The \AF{ν} and \AF{δ} lemmas for codomain (but not domain) operations in \figref{nu-delta-lemmas} are easily generalized as shown in \figrefdef{nu-delta-codomain-lemmas}{Properties of \AF{ν} and \AF{δ} for predicate codomain operations}{\agda{νδ-codomain}} and are all proved automatically by normalization.
+The \AF{ν} and \AF{δ} lemmas for codomain (but not domain) operations in \figref{nu-delta-lemmas} are easily generalized as shown in \figrefdef{nu-delta-codomain-lemmas}{Properties of \AF{ν} and \AF{δ} for predicate codomain operations}{\setstretch{1.5}\agda{νδ-codomain}} and are all proved automatically by normalization.
 Since the decidable language implementations in \secref{From Languages to Parsing} are corollaries of \AF{ν} and \AF{δ} lemmas, those implementations adapt easily to the generalized codomain operations (\AF{pureᵀ}, \AF{mapᵀ}, and \AF{mapᵀ₂}), resulting in a somewhat smaller implementation and broader coverage.

### 2021-02-27 paper.tex
𝒟 ; Clarify tries as "automatic differentiation".

@@
 
-In order to convert languages into parsers, it will help to decompose languages into a regular form.
-Since lists are defined by an algebraic data type, we can decompose languages---and indeed any function from lists---according to constructor, as follows:
-\agda{νδ}
-In terms of languages, {\AF{δ} \AB{P} \AB{a}} comprises all strings \AB{w} such that {\AB{a} \AIC ∷ \AB{w}} is in \AB{P}\out{, i.e., the tails of strings in \AB{P} that start with \AB{a}}.
+In order to convert languages into parsers, it will help to understand how language membership relates to the algebraic structure of lists, specifically the monoid formed by \AIC{[]} and \AF{\_⊙\_}.
+Generalizing from languages to functions from lists to \emph{any} type, define
+\agda{ν𝒟}
+In terms of languages, {\AF{𝒟} \AB{P} \AB{u}} comprises all strings \AB{v} such that {\AB{u} \AF ⊙ \AB{v} \AF ∈ \AB{P}}, i.e., \AB{u}-suffixes of strings in \AB{P}.
 
-Each use of \AF{δ} takes us one step closer to reducing general language membership to nullability.
-In other words,\footnote{For functions \AB{f} and \AB{g}, {\AB{f} \AF{≗} \AB{g}} is extensional equality, i.e., {\AS ∀ \AB x \AS → \AB{f} \AB{x} \AD{≡} \AB{g} \AB{x}}.
-Repeated application is expressed as a standard left fold\out{ \stdlibCitep{Data.List}}:
+Since {\AB{u} \AF ⊙ \AIC{[]} \AD ≡ \AB{u}}, it follows that\footnote{For functions \AB{f} and \AB{g}, {\AB{f} \AF{≗} \AB{g}} is extensional equality, i.e., {\AS ∀ \AB x \AS → \AB{f} \AB{x} \AD{≡} \AB{g} \AB{x}}.}
+\agda{ν∘𝒟}
+Moreover, the function \AF{𝒟} distributes over \AIC{[]} and \AF{\_⊙\_}:\out{\footnote{In algebraic terms, argument-flipped \AF{𝒟} is a monoid homomorphism from lists to language endofunctions.}}
+\agda{𝒟[]⊙}
+These facts suggest that \AB{𝒟} can be computed one list element at a time via a more specialized operation, as is indeed the case:\footnote{Repeated \AF{δ} application is expressed as a standard left fold\out{ \stdlibCitep{Data.List}}:
 \agda{foldl}\vspace{-2ex}
-}%
-\agda{ν∘foldlδ}
+}
+\agda{δ}
+\agda{𝒟foldl}
+Each use of \AF{δ} thus takes us one step closer to reducing general language membership to nullability.
+%% In other words,%
+%% \agda{ν∘foldlδ}
 This simple observation is the semantic heart of the syntactic technique of \emph{derivatives of regular expressions} as used for efficient recognition of regular languages \citep{Brzozowski64}, later extended to parsing general context-free languages \citep{Might2010YaccID}.
-Lemma \AF{ν∘foldlδ} liberates this technique from the assumption that languages are represented \emph{symbolically}, by some form of grammar (e.g., regular or context-free), inviting other representations as we will see in \secref{Automatic Differentiation}.
+Lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} liberate this technique from the assumption that languages are represented \emph{symbolically}, by some form of grammar (e.g., regular or context-free), inviting other representations as we will see in \secref{Automatic Differentiation}.
 
+\begin{comment}
 Lemma \AF{ν∘foldlδ} relates to some other powerful perspectives and principles as well:
@@
 \end{itemize}
+\end{comment}
 
@@
 
-%% To more clearly see the parallel with automatic differentiation, note that
+\rnc\source{Calculus}
+
+To more clearly see the parallel with automatic differentiation (AD), note that AD implementations generally sample a function \emph{and} its derivative together to exploit the fact that these two computations typically share much common work.
+This work sharing also applies when differentiating languages.
+Without this optimization we have
+\agda{𝒟′}
+The more efficient, work-sharing version:
+\agda{ʻ𝒟}
+Their equivalence follows from lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} of \secref{Decomposing Languages}:
+\agda{𝒟′≡ʻ𝒟}
+The trie representation accomplishes this sharing by weaving \AF{ν} and \AF{δ} into a single structure based on common prefixes.
 
@@
 \hfill\;
-} shows two parallel collections of predicate operations, each covariantly transforming the codomain (left) or domain (right).\footnote{The types of operations on the left can be further relaxed from \AF{Set ℓ} to any codomain type, while keeping \AB{A} fixed, resulting in the usual covariant applicative functor of functions \emph{from} a fixed type \citep{McBride2008APE}.
+} shows two parallel collections of predicate operations, each covariantly transforming the codomain (left) or domain (right).\footnote{The types of operations on the left can be further relaxed from {\APT{Set} \AB ℓ} to any codomain type, while keeping \AB{A} fixed, resulting in the usual covariant applicative functor of functions \emph{from} a fixed type \citep{McBride2008APE}.
 Dually, the operations on the right (in their current generality) form a covariant applicative functor of functions \emph{to} types.}

### 2021-02-27 paper.tex
automata perspective

@@
 
-\begin{comment}
-Lemma \AF{ν∘foldlδ} relates to some other powerful perspectives and principles as well:
-\begin{itemize}
-
-\item 
-\emph{Automata theory}: {\AF{ν} \AF ∘ \AF{foldl} \AF{δ} \AB{P}} is the execution of a state machine.
-Each state is a language, with \AB{P} being the initial state; \AF{δ} is the state transition function; and \AB{ν} is the set of accepting states\out{ \needcite{}}.
-Lemma \AF{ν∘foldlδ} captures the correctness of this state machine as a recognizer for the language \AB{P}.
-
-\item
-\emph{Incremental computation}: {\AF{ν} \AF ∘ \AF{foldl} \AF{δ} \AB{P}} computes \AB{P} by transforming successive incremental modifications to the empty string (each in the form of {\AB{a} \AIC ∷\_} for a list element \AB{a}) into incremental modifications to the membership question for the related language {\AF{δ} \AB{P} \AB{a}}\out{ \needcite{}}.
-Lemma \AF{ν∘foldlδ} again captures the correctness of this incremental implementation of \AB{P}.
-
-\item 
-\emph{Calculus}: \AB{δ} is differentiation; {\AF{ν} \AF ∘ \AF{foldl} \AF{δ} \AB{P}} is integration; and lemma \AF{ν∘foldlδ} is the second fundamental theorem of calculus, which says that a function can be recovered by integrating its derivative\out{ \needcite{}}.
-
-\end{itemize}
-\end{comment}
+In terms of automata theory, derived languages are states, with \AB{P} being the initial state and {\AF{𝒟} \AB{P} \AB{u}} the state reached from \AB{P} after consuming the string \AB{u}; \AF{δ} is the state transition function; and \AB{ν} is the set of accepting states\out{ \needcite{}}.
+Lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} capture the correctness of this state machine as a recognizer for the language \AB{P}.
 

### 2021-02-27 paper.tex
spacing and wording

@@
 
-\acmtrue
+%% \acmtrue
 
@@
 \documentclass[hidelinks]{article}  % fleqn,12pt
-\usepackage[margin=1in]{geometry}  % 0.12in, 0.9in, 1in
+\usepackage[margin=0.9in]{geometry}  % 0.12in, 0.9in, 1in
 \usepackage[useregional]{datetime2}
@@
 \title{\tit}
-\date{Draft of \DTMnow{} GMT-8}
+\date{Version of \DTMnow{} GMT-8}
 
@@
 \agda{ν𝒟}
-In terms of languages, {\AF{𝒟} \AB{P} \AB{u}} comprises all strings \AB{v} such that {\AB{u} \AF ⊙ \AB{v} \AF ∈ \AB{P}}, i.e., \AB{u}-suffixes of strings in \AB{P}.
+In terms of languages, {\AF{𝒟} \AB{P} \AB{u}} comprises all strings \AB{v} such that {\AB{u} \AF ⊙ \AB{v} \AF ∈ \AB{P}}, i.e., the \AB{u}-suffixes of strings in \AB{P}.
 
@@
 \agda{δ}
+\vspace{-4ex}
 \agda{𝒟foldl}

### 2021-02-28 paper.tex
misc tidying

@@
 
-%% \acmtrue
+\acmtrue
 
@@
 \agda{ν∘𝒟}
-Moreover, the function \AF{𝒟} distributes over \AIC{[]} and \AF{\_⊙\_}:\out{\footnote{In algebraic terms, argument-flipped \AF{𝒟} is a monoid homomorphism from lists to language endofunctions.}}
+Moreover, the function \AF{𝒟} distributes over \AIC{[]} and \AF{\_⊙\_}:\footnote{In algebraic terms, argument-flipped \AF{𝒟} is a monoid homomorphism from lists to endofunctions.}
 \agda{𝒟[]⊙}
-These facts suggest that \AB{𝒟} can be computed one list element at a time via a more specialized operation, as is indeed the case:\footnote{Repeated \AF{δ} application is expressed as a standard left fold\out{ \stdlibCitep{Data.List}}:
+These facts suggest that \AB{𝒟} can be computed one list element at a time via a more specialized operation, as is indeed the case:\footnote{Repeated \AF{δ} application is expressed here via a standard left fold\out{ \stdlibCitep{Data.List}}:
 \agda{foldl}\vspace{-2ex}
+In \AF{𝒟foldl}, {\AB{X} \AS = \AB{A} \AF ✶ \AS → \AB B}.
 }
 \agda{δ}
-\vspace{-4ex}
+%\vspace{-4ex}
 \agda{𝒟foldl}
@@
 
-Decidable parsing (via {\AF{⟦\_⟧}}) relies only on \AF{ν} and \AF{δ}, which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\out{\notefoot{Describe this derivation.}}
+Decidable parsing relies only on \AF{ν} and \AF{δ} (via {\AF{⟦\_⟧}}), which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\out{\notefoot{Describe this derivation.}}
 These definitions are meant to be read in ``column-major'' order, i.e., each column is one function definition.

### 2021-02-28 paper.tex
fix wonky footnote; trim some words

@@
 \agda{𝒟[]⊙}
-These facts suggest that \AB{𝒟} can be computed one list element at a time via a more specialized operation, as is indeed the case:\footnote{Repeated \AF{δ} application is expressed here via a standard left fold\out{ \stdlibCitep{Data.List}}:
-\agda{foldl}\vspace{-2ex}
-In \AF{𝒟foldl}, {\AB{X} \AS = \AB{A} \AF ✶ \AS → \AB B}.
+These facts suggest that \AB{𝒟} can be computed one list element at a time via a more specialized operation, as is indeed the case:\footnote{Repeated \AF{δ} application is expressed here via a standard left fold (with {\AB{X} \AS = \AB{A} \AF ✶ \AS → \AB B} in \AF{𝒟foldl}):
+\agda{foldl}
 }
@@
 
-For an effective implementation, we must bridge the gap between a type (of membership proofs) and its decidable inhabitation.
+For effective implementations, we must bridge the gap between a type (of membership proofs) and its decidable inhabitation.
 Fortunately, there is a convenient and compositional way to do so.
-For any type \AB{A}, the type {\AF{Dec} \AB{A}} contains proof of \AB{A} or a proof of {\AF ¬ \AB{A}} (defined as usual to mean {\AB{A} \AS → \AD{⊥}}):\footnote{This simple \AF{Dec} definition been superceded by a more efficient but more complex version \stdlibCitep{Relation.Nullary}.}
+For type \AB{A}, the type {\AF{Dec} \AB{A}} contains proof of \AB{A} or a proof of {\AF ¬ \AB{A}} (defined as usual to mean {\AB{A} \AS → \AD{⊥}}):\footnote{This simple \AF{Dec} definition been superceded by a more efficient but more complex version \stdlibCitep{Relation.Nullary}.}
 \agda{Dec}

### 2021-02-28 paper.tex
tune margins for footnotes
a little more on proof relevance

@@
 \documentclass[acmsmall,screen,anonymous,timestamp,review]{acmart}
+%% \documentclass[acmsmall,screen,timestamp,nonacm]{acmart}
+%% \authorsaddresses{}
 \else
 \documentclass[hidelinks]{article}  % fleqn,12pt
-\usepackage[margin=0.9in]{geometry}  % 0.12in, 0.9in, 1in
+\usepackage[margin=1.3in]{geometry}  % 0.12in, 0.9in, 1in
 \usepackage[useregional]{datetime2}
@@
 \item
-  The theory and algorithms are specified in terms of proof isomorphism rather than mere logical equivalence.
-  As such, the differentiation lemmas for language concatenation and Kleene closure are more informative than in previous work, capturing \emph{all} parsings, including closure of languages containing the empty string.
+  The theory and algorithms are specified in terms of proof isomorphism rather than the usual, coarser relation of logical equivalence.
+  As such, the lemmas for language concatenation and Kleene closure are more informative than in previous work, capturing \emph{all} parsings distinctly, including closure of languages containing the empty string.
 \item
-  The vocabulary of languages is generalized to predicates over arbitrary types, and then re-specialized to the usual vocabulary (\secref{Predicate Algebra}).
+  The vocabulary of languages is generalized from predicates on lists to predicates on arbitrary types, and then re-specialized to the usual vocabulary (\secref{Predicate Algebra}).
   This generalization consists of two useful covariant applicative functors on functions (the usual one on functions-from-\AF{X} and another on functions-to-\APT{Set}).

### 2021-02-28 paper.tex
Use acmsmall for the public/non-anonymous version of my paper, tweaking some features
proofreading pass (misc)

@@
+\newif\iftalk
 \newif\ifacm
@@
 \ifacm
-\documentclass[acmsmall,screen,anonymous,timestamp,review]{acmart}
-%% \documentclass[acmsmall,screen,timestamp,nonacm]{acmart}
-%% \authorsaddresses{}
+\documentclass[acmsmall,screen,timestamp,anonymous,review]{acmart}
 \else
-\documentclass[hidelinks]{article}  % fleqn,12pt
-\usepackage[margin=1.3in]{geometry}  % 0.12in, 0.9in, 1in
-\usepackage[useregional]{datetime2}
+\documentclass[acmsmall,screen,timestamp,nonacm]{acmart}
 \fi
+\authorsaddresses{}
 
@@
 
-%% \usepackage{balance}  % even out final two-column page
-
-\ifacm\else
-%% Already in acmart
-\RequirePackage{amssymb, unicode-math}
-\fi
 \RequirePackage{newunicodechar, stmaryrd, setspace, comment, scalerel}
@@
 
-\ifacm
 \bibliographystyle{ACM-Reference-Format}
 \citestyle{acmauthoryear}
-\else
-\usepackage[round]{natbib} % [square]
-\bibliographystyle{plainnat}
-\fi
+\setcitestyle{nosort}
 
@@
 %% work with acmart.
-\ifacm
 \newunicodechar{◇}{\raisebox{0.02ex}{\ensuremath{_{\diamond}\hspace{-0.323em}}}}
 \newunicodechar{□}{\raisebox{-0.12ex}{\ensuremath{\scaleobj{0.4}{\Box}}\hspace{-0.238em}}}
-\else
-\setmathfont{XITSMath-Regular.otf}
-\newunicodechar{◇}{\raisebox{0.23ex}{\ensuremath{_{\diamond}\hspace{-0.358em}}}}
-\newunicodechar{□}{\raisebox{-0.13ex}{\ensuremath{\scaleobj{0.3}{\Box}}\hspace{-0.23em}}}
-\newunicodechar{∷}{\ensuremath{\mathbin{\hspace{-0.2em}:\hspace{-0.27em}:\hspace{-0.15em}}}}
-\fi
 
@@
 
-\ifacm\else
-\maketitle
-\fi
-
 \begin{abstract}
@@
 
-\ifacm
 \maketitle
 \mathindent2em
-\fi
 
@@
 For instance, every proof of {\AB{w} \AF ∈ \AB{P} \AF ∪ \AB{Q}} contains a proof of {\AB{w} \AF ∈ \AB{P}} or of {\AB{w} \AF ∈ \AB{Q}} \emph{and} the knowledge of which one was chosen (with different proofs reflecting different choices).
-Likewise, a proof of {w \AF ∈ \AB{P} \AF ⋆ \AB{Q}}\out{ (language concatenation)} includes the choice of strings \AB{u} and \AB{v}, a proof that {\AB{u} \AF ⊙ \AB{v} \AD ≡ \AB{w}}, and proofs that {\AB{u} \AF ∈ \AB{P}} and {\AB{v} \AF ∈ \AB{Q}}.
+Likewise, a proof of {w \AF ∈ \AB{P} \AF ⋆ \AB{Q}}\out{ (language concatenation)} includes the choice of strings \AB{u} and \AB{v}, a proof that {\AB{w} \AD ≡ \AB{u} \AF ⊙ \AB{v}}, and proofs that {\AB{u} \AF ∈ \AB{P}} and {\AB{v} \AF ∈ \AB{Q}}.
 
@@
 \end{itemize}
-Although many proofs are elided below, the source code for this paper is freely available and contains fully verified Agda code with no postulates.
+Although many proofs are elided below, the source code for this paper is freely available and contains fully verified Agda code with no postulates%
+\ifacm\else
+%% \citep{Elliott2021-derivatives-repo}.
+%% (\repoURL).
+.\footnote{\protect\repoURL}
+\fi
 All type signatures and definitions in the paper are generated by the Agda compiler.
@@
 
-Since {\AB{u} \AF ⊙ \AIC{[]} \AD ≡ \AB{u}}, it follows that\footnote{For functions \AB{f} and \AB{g}, {\AB{f} \AF{≗} \AB{g}} is extensional equality, i.e., {\AS ∀ \AB x \AS → \AB{f} \AB{x} \AD{≡} \AB{g} \AB{x}}.}
+Since {\AB{u} \AF ⊙ \AIC{[]} \AD ≡ \AB{u}}, it follows that\footnote{Agda variable names can contain most characters other than spaces and parentheses, e.g., ``\AB{ν∘𝒟}'' as declared here.
+For functions \AB{f} and \AB{g}, {\AB{f} \AF{≗} \AB{g}} is extensional equality, i.e., {\AS ∀ \AB x \AS → \AB{f} \AB{x} \AD{≡} \AB{g} \AB{x}}.}
 \agda{ν∘𝒟}
@@
 The equalities in \figref{nu-delta-lemmas} are all proved automatically by normalization (i.e., their proofs are simply \AIC{refl}), while the other relations require a bit more work.
-As with all lemmas in this paper shown with signatures only, full proofs are in the paper's source code and are formally verified by the Agda compiler.
+As with all lemmas in this paper shown with signatures only, full proofs are in the paper's source code and are formally verified by the Agda compiler (including during typesetting).
 
@@
 These lemmas and definitions cannot be applied \emph{automatically} in their present form, however, because languages are functions, as are \AF{ν} and \AF{δ}, and so are not subject to pattern-matching.
-An automatic solution would need some form of \emph{reflection} of language operations or of \AF{ν} and \AF{δ} as inspectable data.
+An automatic solution would need some form of \emph{reflection} of language-constructing operations or of \AF{ν} and \AF{δ} as inspectable data.
 
@@
 In the former, functions are represented symbolically in some suitable vocabulary, and pattern-matching is used to apply rules of differentiation to those symbolic representations.
-The latter solution involves re-interpreting the vocabulary of functions to construct not just the usual functions, but also their derivatives \citep{Griewank89onAD, GriewankWalther2008EvalDerivs, Elliott-2018-ad-icfp}.
-This latter option is often much more efficient than the former, since it easily avoids a good deal of redundant computation due to significant amounts of duplicated work between function ``primals'' and their derivatives.
+The latter solution involves re-interpreting the function-building vocabulary to construct not just the usual functions, but also their derivatives \citep{Griewank89onAD, GriewankWalther2008EvalDerivs, Elliott-2018-ad-icfp}.
+This latter option is often much more efficient than the former, since it easily avoids a good deal of redundant computation due to significant amounts of duplicated work between a function and its derivative.
 
@@
 
-To more clearly see the parallel with automatic differentiation (AD), note that AD implementations generally sample a function \emph{and} its derivative together to exploit the fact that these two computations typically share much common work.
+To more clearly see the parallel with automatic differentiation (AD), note that AD implementations sample a function \emph{and} its derivative together to exploit the fact that these two computations typically share much common work.
 This work sharing also applies when differentiating languages.
@@
 \figrefdef{predicates}{Predicate operations}{
-\ifacm
 \small
-\fi
 \begin{center}
@@
 \end{center}
-\ifacm
 \vspace{1.5ex}
-\else
-\vspace{-2ex}
-\fi
 \hfill
@@
 \end{center}
-\ifacm
 \vspace{1ex}
-\else
-\vspace{-2ex}
-\fi
 \hfill
@@
 Likewise for other algebraic structures.
-(As always, proofs are in this paper's source code.)
+(As always, compiler-verified proofs are in this paper's source code.)
 
@@
 The dimension of the semimodule is the cardinality of the domain type \AB{A} (typically infinite), with ``basis vectors'' having the form {\AF{pureⱽ} \AB{a}} for {\AB{a} \AS : \AB{A}}.
-A general vector (predicate) is a linear combination (often infinite) of these basis vectors, with addition being union and scaling defined by conjoining membership proofs (as in \figref{lang-via-pred}).
+A general vector (predicate) is a linear combination (often infinite) of these basis vectors, with addition being union and scaling defined by pairing membership proofs (as in \figref{lang-via-pred}).
 
@@
 We have seen that type-level languages have useful and illuminating algebraic properties and that the decidable implementations in \secref{From Languages to Parsing} are tightly connected to languages.
-Must we prove these properties again for each implementation---here there are more operational details to manage---or do the algebraic properties somehow transfer to those implementations?
+Must we prove these properties again for each implementation---where there are more operational details to manage---or do the algebraic properties somehow transfer to those implementations?
 This question immediately raises a technical obstacle.
@@
 \rnc\source{Transport}
-As examples, we can lift the commutative semiring of types to wrappings of the decidable counterparts from \figref{compositional-dec} in \secref{Decidability}, and lift both language semirings to the decidable implementations in \secref{From Languages to Parsing}, as in \figrefdef{transport-examples}{Examples of algebraic instances transported from indices to existential wrappings}{\agda{examples}}.
+As examples, we can lift the commutative semiring of types to wrappings of the decidable counterparts from \figref{compositional-dec} in \secref{Decidability}, and lift both language semirings to the decidable implementations in \secref{From Languages to Parsing}, e.g., as in \figrefdef{transport-examples}{Examples of algebraic instances transported from indices to existential wrappings}{\agda{examples}}.
 
@@
 
-There has also been some work using type theory to capture languages as type-level predicates much like in \secref{Specifying Languages}:
+There has also been some work using type theory to capture languages as type-level predicates much as in \secref{Specifying Languages}:
 \begin{itemize}
@@
 \item \citet{DoczkalEtAl2013} formalized regular languages constructively in Coq, also in terms of a type-level membership predicate.
-They then proved several classic results about finite automata, including minimization and relationship to Nerode and Myhill partitions.
+They then proved several classic results about finite automata, including minimization and relationships to Nerode and Myhill partitions.
 
@@
 
-Of the previous work involving language derivatives, none appear to have based their formal specification and proofs on the essential, non-inductive, nearly-trivial definition in terms of functions of lists from \secref{Decomposing Languages}, although \citet{Abel2016} did indeed mention this definition informally as motivation for the more complex, coinductively defined operations on tries.
+Of the previous work involving language derivatives, none appear to have based their formal specification and proofs on the essential, non-inductive, nearly-trivial definition in terms of functions of lists from \secref{Decomposing Languages}.
+\citet{Abel2016} mentioned this definition informally as motivation for the more complex, coinductively defined operations on tries.
 \citet{Brzozowski64} also made clear mention of the underlying meaning on sets of strings in his original work on regular expression derivatives.

### 2021-02-28 paper.tex
"(which are essentially \emph{proof-relevant} recognizers)"

@@
 Those laws were based on the assumption that union (``addition'' in both language semirings of \secref{Predicate Algebra}) is idempotent, which is not true for type-level language predicates related by proof isomorphism (rather than mere logical equivalence).
-The lack of idempotence is crucial for going beyond recognizers to parsers, in which we might want to extract more than one proof (parsing).
+The lack of idempotence is crucial for going beyond recognizers to parsers (which are essentially \emph{proof-relevant} recognizers), in which we might want to extract more than one proof (parsing).
 The difference is mainly visible in the \AF ν and \AF δ lemmas for {\AB P ⋆ \AB Q} and {\AB P ☆} in \figref{nu-delta-lemmas}.

### 2021-03-01 paper.tex
switching stars
vertical spacing in languages via predicate operations

@@
 
+%% I think acmart minimizes these two.
+
 %% Switching to \textsf for AgdaFunction messes up vertical alignment.
@@
 \begin{document}
+
+\setlength{\abovedisplayskip}{2ex}
+\setlength{\belowdisplayskip}{2ex}
+\setlength{\abovedisplayshortskip}{0ex}
+\setlength{\belowdisplayshortskip}{0ex}
 
@@
 
+In the definition of {\AB P \AF ☆}, \AB{concat} flattens a list of lists into a single list, defined as a right fold:
+
+\agda{foldr-concat}
+
 Note that for a language \AB{P} and string \AB{w}, {\AB{P} \AB{w}} is the type of \emph{proofs} that {\AB{w} \AF ∈ \AB{P}}, in other words \emph{explanations} of membership, or \emph{parsings}.
@@
 
-Since {\AB{u} \AF ⊙ \AIC{[]} \AD ≡ \AB{u}}, it follows that\footnote{Agda variable names can contain most characters other than spaces and parentheses, e.g., ``\AB{ν∘𝒟}'' as declared here.
-For functions \AB{f} and \AB{g}, {\AB{f} \AF{≗} \AB{g}} is extensional equality, i.e., {\AS ∀ \AB x \AS → \AB{f} \AB{x} \AD{≡} \AB{g} \AB{x}}.}
+Since {\AB{u} \AF ⊙ \AIC{[]} \AD ≡ \AB{u}}, it follows that\footnote{Agda variable names can contain most characters other than spaces and parentheses, e.g., ``\AB{ν∘𝒟}'' as declared here.}\footnote{For functions \AB{f} and \AB{g}, {\AB{f} \AF{≗} \AB{g}} is extensional equality, i.e., {\AS ∀ \AB x \AS → \AB{f} \AB{x} \AD{≡} \AB{g} \AB{x}}.}
 \agda{ν∘𝒟}
@@
 \setstretch{1.7}
-%% (any domain type)
 \agda{codomain-ops}
@@
 \begin{minipage}[c]{15em}
-%% (any monoid)
+\setstretch{2.5}
 \agda{domain-ops}
-%% (lists)
-\agda{list-ops}
 \end{minipage}

### 2021-03-01 paper.tex
some translation around All; trim some paragraphs for better page breaks.

@@
 
-\setlength{\abovedisplayskip}{2ex}
-\setlength{\belowdisplayskip}{2ex}
+\setlength{\abovedisplayskip}{1.5ex}
+\setlength{\belowdisplayskip}{1.5ex}
 \setlength{\abovedisplayshortskip}{0ex}
@@
 
-In the definition of {\AB P \AF ☆}, \AB{concat} flattens a list of lists into a single list, defined as a right fold:
-
+In the definition of {\AB P \AF ☆}, \AF{concat} flattens a list of lists into a single list via a right fold.
+The predicate {\AD{All} \AB P} holds of a list \AB{ws} when \AB{P} holds for every element of \AB{ws}:\footnote{The data type constructors \AIC{[]} and \AIC{\_∷\_} are overloaded here, referring to both lists (as indices) and \AD{All}.}\\
 \agda{foldr-concat}
+\vspace{-1ex}
+\agda{All}
+These three definitions are provided in the Agda standard library \citep{agda-stdlib}.
 
@@
 \agda{ν𝒟}
-In terms of languages, {\AF{𝒟} \AB{P} \AB{u}} comprises all strings \AB{v} such that {\AB{u} \AF ⊙ \AB{v} \AF ∈ \AB{P}}, i.e., the \AB{u}-suffixes of strings in \AB{P}.
+For languages, {\AF{𝒟} \AB{P} \AB{u}} is the set of \AB{u}-suffixes from \AB{P}, i.e., the strings \AB{v} such that {\AB{u} \AF ⊙ \AB{v} \AF ∈ \AB{P}}.
 
@@
 This simple observation is the semantic heart of the syntactic technique of \emph{derivatives of regular expressions} as used for efficient recognition of regular languages \citep{Brzozowski64}, later extended to parsing general context-free languages \citep{Might2010YaccID}.
-Lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} liberate this technique from the assumption that languages are represented \emph{symbolically}, by some form of grammar (e.g., regular or context-free), inviting other representations as we will see in \secref{Automatic Differentiation}.
+Lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} liberate this technique from the assumption that languages are represented \emph{symbolically}, by some form of grammar (e.g., regular or context-free), as we will exploit in \secref{Automatic Differentiation}.
 
@@
 Fortunately, there is a convenient and compositional way to do so.
-For type \AB{A}, the type {\AF{Dec} \AB{A}} contains proof of \AB{A} or a proof of {\AF ¬ \AB{A}} (defined as usual to mean {\AB{A} \AS → \AD{⊥}}):\footnote{This simple \AF{Dec} definition been superceded by a more efficient but more complex version \stdlibCitep{Relation.Nullary}.}
+For type \AB{A}, the type {\AF{Dec} \AB{A}} contains proof of \AB{A} or a proof of {\AF ¬ \AB{A}} (defined as usual to mean {\AB{A} \AS → \AD{⊥}}):\footnote{This simple \AF{Dec} definition been superceded by a more efficient version \stdlibCitep{Relation.Nullary}.}
 \agda{Dec}
@@
 With these definitions, we can formulate the problem of decidable language recognition: given a language \AB{P}, construct a term of type {\AF{Decidable} \AB{P}}.
-Since this transformation cannot be fully automated, we will instead look for decidable building blocks that mirror the predicate vocabulary defined in \secref{Specifying Languages}.
+To accomplish this transformation, we will look for decidable building blocks that mirror the predicate vocabulary defined in \secref{Specifying Languages}.
 
@@
 \agda{𝒟′}
-The more efficient, work-sharing version:
+The potential work-sharing version:
 \agda{ʻ𝒟}
-Their equivalence follows from lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} of \secref{Decomposing Languages}:
+Equivalence follows from lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} of \secref{Decomposing Languages}:
 \agda{𝒟′≡ʻ𝒟}
-The trie representation accomplishes this sharing by weaving \AF{ν} and \AF{δ} into a single structure based on common prefixes.
+The trie representation exploits this sharing potential by weaving \AF{ν} and \AF{δ} into a single structure based on common prefixes.
 
@@
 The first category applies to predicates over all types (not just over lists).
-Within the second category, \AF{𝟏}, \AF{\_⋆\_}, and \AF{\_☆} apply to all monoids\footnote{The \emph{definitions} of those operations do not require the monoid properties, but their properties will (\secref{Predicate Algebra}).}, while \AF{`} is specific to lists.
+Within the second category, \AF{𝟏}, \AF{\_⋆\_}, and \AF{\_☆} can apply to any monoid, while \AF{`} is specific to lists.
 
@@
 \hfill\;
-}, which adds language complement (\AF ∁).
+}, which adds language complement (\AF ∁).\footnote{The list-specific operations \AF{\_⊙\_}, \AIC{[]}, and \AF{concat} used in \figref{Lang-ops} have been generalized to monoid operations {\AF{\_∙\_} and \AF{ε}}.
+The \emph{definitions} of \AF{𝟏}, \AF{\_⋆\_}, and \AF{\_☆} do not require the monoid properties, but their properties will (\secref{Predicate Algebra}).
+}
 

### 2021-03-01 paper.tex
wording

@@
 
-\item \citet{FirsovUustalu2013} used Agda to formalize regular expressions with a corresponding inductive type of language membership proofs and then related regular expressions to nondeterministic finite automata (NFAs) represented as boolean matrices, as well as operations on that representation that correspond to those on regular expressions.
-They then proved soundness and completeness of the NFA representation with respect to regular language membership.
+\item \citet{FirsovUustalu2013} formalized regular expressions in Agda with a corresponding inductive type of language membership proofs and related regular expressions to nondeterministic finite automata (NFAs) represented as boolean matrices.
+They also defined operations on that representation corresponding to those on regular expressions and proved soundness and completeness of the NFA representation with respect to regular language membership.
 

### 2021-03-01 paper.tex
Swap "✪" and "☆"

@@
 
-A simple characterization of languages (and indeed functions from lists to any type) captures the essential idea behind language ``differentiation'' (and integration) as used for recognizing languages, leading to a collection of lemmas about type-level predicates.
+A simple characterization of languages (and indeed functions from lists to any type) captures the essential idea behind language ``differentiation'' as used for recognizing languages, leading to a collection of lemmas about type-level predicates.
 These lemmas are the heart of two dual parsing implementations---using (inductive) regular expressions and (coinductive) tries---each containing the same code but in dual arrangements (with representation and primitive operations trading places).
@@
 Alternatively, one can use type theory, so that a language is a type-level predicate on ``strings'' (lists of an arbitrary type \AB{A} of ``characters'').
-The usual language operations are defined in \figrefdef{Lang-ops}{Languages as type-level predicates}{\agda{Lang-ops}}, including language union and intersection ({\AB{P} \AF ∪ \AB{Q}} and {\AB{P} \AF ∩ \AB{Q}}) with their identities (the empty language \AF ∅ and universal language \AF 𝒰), language concatenation ({\AB{P} \AF ⋆ \AB{Q}}) and its identity (\AF 𝟏, containing only the empty string), single-character languages ({\AF ` \AB{c}}), Kleene star ({\AB{P} \AF ☆}), and ``scalar multiplication'' (\mbox{\AB{s} \AF · \AB{P}}, which will prove useful later).
+The usual language operations are defined in \figrefdef{Lang-ops}{Languages as type-level predicates}{\agda{Lang-ops}}, including language union and intersection ({\AB{P} \AF ∪ \AB{Q}} and {\AB{P} \AF ∩ \AB{Q}}) with their identities (the empty language \AF ∅ and universal language \AF 𝒰), language concatenation ({\AB{P} \AF ⋆ \AB{Q}}) and its identity (\AF 𝟏, containing only the empty string), single-character languages ({\AF ` \AB{c}}), Kleene star ({\AB{P} \AF ✪}), and ``scalar multiplication'' (\mbox{\AB{s} \AF · \AB{P}}, which will prove useful later).
 
@@
 
-In the definition of {\AB P \AF ☆}, \AF{concat} flattens a list of lists into a single list via a right fold.
+In the definition of {\AB P \AF ✪}, \AF{concat} flattens a list of lists into a single list via a right fold.
 The predicate {\AD{All} \AB P} holds of a list \AB{ws} when \AB{P} holds for every element of \AB{ws}:\footnote{The data type constructors \AIC{[]} and \AIC{\_∷\_} are overloaded here, referring to both lists (as indices) and \AD{All}.}\\
@@
 One category transforms the codomain (types): \AF{∅}, \AF 𝒰, \AF{\_∪\_}, and \AF{\_∩\_}.
-The other transforms the domain (lists): \AF{𝟏}, \AF{\_⋆\_}, \AF{\_☆}, and \AF{`}.
+The other transforms the domain (lists): \AF{𝟏}, \AF{\_⋆\_}, \AF{\_✪}, and \AF{`}.
 The first category applies to predicates over all types (not just over lists).
-Within the second category, \AF{𝟏}, \AF{\_⋆\_}, and \AF{\_☆} can apply to any monoid, while \AF{`} is specific to lists.
+Within the second category, \AF{𝟏}, \AF{\_⋆\_}, and \AF{\_✪} can apply to any monoid, while \AF{`} is specific to lists.
 
@@
 }, which adds language complement (\AF ∁).\footnote{The list-specific operations \AF{\_⊙\_}, \AIC{[]}, and \AF{concat} used in \figref{Lang-ops} have been generalized to monoid operations {\AF{\_∙\_} and \AF{ε}}.
-The \emph{definitions} of \AF{𝟏}, \AF{\_⋆\_}, and \AF{\_☆} do not require the monoid properties, but their properties will (\secref{Predicate Algebra}).
+The \emph{definitions} of \AF{𝟏}, \AF{\_⋆\_}, and \AF{\_✪} do not require the monoid properties, but their properties will (\secref{Predicate Algebra}).
 }
@@
 The boolean semiring is idempotent, while the type semiring is not (in order to distinguish multiple parsings in ambiguous languages).
-The latter non-idempotence accounts for the difference between the \AF ν and \AF δ lemmas for {\AB P ⋆ \AB Q} and {\AB P ☆} in \figref{nu-delta-lemmas} and the corresponding rules in previous work, as mentioned in \secref{Related Work}.}
+The latter non-idempotence accounts for the difference between the \AF ν and \AF δ lemmas for {\AB P ⋆ \AB Q} and {\AB P ✪} in \figref{nu-delta-lemmas} and the corresponding rules in previous work, as mentioned in \secref{Related Work}.}
 They are both closed as well.
@@
 The lack of idempotence is crucial for going beyond recognizers to parsers (which are essentially \emph{proof-relevant} recognizers), in which we might want to extract more than one proof (parsing).
-The difference is mainly visible in the \AF ν and \AF δ lemmas for {\AB P ⋆ \AB Q} and {\AB P ☆} in \figref{nu-delta-lemmas}.
+The difference is mainly visible in the \AF ν and \AF δ lemmas for {\AB P ⋆ \AB Q} and {\AB P ✪} in \figref{nu-delta-lemmas}.
 

### 2021-03-03 paper.tex
misc wording tweaks

@@
 The use of type-level predicates makes for simple, direct specification, including the use of existential quantification.
-As described in \secref{Properties}, it's also fairly easy to prove algebraic properties of predicates\out{, including the semimodule and semiring laws for both predicate semirings}.
+As described in \secref{Properties}, it is also fairly easy to prove algebraic properties of predicates\out{, including the semimodule and semiring laws for both predicate semirings}.
 These definitions do not give us decidable parsing, because general type inhabitation amounts to theorem proving.
@@
   Properties are stated first at the level of predicates, making them easy to understand and prove, thanks to elimination of operational details.
-  These properties are then transported effortlessly to the computable parsing implementations (\secref{Transporting Properties}).
+  These properties are then transported to the computable parsing implementations (\secref{Transporting Properties}) with no additional proofs needed.
 \end{itemize}
-Although many proofs are elided below, the source code for this paper is freely available and contains fully verified Agda code with no postulates%
+Although many proofs are elided below, the source code for this paper is freely available and contains fully verified Agda code with no postulates.%
 \ifacm\else
@@
 %% (\repoURL).
-.\footnote{\protect\repoURL}
-\fi
+\footnote{\protect\repoURL}
+\fi{}
 All type signatures and definitions in the paper are generated by the Agda compiler.
@@
 
-In order to convert languages into parsers, it will help to understand how language membership relates to the algebraic structure of lists, specifically the monoid formed by \AIC{[]} and \AF{\_⊙\_}.
+In order to convert languages into parsers, it will help to understand how language membership relates to the algebraic structure of lists, specifically the monoid formed by \AF{\_⊙\_} and \AIC{[]}.
 Generalizing from languages to functions from lists to \emph{any} type, define
@@
 \agda{𝒟[]⊙}
-These facts suggest that \AB{𝒟} can be computed one list element at a time via a more specialized operation, as is indeed the case:\footnote{Repeated \AF{δ} application is expressed here via a standard left fold (with {\AB{X} \AS = \AB{A} \AF ✶ \AS → \AB B} in \AF{𝒟foldl}):
+These facts suggest that \AB{𝒟} can be computed one list element at a time via a more specialized operation, as is indeed the case:\footnote{Repeated \AF{δ} application is expressed here via a standard left fold (with {\AB{X} \AS = \AB{A} \AF ✶ \AS → \AB B} for \AF{𝒟foldl}):
 \agda{foldl}
@@
 This simple observation is the semantic heart of the syntactic technique of \emph{derivatives of regular expressions} as used for efficient recognition of regular languages \citep{Brzozowski64}, later extended to parsing general context-free languages \citep{Might2010YaccID}.
-Lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} liberate this technique from the assumption that languages are represented \emph{symbolically}, by some form of grammar (e.g., regular or context-free), as we will exploit in \secref{Automatic Differentiation}.
+Lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} liberate this technique from the assumption that languages are represented \emph{symbolically}, by some form of grammar (e.g., regular or context-free), as will be exploited in \secref{Automatic Differentiation}.
 
@@
 The equalities in \figref{nu-delta-lemmas} are all proved automatically by normalization (i.e., their proofs are simply \AIC{refl}), while the other relations require a bit more work.
-As with all lemmas in this paper shown with signatures only, full proofs are in the paper's source code and are formally verified by the Agda compiler (including during typesetting).
+As with all lemmas in this paper shown with signatures only, full proofs are in the paper's source code and are formally verified by the Agda compiler (including during typesetting of this paper).
 
@@
 Fortunately, there is a convenient and compositional way to do so.
-For type \AB{A}, the type {\AF{Dec} \AB{A}} contains proof of \AB{A} or a proof of {\AF ¬ \AB{A}} (defined as usual to mean {\AB{A} \AS → \AD{⊥}}):\footnote{This simple \AF{Dec} definition been superceded by a more efficient version \stdlibCitep{Relation.Nullary}.}
+For type \AB{A}, the type {\AD{Dec} \AB{A}} contains proof of \AB{A} or a proof of {\AF ¬ \AB{A}} (defined as usual to mean {\AB{A} \AS → \AD{⊥}}):\footnote{The Agda standard library contains a more efficient version of \AD{Dec} \stdlibCitep{Relation.Nullary}.}
 \agda{Dec}
@@
 One direction of the isomorphism proves \AB{B} from \AB{A}, while the other proves {\AF ¬ B} from {\AF ¬ A}.
-In fact, logical \emph{equivalence} suffices for the results in this paper, but the stronger condition of isomorphism enables variations such as generating all parses/proofs rather than just one.
+In fact, logical \emph{equivalence} suffices for the results in this paper, but the stronger condition of isomorphism enables variations such as generating many parses/proofs rather than just one.
 
@@
 The latter solution involves re-interpreting the function-building vocabulary to construct not just the usual functions, but also their derivatives \citep{Griewank89onAD, GriewankWalther2008EvalDerivs, Elliott-2018-ad-icfp}.
-This latter option is often much more efficient than the former, since it easily avoids a good deal of redundant computation due to significant amounts of duplicated work between a function and its derivative.
+This latter option is often much more efficient than the former, since it easily avoids a good deal of work duplicated between a function and its derivative.
 
 \secreftwo{Symbolic Differentiation}{Automatic Differentiation} applies the symbolic and automatic strategies respectively to languages.
-In both cases, the algorithms are automatically proved correct by type checking as corollaries of the lemmas in \figref{nu-delta-lemmas}.
+Both algorithms are corollaries of the lemmas in \figref{nu-delta-lemmas} and are automatically proved correct by type checking.
 
@@
 Decidable parsing relies only on \AF{ν} and \AF{δ} (via {\AF{⟦\_⟧}}), which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\out{\notefoot{Describe this derivation.}}
-These definitions are meant to be read in ``column-major'' order, i.e., each column is one function definition.
+These clauses are meant to be read in ``column-major'' order, i.e., each column is a complete definition.
 Correctness is guaranteed by the types of \AF{ν} and \AF{δ} and so is proved automatically by type-checking \figref{symbolic-defs}.
-(We could thus use \emph{any} definitions of \AF{ν} and \AF{δ} that type- and termination-check, although not many definitions would.)
+(We could thus use \emph{any} definitions of \AF{ν} and \AF{δ} that type- and termination-check, although not many definitions would do so.)
 
@@
 
-Decidable recognition again relies only on \AF{ν} and \AF{δ}, which are defined in \figrefdef{automatic-defs}{Automatic differentiation (row-major/copatterns)}{\agda{defs}}.
-These definitions are meant to be read in ``row-major'' order, i.e., each row is one definition.
-Note that the definitions are syntactically identical to those in \figref{symbolic-defs} but are organized dually.
+Decidable recognition again relies only on \AF{ν} and \AF{δ}, which are defined for each language operation in \figrefdef{automatic-defs}{Automatic differentiation (row-major/copatterns)}{\agda{defs}} (though with a problem to be noted and fixed in the next paragraph).
+These clauses are meant to be read in ``row-major'' order, i.e., each row is a complete definition.
+Note that the clauses in \figref{automatic-defs} are syntactically identical to those in \figref{symbolic-defs} but are organized dually.
 (The compiler-generated syntax coloring differs to reflect the changed interpretation of the definitions.)
@@
 } shows two parallel collections of predicate operations, each covariantly transforming the codomain (left) or domain (right).\footnote{The types of operations on the left can be further relaxed from {\APT{Set} \AB ℓ} to any codomain type, while keeping \AB{A} fixed, resulting in the usual covariant applicative functor of functions \emph{from} a fixed type \citep{McBride2008APE}.
-Dually, the operations on the right (in their current generality) form a covariant applicative functor of functions \emph{to} types.}
+Dually, the operations on the right (in their current generality) form a second covariant applicative functor of functions \emph{to} types.}
 These generalized operations then specialize to language operations as in \figrefdef{lang-via-pred}{Languages via predicate operations}{
@@
 \rnc\source{Calculus}
-The \AF{ν} and \AF{δ} lemmas for codomain (but not domain) operations in \figref{nu-delta-lemmas} are easily generalized as shown in \figrefdef{nu-delta-codomain-lemmas}{Properties of \AF{ν} and \AF{δ} for predicate codomain operations}{\setstretch{1.5}\agda{νδ-codomain}} and are all proved automatically by normalization.
+The \AF{ν} and \AF{δ} lemmas in \figref{nu-delta-lemmas} for codomain (but not domain) operations are easily generalized as shown in \figrefdef{nu-delta-codomain-lemmas}{Properties of \AF{ν} and \AF{δ} for predicate codomain operations}{\setstretch{1.5}\agda{νδ-codomain}}, all proved automatically by normalization.
 Since the decidable language implementations in \secref{From Languages to Parsing} are corollaries of \AF{ν} and \AF{δ} lemmas, those implementations adapt easily to the generalized codomain operations (\AF{pureᵀ}, \AF{mapᵀ}, and \AF{mapᵀ₂}), resulting in a somewhat smaller implementation and broader coverage.
@@
 Again, nothing is needed from the codomain type \AB{B} beyond the properties of a semiring, so again predicates get these algebraic structures for free (already paid for by types).
-For every {\AB{P} \AS : \AF{Pred} \AB{A}}, \AB{P} is a vector in the semimodule {\AF{Pred} \AB{A}}, including the special case of ``languages'', for which \AB{A} is a list type.
+Every {\AB{P} \AS : \AF{Pred} \AB{A}} is a vector in the semimodule {\AF{Pred} \AB{A}}, including the special case of ``languages'', for which \AB{A} is a list type.
 The dimension of the semimodule is the cardinality of the domain type \AB{A} (typically infinite), with ``basis vectors'' having the form {\AF{pureⱽ} \AB{a}} for {\AB{a} \AS : \AB{A}}.
@@
 \rnc\source{Existential}
-Indexed operations of all arities for a type family \AB{F} can be systematically wrapped in this style, as shown in
+Indexed operations of all arities\out{ for a type family \AB{F}} can be systematically wrapped in this style, as shown in
 \figrefdef{inj}{Existential wrappers}{\mathindent0ex\agda{inj}}, again with automatically inferable components given explicitly for clarity.
-Then wrapped union can be defined simply as {\AF{\_∪\_} \AS = \AF{inj₂} \AIC{□.\_∪\_}} and likewise for the other operations.
+Wrapped union can then be defined simply as {\AF{\_∪\_} \AS = \AF{inj₂} \AIC{□.\_∪\_}} and likewise for the other operations.
 
@@
 \rnc\source{Transport}
-As examples, we can lift the commutative semiring of types to wrappings of the decidable counterparts from \figref{compositional-dec} in \secref{Decidability}, and lift both language semirings to the decidable implementations in \secref{From Languages to Parsing}, e.g., as in \figrefdef{transport-examples}{Examples of algebraic instances transported from indices to existential wrappings}{\agda{examples}}.
+As examples, we can lift the commutative semiring of types to wrappings of the decidable counterparts from \figref{compositional-dec} in \secref{Decidability}, and lift both language semirings to both decidable implementations in \secref{From Languages to Parsing}, as in \figrefdef{transport-examples}{Examples of algebraic instances transported from indices to existential wrappings}{\agda{examples}}.
 

### 2021-03-03 paper.tex
tiny tweaks

@@
 (\figref{automatic-defs} compiles only due to suppressing termination checking for the problematic clause with a compiler pragma.)
-Fortunately, this issue was already identified and solved by Andreas \citet{Abel2016}---also in the setting of trie-based language recognition---by using \emph{sized types} \citep{Abel2008, AbelPientka2016}.
+Fortunately, this issue was already identified and solved by \citet{Abel2016}---also in the setting of trie-based language recognition---by using \emph{sized types} \citep{Abel2008, AbelPientka2016}.
 \rnc\source{SizedAutomatic}
@@
 In the first predicate semiring, which is \emph{commutative} (i.e., multiplication commutes), zero and addition are \AF ∅ and \AF{\_∪\_}, while one and multiplication are \AF 𝒰 and \AF{\_∩\_}.
-Closure adds a star/closure operation \AF{\_✯} with \AF{starˡ} law: {\AB x \AF ✯ \AF ≈ 1 \AF + \AB x \AF ✲ \AB x \AF ✯}.
+Closure adds a star/closure operation \AF{\_✯} with \AF{star} law: {\AB x \AF ✯ \AF ≈ 1 \AF + \AB x \AF ✲ \AB x \AF ✯}.
 
@@
 They are both closed as well.
-For booleans, closure maps both \AIC{false} and \AIC{true} to \AIC{true}, with the \AF{starˡ} law holding definitionally.
-For types, the closure of \AB{A} is {\AB{A} ✶} (the usual inductive list type) with a simple, non-inductive proof of \AF{starˡ}.
+For booleans, closure maps both \AIC{false} and \AIC{true} to \AIC{true}, with the \AF{star} law holding definitionally.
+For types, the closure of \AB{A} is {\AB{A} ✶} (the usual inductive list type) with a simple, non-inductive proof of \AF{star}.
 

### 2021-05-31 response-to-reviews.md
processed review comments & requirements

# Paper #87 review responses

***Note:** I say "we" instead of "I" in these responses, because the blind review process was still in effect when I wrote them.*

> The authors’ response should address reviewer concerns, answer reviewer questions, and correct misunderstandings. Reviewers are not required to read past the 500 word limit, so please focus your response there. You may provide additional (e.g. point-by-point, per-reviewer) response beyond the limit if it may help the reviewers.

***Oops. I read this clarification late in the game. To do: reorganize the following remarks accordingly.***

## Review A

> My major concern is novelty, but unfortunately I am not familiar with the prior work. Given that I was a bit surprised by the absence of prior work, and that there were already so many papers on related topics, my rating is B. I might advocate for this paper more enthusiastically once novelty is confirmed.

We were surprised as well not to find formalized & mechanized connections between the simple, non-operational semantics of language specification and the dual decidable representations of regular expressions and tries.
Note also that proof relevance yields *parsing* (explanations) rather than mere recognition.
We are not aware of formalized (or even rigorous) accounts of *parsing* (as opposed to mere recognition) based on non-operational specification of languages and derivatives.

> I recognize that a note on the versions of Agda and its standard library would help the reader.

Will do. Thank you for the suggestion!

## Review B

> It seems that in Fig. 1 "a language over an alphabet A" is not A^* but A^* -> Set l. This is at least nonstandard --- a standard definition should be A*. Explain the reason.

`A^*` is the set of *all* strings over an alphabet `A`, i.e., the "universal language".
Other languages are proper subsets, which are neatly formalized in type theory as `A^* -> Set l`, i.e., a function mapping each string `w` to the type of proofs that `w` is in the language.

> By the way, can you use Prop instead of Set l? I suspect the former is more intuitive, especially for those who are not experts on type theory.

`Set` appears to be much more widely used in the Agda ecosystem than `Prop` and so using `Set` allows reusing many more existing definitions and proofs, including type isomorphisms.
More fundamentally, `Prop` is proof-irrelevant, so we would only get language recognition.
In contrast, using `Set` gives proof-relevant recognition, i.e., parsing.

> In Fig. 1, while I believe c :: A, I could not figure out the type of scalar, s.

In the definition of scalar multiplication (`(s · P) w = s × P w`), `s` appears as an argument to the type product operator and so is a type (i.e., has type `Set ℓ`).
Logically, it's a proposition being conjoined with the property of the string `w` being in the language `P`.
This operator makes for more elegant and more general formulations of derivatives of concatenated and starred languages than Brzozowski's definitions.
The additional generality comes from proof relevance ("parsing").
It's also fundamentally important to the algebraic setting of semimodules (generalized vector spaces), which is a natural fit to proof-relevant languages.

> This paper carefully distinguished three kinds of equalities, namely propositional equality, isomorphism, and extensional isomorphism. However, I did not understand why that equality should be used for each occurrence. For example, why in Fig. 1 is propositional equality necessary?

Figure 1 uses both propositional equality and extensional isomorphism.
We could not have used (propositional) equality for all of the lemmas, since this strong condition does not hold in all cases.
We could have used the weaker isomorphism condition for all of the lemmas, but the equalities are more specific and thus more informative.
Moreover, isomorphisms require some notational overhead to apply, as shown where they are invoked in figures 5 and 7.
In fact, there might be some notational overhead even for propositional equalities, but these ones are actually *definitional* (meaning proved by normalization, as noted on line 194), making them especially convenient.

The reason for using both isomorphism and extensional isomorphism is simply that the `ν` lemmas are about types/propositions (nullability) while the `δ` lemmas are about predicates (general, proof-relevant language membership).

*Correction:* My answer was about Figure 3.
As for Figure 1, the uses of propositional equality are for lists (not types or type-level predicates), to which the other mentioned equality notions do not apply.

> Section 2 uses A* -> B instead of A* -> Set l. Perhaps, the reason is that all the definitions can be polymorphic on the result, B. However, this difference at least makes the meaning difficult to understand.

Indeed, this generalization is important because it points out that the central notion of derivatives is much more general than its usual setting of languages (even beyond generalizing from `Set` to arbitrary semirings), and some definitions and properties hold in the more general setting.

> p5. "isomorphic types or predicates are equivalently decidable":
> I could not understand the reason.

The reason (proof) is outlined in the next sentence ("One direction of the isomorphism proves \AB{B} from \AB{A}, while the other proves {\AF ¬ B} from {\AF ¬ A}") and formalized in the definitions/proofs of the `_▹_` and `_▸_` signatures/lemmas stated in between.
As with the rest of the paper, we mostly state the lemmas & theorems (signatures), leaving the interested reader to view proofs as desired in the paper/project's Agda source code.

> Figs. 4 and 5 are nearly identical to Figs. 6 and 7. While I could understand that it is because of the duality, I still wonder what is the essential differences.

The essential difference is that the representation data constructors and the representation examiners trade places (the duality).
More specifically

*   For regular expressions, language building blocks (`∅`, `_∪_`, `𝟙`, `_*_`, etc) are data constructors, while `ν` and `δ` are functions that examine the data.
*   For tries, `ν` and `δ` are codata constructors, while language building blocks are functions that examine the codata.

This duality is visually highlighted in the code arrangement in figures 4 and 6 and in the syntax coloring in figures 5 and 7.
(Line 419: "The compiler-generated syntax coloring differs to reflect the changed interpretation of the definitions.")

> Usually, an inductive definition gives a datatype, whereas a coinductive one does a function (or stream). However, the automatic differentiation requires the concrete construction process of the function even for the coinductive case. Another possible difference is the potential of the infiniteness of the coinductive definition. However, for this particular case, I am not aware of the usefulness of infiniteness.

The same pattern occurs in this paper.
Streams are a special case of string tries (for a singleton alphabet), being infinite in the specific and more general case.
The pattern also generalizes beyond string tries to tries over arbitrary regular algebraic data types.

By "concrete construction process", do you mean that there are still (corecursively defined) examination functions?
If so, this situation is also customary in our experience when working with coinductive structures.

> On page 9. it is argued that a difference is a work-sharing between the function and its derivative. I agree with this point but do not think it to be critical. Existing parsing methods based on the derivatives do such work sharing. Moreover, automatic differentiation alone cannot share repeated calculations of derivatives when we arrive at the same state more than once.

We thought this work sharing issue worth pointing out, because it makes an enormous difference for symbolic vs automatic differentiation in the usual calculus sense, and indeed is the main argument usually given in the literature for preferring the latter over the former.
We are not aware of a clear or rigorous discussion of the issue in derivative-based parsing work (though there are many optimizations described informally and without semantic basis).
It is not clear what the practical implications are in the setting of languages, but it seems worth mentioning to stimulate closer examination.
If the performance difference is much less than with differential calculus, then it's important to understand why.
For instance, maybe there is a dual technique for regular expressions that could be examined clearly for useful practical and theoretical insight.
On the other hand, if the difference is the same or greater, then it's important to exploit the potential with better justification and perhaps implementation.

We certainly agree with your remark about the potential for improved work sharing when a "state" is reached in more than one way!

## Review C

> The specification of language membership (language operations) is non-operational and it seems to be one of the key ideas that the authors are advocating, i.e., to have as simple as possible specifications with respect to which other operations (or implementations) can be proved correct.

Yes, exactly.

> The result of this approach is that the provided language membership deciders (defined using differentiation on inductive or coinductive representations of the language) are correct. In other words, the function that decides language membership is a theorem saying that if there is a representation for a language P, then language membership in P is decidable, i.e., for any w, the function either computes a witness of P w (the type of language membership proofs of w in P) or constructs a proof that no such exists.

Indeed. Well-summarized. Thank you.

> Overall, I think this is a nice, clean, well-written and compact paper. Working things out so that clean and simple definitions are also "magically" correct according to Agda is often a non-trivial task. The idea that specifications should be as simple as possible is a very reasonable one. Therefore, I think these ideas deserve to be communicated.

Thank you!
The clean, simple, automatically verified nature is exactly what we were wanting to highlight.

> One issue I have is that reading the paper might give the impression that the choice of type theory (instead of set theory) or type-level languages is necessary for achieving non-operational specification of language membership or more informative language membership than just true/false. >The paper does demonstrate that type theory can be convenient for achieving this.
>
> For example, the concatenation of P and Q (subsets of A*) is a certain subset of A* 
>
> P * Q = { w \in A* | exists u, v. w = uv \and u \in P \and v \in Q }
>
> and the characteristic function of this set corresponds to what is given in Figure 1. This could be done in Agda with decidable languages (A* -> 2) and thus also achieve non-operational specification (values of A* -> 2 as type indices). (Assumption: a word w can be decomposed in a finite number of ways.)

We wrote a comparison with set-theoretical and decidable specifications and discarded it for better focus.
Yes, one can start with decidable languages and exploit the property that every finite list can be unappended in finitely many ways.
Doing so would go against the principle of simple, elegant, non-operational *specifications*, since it relies on cleverness involving and motivated by decidability of language *recognition* rather than the simpler nature of language *definition*.

> line 18: I think it would be helpful to later mention what properties would have required induction/coinduction?

Thanks. We might refer to Abel [2016], which proves a great number of algebraic properties on tries, all of them needing coinduction/bisimulation.
The key to eliminating these much more complex proofs is inheriting properties from the *indices* (languages specified as type-level predicates), which are simple and neither inductive nor coinductive.

> Contribution 1, line 93: Is this meant to say that capturing the semantic essence of Brzozowski's derivatives is something new? The semantic operation was defined before the syntactic one. Brzozowski references three papers where (under different names) derivatives of sequences are used and then defines the corresponding operation on syntax.

*Formalizing* (stating and proving in a formal system) the semantic essence is new, as far as we know. Generalizing this essence from languages to functions from lists might be new as well. Moreover, most recent work we know of seems to identify derivatives with regular expressions, leading readers away from the more natural and general notion in practice and its alternative realizations, such as tries. (Abel [2016] is an exception but lacks a formal semantic specification separate from the implementation itself.)

See also the paragraph starting on line 717, including the mention of Brzozowski on line 720.
I imagine you will agree that it is easier to make clear comparisons after the technical substance is presented (in conclusions) than before (in introduction); and yet readers want at least a general sense of comparison before the technical substance.

> Contribution 2, line 99: The duality is between regular expressions and a certain subset of tries (those constructed using the language building operations). The two things named Lang in Figures 4 and 6 are not equally expressive.

We agree. This comparison can be expressed more clearly, including the fact that regular expressions *and their derivatives* also naturally generalize to a much broader class of languages, as Might and Darais [2010] and follow-up work has noted (though without much specification or rigor).

> line 103: Is "proof isomorphism" (throughout the paper) the same as "type isomorphism" on line 188?

Yes. "Type isomorphism" is more technically specific (suiting the technical meat of the paper), while its use as "proof isomorphism" seems more suitable for the introduction and the contrast with "logical equivalence" mentioned in line 103.

> Contribution 3, line 103: Is this statement considering previous work on formalisations or formal languages in general? As soon as something more general is considered instead of nullability (true/false empty word property), the lemmas (or definitions) for concatenation and closure start to look like those in Figure 2.

The statement refers to work we are aware of in formal languages as well as its formalization. Have we missed something important?

> line 179: What is meant exactly by "liberation"? I think it is fairly standard to consider left and right quotients (wrt. a word) of languages (sets of strings, no grammars) as monoid actions and thus reducing language membership to nullability also in a step-by-step manner.

This sentence is in the context of the immediately preceding sentence, which refers to Brzozowski's paper ("Derivatives of *Regular Expressions*"---emphasis mine) and its extensions by Might and Darais [2010] etc.
We are trying to encourage a shift from recent near-identification of derivatives (for parsing) as about a particular grammatical/inductive language representation back to the underlying semantic basis you mention (though also tilted toward proof relevance for parsing and other applications), so that the formal development has a simpler formal starting point (and neither inductive nor coinductive) and alternatives to grammatical representation are easier to notice.

> line 283: I think it would be good to also say what is the correctness condition (in addition to the fact that it is guaranteed by types).

The condition is stated on line 228 and is captured in the type of `⟦_⟧` on line 310, namely that `⟦_⟧` is both computable and converts a language into a decidable recognizer/parser.
We will add a reminder at line 283.
Thanks.

> line 308: This left triangle looks like a map operation. Maybe a comment about why this has to be part of the syntax.

The left triangle *reifies* a map operation.
The reason it must be part of the inductive representation is the same as the other constructors, namely so that the core lemmas (figure 2) translate into an implementation.
We will add a comment.

> line 333: Maybe a comment about why f works on the left hand side of the triangle with both p and with p differentiated.

Thanks.
The reason is that both `p` and `δ p a` are languages/predicates.

> Figures 4 and 5: there is no definition of the language (A* -> Set) denoted by an expression and thus no formal statement that it indeed denotes the language in its type.

Thank you.
This missing semantic function maps every `p : Lang P` to `P`, regardless of the specifics of `p` (which is why the algebraic properties transfer effortlessly from type-level languages without having to consider the structure of `p`).
We agree that it's worth stating explicitly and will do so.

> line 623: By a choice of an equivalence relation we are quotienting the (encapsulated) representation. It would be good to give a bit more intuition here (which quotient is this).

Indeed. This comment is closely related to your remark above about the denotation of the decidable language representations and has the same answer. Equivalence of representations is *semantic* equivalence, which is to say equivalence of the indices (type-level languages) that capture the semantics.
We will add this clarification.

> line 723: What are transformations?

The rules of differentiation, as transformations of symbolic/grammatical representations.

> line 724: "based *on* the"?

Fixed. Thanks!

## Review D

> Section 6.2 is quite dense, and its motivation is not entirely clear to me. Is its purpose to show that existentials can be used to force the dependently-typed definitions developed throughout the paper into the simply-typed algebraic structures? If so, I believe a single paragraph would to, and would perhaps be less confusing. Furthermore, the section title is confusing: transport properties to what? We haven't defined the existentially quantifier variants yet, so it's not yet obvious that we have anywhere to transport properties to.

The purpose is to show that the properties of non-decidable, type-level languages transfer effortlessly to the decidable implementations.
The importance of this fact is that one gets both (a) easy proofs of useful correctness properties (without complications due to efficiency and even decidability) *and* (b) efficient implementations.

This simple solution for property transfer, however, raises a "technical obstacle" (line 581): the indexed data types do not match the types of standard algebraic properties.
Existential wrapping is a simple, generic solution to this obstacle.

I renamed the section from "Transporting Properties" to "Transporting Properties from Specification to Implementations"

> Line 86: The clause "including the use of existential quantifiers" is confusing, as it in no way makes clear what the existential quantifiers are used for. I'd suggest dropping the clause entirely.

I've appended "(as in Figure 1) to that sentence".

> Line 500: The types of the primitive language constructs are elided throughout the paper. Therefore, it is a bit surprising to see that the variant of star which, ostensibly, abstracts away from lists, still references the list primitives foldr and All, as well as the monoid properties.

Before generalizing, the list plays two different algebraic roles: monoid and foldable.
Either can be generalized independently.

> Line 608: Why are the existentially-wrapped versions specifically based on the inductive variant? Is there anything specific that makes these better or more amenable? The text so far led me to believe the coinductive variant was better and more performant.

It was an arbitrary choice. Both implementations (inductive and coinductive) are wrapped alike. We will say so in the text.

> Line 638: Is the pattern matching in the type valid Agda syntax? Again, I don't think these liftings really add anything to the paper.

Yes. All such declarations and definitions in the paper (and those elided from the paper) are fully type- (and thus correctness-)checked when the paper is typeset.

* * * * * * * * * * * * * * * * * * * *

From the final review:
<blockquote>
The reviewers want to thank the author for this interesting submission and are happy to accept it with the understanding that the author will make all of the following changes:

1. Clarify what's new in the context of proof mechanization or type theory, what's new in the context of parsing, and what's new in the context of formal languages & automata. In particular, some contributions do not seem to be novel in the last context. The author did not respond to Reviewer C's comment about the work of Lombardy and Sakarovitch, but their work seems to be similar to the derivative operations in the paper and the author should pay attention to it (in addition to other related works).

2. Explain the Agda syntax more for the readers who are not familiar with Agda. For example, add concrete examples showing how operators in Fig. 1 apply to concrete regular expressions and words, including the fully expanded types and values (as witnesses of the membership). The current draft has only 16 pages so space should not be an issue.

3. Clarify the difference between data and codata: both generally and in terms of difference it makes for main results of this paper.

4. Consider explaining how to recover the ordinary setting where a language is a set of words and language union is idempotent. For example, explain how the current work can be used to prove the classical theorem for the ordinary setting that the set of derivatives of a regular expression is finite modulo associativity, commutativity and idempotence of union (although the derivative defined here might require more).

5. Fix typos and other issues that have already been discussed in the reviews and the response.
</blockquote>

*My response:*

1.  I added the following paragraph to the beginning of section 7 (Related Work):

    > The shift from languages as sets of strings to type-level predicates (i.e., proof relevance or ``parsing'') is akin to weighted automata \citep{Schutzenberger1961, DrosteKuske2019} and more generally to semiring-based parsing \citep{Chomsky1959CFL, Goodman1998PIO, Goodman1999SP, Liu2004}, noting that types (``\APT{Set}'' in Agda) form a (commutative) semiring (up to isomorphism).
    > In particular, \citet{Lombardy2005} investigated Brzozowski-style derivatives is this more general setting, formalizing and generalizing the work of \citet{Antimirov1996}, who decomposed derivatives into simpler components (``partial derivatives'') that sum to the full derivative and lead to efficient recognizers in the form of nondeterministic automata.

    Also an initial sentence to the (now) third paragraph of that section:

    > The main contributions of this paper are to formalization and mechanization of language recognition and parsing.

2.  I added a new figure 2 with examples of languages and membership proofs, along with a new footnote 2 giving some detailed explanations.

3.  I added a new footnote 10.

4.  Appended to the final paragraph of section 7:

    > As mentioned at the end of \secref{Decidability}, one can simplify the development above somewhat by replacing proof isomorphism with the coarser notion of logical equivalence.
    > Doing so restores union idempotence (modulo equivalence), losing proof relevance and thus replacing parsing by recognition.
    > One might then also prove that the set of derivatives of a regular language is finite (again, modulo equivalence).

5.  Done!

### 2021-05-31 short-review-response.md
processed review comments & requirements

--- 2021-05-31 response-to-reviews.md
@@
 # Paper #87 review responses
-
-***Note:** I say "we" instead of "I" in these responses, because the blind review process was still in effect when I wrote them.*
-
-> The authors’ response should address reviewer concerns, answer reviewer questions, and correct misunderstandings.
-Reviewers are not required to read past the 500 word limit, so please focus your response there.
-You may provide additional (e.g.
-point-by-point, per-reviewer) response beyond the limit if it may help the reviewers.
-
-***Oops. I read this clarification late in the game. To do: reorganize the following remarks accordingly.***
 
@@
 We are not aware of formalized (or even rigorous) accounts of *parsing* (as opposed to mere recognition) based on non-operational specification of languages and derivatives.
-
-> I recognize that a note on the versions of Agda and its standard library would help the reader.
-
-Will do. Thank you for the suggestion!
+> It seems that in Fig. 1 "a language over an alphabet A" is not A^* but A^* -> Set l. This is at least nonstandard --- a standard definition should be A*. Explain the reason.
 
@@
 It's also fundamentally important to the algebraic setting of semimodules (generalized vector spaces), which is a natural fit to proof-relevant languages.
-
-> This paper carefully distinguished three kinds of equalities, namely propositional equality, isomorphism, and extensional isomorphism.
-However, I did not understand why that equality should be used for each occurrence.
-For example, why in Fig.
-1 is propositional equality necessary?
-
-Figure 1 uses both propositional equality and extensional isomorphism.
-We could not have used (propositional) equality for all of the lemmas, since this strong condition does not hold in all cases.
-We could have used the weaker isomorphism condition for all of the lemmas, but the equalities are more specific and thus more informative.
-Moreover, isomorphisms require some notational overhead to apply, as shown where they are invoked in figures 5 and 7.
-In fact, there might be some notational overhead even for propositional equalities, but these ones are actually *definitional* (meaning proved by normalization, as noted on line 194), making them especially convenient.
-
-The reason for using both isomorphism and extensional isomorphism is simply that the `ν` lemmas are about types/propositions (nullability) while the `δ` lemmas are about predicates (general, proof-relevant language membership).
-
-*Correction:* My answer was about Figure 3.
-As for Figure 1, the uses of propositional equality are for lists (not types or type-level predicates), to which the other mentioned equality notions do not apply.
-
-> Section 2 uses A* -> B instead of A* -> Set l.
-Perhaps, the reason is that all the definitions can be polymorphic on the result, B.
-However, this difference at least makes the meaning difficult to understand.
-
-Indeed, this generalization is important because it points out that the central notion of derivatives is much more general than its usual setting of languages (even beyond generalizing from `Set` to arbitrary semirings), and some definitions and properties hold in the more general setting.
-
-> p5. "isomorphic types or predicates are equivalently decidable":
-> I could not understand the reason.
-
-The reason (proof) is outlined in the next sentence ("One direction of the isomorphism proves \AB{B} from \AB{A}, while the other proves {\AF ¬ B} from {\AF ¬ A}") and formalized in the definitions/proofs of the `_▹_` and `_▸_` signatures/lemmas stated in between.
-As with the rest of the paper, we mostly state the lemmas & theorems (signatures), leaving the interested reader to view proofs as desired in the paper/project's Agda source code.
-
-> Figs. 4 and 5 are nearly identical to Figs. 6 and 7. While I could understand that it is because of the duality, I still wonder what is the essential differences.
-
-The essential difference is that the representation data constructors and the representation examiners trade places (the duality).
-More specifically
-
-*   For regular expressions, language building blocks (`∅`, `_∪_`, `𝟙`, `_*_`, etc) are data constructors, while `ν` and `δ` are functions that examine the data.
-*   For tries, `ν` and `δ` are codata constructors, while language building blocks are functions that examine the codata.
-
-This duality is visually highlighted in the code arrangement in figures 4 and 6 and in the syntax coloring in figures 5 and 7.
-(Line 419: "The compiler-generated syntax coloring differs to reflect the changed interpretation of the definitions.")
-
-> Usually, an inductive definition gives a datatype, whereas a coinductive one does a function (or stream).
-However, the automatic differentiation requires the concrete construction process of the function even for the coinductive case.
-Another possible difference is the potential of the infiniteness of the coinductive definition.
-However, for this particular case, I am not aware of the usefulness of infiniteness.
-
-The same pattern occurs in this paper.
-Streams are a special case of string tries (for a singleton alphabet), being infinite in the specific and more general case.
-The pattern also generalizes beyond string tries to tries over arbitrary regular algebraic data types.
-
-By "concrete construction process", do you mean that there are still (corecursively defined) examination functions?
-If so, this situation is also customary in our experience when working with coinductive structures.
-
-> On page 9.
-it is argued that a difference is a work-sharing between the function and its derivative.
-I agree with this point but do not think it to be critical.
-Existing parsing methods based on the derivatives do such work sharing.
-Moreover, automatic differentiation alone cannot share repeated calculations of derivatives when we arrive at the same state more than once.
-
-We thought this work sharing issue worth pointing out, because it makes an enormous difference for symbolic vs automatic differentiation in the usual calculus sense, and indeed is the main argument usually given in the literature for preferring the latter over the former.
-We are not aware of a clear or rigorous discussion of the issue in derivative-based parsing work (though there are many optimizations described informally and without semantic basis).
-It is not clear what the practical implications are in the setting of languages, but it seems worth mentioning to stimulate closer examination.
-If the performance difference is much less than with differential calculus, then it's important to understand why.
-For instance, maybe there is a dual technique for regular expressions that could be examined clearly for useful practical and theoretical insight.
-On the other hand, if the difference is the same or greater, then it's important to exploit the potential with better justification and perhaps implementation.
-
-We certainly agree with your remark about the potential for improved work sharing when a "state" is reached in more than one way!
-
-## Review C
-
-> The specification of language membership (language operations) is non-operational and it seems to be one of the key ideas that the authors are advocating, i.e., to have as simple as possible specifications with respect to which other operations (or implementations) can be proved correct.
-
-Yes, exactly.
-
-> The result of this approach is that the provided language membership deciders (defined using differentiation on inductive or coinductive representations of the language) are correct.
-In other words, the function that decides language membership is a theorem saying that if there is a representation for a language P, then language membership in P is decidable, i.e., for any w, the function either computes a witness of P w (the type of language membership proofs of w in P) or constructs a proof that no such exists.
-
-Indeed. Well-summarized. Thank you.
-
-> Overall, I think this is a nice, clean, well-written and compact paper.
-Working things out so that clean and simple definitions are also "magically" correct according to Agda is often a non-trivial task.
-The idea that specifications should be as simple as possible is a very reasonable one.
-Therefore, I think these ideas deserve to be communicated.
-
-Thank you!
-The clean, simple, automatically verified nature is exactly what we were wanting to highlight.
-
-> One issue I have is that reading the paper might give the impression that the choice of type theory (instead of set theory) or type-level languages is necessary for achieving non-operational specification of language membership or more informative language membership than just true/false.
->The paper does demonstrate that type theory can be convenient for achieving this.
->
-> For example, the concatenation of P and Q (subsets of A*) is a certain subset of A* 
->
-> P * Q = { w \in A* | exists u, v. w = uv \and u \in P \and v \in Q }
->
-> and the characteristic function of this set corresponds to what is given in Figure 1.
-This could be done in Agda with decidable languages (A* -> 2) and thus also achieve non-operational specification (values of A* -> 2 as type indices).
-(Assumption: a word w can be decomposed in a finite number of ways.)
-
-We wrote a comparison with set-theoretical and decidable specifications and discarded it for better focus.
-Yes, one can start with decidable languages and exploit the property that every finite list can be unappended in finitely many ways.
-Doing so would go against the principle of simple, elegant, non-operational *specifications*, since it relies on cleverness involving and motivated by decidability of language *recognition* rather than the simpler nature of language *definition*.
-
-> line 18: I think it would be helpful to later mention what properties would have required induction/coinduction?
-
-Thanks. We might refer to Abel [2016], which proves a great number of algebraic properties on tries, all of them needing coinduction/bisimulation.
-The key to eliminating these much more complex proofs is inheriting properties from the *indices* (languages specified as type-level predicates), which are simple and neither inductive nor coinductive.
-
-> Contribution 1, line 93: Is this meant to say that capturing the semantic essence of Brzozowski's derivatives is something new?
-The semantic operation was defined before the syntactic one.
-Brzozowski references three papers where (under different names) derivatives of sequences are used and then defines the corresponding operation on syntax.
-
-*Formalizing* (stating and proving in a formal system) the semantic essence is new, as far as we know.
-Generalizing this essence from languages to functions from lists might be new as well.
-Moreover, most recent work we know of seems to identify derivatives with regular expressions, leading readers away from the more natural and general notion in practice and its alternative realizations, such as tries.
-(Abel [2016] is an exception but lacks a formal semantic specification separate from the implementation itself.)
-
-See also the paragraph starting on line 717, including the mention of Brzozowski on line 720.
-I imagine you will agree that it is easier to make clear comparisons after the technical substance is presented (in conclusions) than before (in introduction); and yet readers want at least a general sense of comparison before the technical substance.
-
-> Contribution 2, line 99: The duality is between regular expressions and a certain subset of tries (those constructed using the language building operations).
-The two things named Lang in Figures 4 and 6 are not equally expressive.
-
-We agree.
-This comparison can be expressed more clearly, including the fact that regular expressions *and their derivatives* also naturally generalize to a much broader class of languages, as Might and Darais [2010] and follow-up work has noted (though without much specification or rigor).
-
-> line 103: Is "proof isomorphism" (throughout the paper) the same as "type isomorphism" on line 188?
-
-Yes.
-"Type isomorphism" is more technically specific (suiting the technical meat of the paper), while its use as "proof isomorphism" seems more suitable for the introduction and the contrast with "logical equivalence" mentioned in line 103.
-
-> Contribution 3, line 103: Is this statement considering previous work on formalisations or formal languages in general?
-As soon as something more general is considered instead of nullability (true/false empty word property), the lemmas (or definitions) for concatenation and closure start to look like those in Figure 2.
-
-The statement refers to work we are aware of in formal languages as well as its formalization. Have we missed something important?
-
-> line 179: What is meant exactly by "liberation"?
-I think it is fairly standard to consider left and right quotients (wrt.
-a word) of languages (sets of strings, no grammars) as monoid actions and thus reducing language membership to nullability also in a step-by-step manner.
-
-This sentence is in the context of the immediately preceding sentence, which refers to Brzozowski's paper ("Derivatives of *Regular Expressions*"---emphasis mine) and its extensions by Might and Darais [2010] etc.
-We are trying to encourage a shift from recent near-identification of derivatives (for parsing) as about a particular grammatical/inductive language representation back to the underlying semantic basis you mention (though also tilted toward proof relevance for parsing and other applications), so that the formal development has a simpler formal starting point (and neither inductive nor coinductive) and alternatives to grammatical representation are easier to notice.
-
-> line 283: I think it would be good to also say what is the correctness condition (in addition to the fact that it is guaranteed by types).
-
-The condition is stated on line 228 and is captured in the type of `⟦_⟧` on line 310, namely that `⟦_⟧` is both computable and converts a language into a decidable recognizer/parser.
-We will add a reminder at line 283.
-Thanks.
-
-> line 308: This left triangle looks like a map operation. Maybe a comment about why this has to be part of the syntax.
-
-The left triangle *reifies* a map operation.
-The reason it must be part of the inductive representation is the same as the other constructors, namely so that the core lemmas (figure 2) translate into an implementation.
-We will add a comment.
-
-> line 333: Maybe a comment about why f works on the left hand side of the triangle with both p and with p differentiated.
-
-Thanks.
-The reason is that both `p` and `δ p a` are languages/predicates.
-
-> Figures 4 and 5: there is no definition of the language (A* -> Set) denoted by an expression and thus no formal statement that it indeed denotes the language in its type.
-
-Thank you.
-This missing semantic function maps every `p : Lang P` to `P`, regardless of the specifics of `p` (which is why the algebraic properties transfer effortlessly from type-level languages without having to consider the structure of `p`).
-We agree that it's worth stating explicitly and will do so.
-
-> line 623: By a choice of an equivalence relation we are quotienting the (encapsulated) representation. It would be good to give a bit more intuition here (which quotient is this).
-
-Indeed.
-This comment is closely related to your remark above about the denotation of the decidable language representations and has the same answer.
-Equivalence of representations is *semantic* equivalence, which is to say equivalence of the indices (type-level languages) that capture the semantics.
-We will add this clarification.
-
-> line 723: What are transformations?
-
-The rules of differentiation, as transformations of symbolic/grammatical representations.
-
-> line 724: "based *on* the"?
-
-Fixed. Thanks!
-
-## Review D
-
-> Section 6.2 is quite dense, and its motivation is not entirely clear to me.
-Is its purpose to show that existentials can be used to force the dependently-typed definitions developed throughout the paper into the simply-typed algebraic structures?
-If so, I believe a single paragraph would to, and would perhaps be less confusing.
-Furthermore, the section title is confusing: transport properties to what?
-We haven't defined the existentially quantifier variants yet, so it's not yet obvious that we have anywhere to transport properties to.
-
-The purpose is to show that the properties of non-decidable, type-level languages transfer effortlessly to the decidable implementations.
-The importance of this fact is that one gets both (a) easy proofs of useful correctness properties (without complications due to efficiency and even decidability) *and* (b) efficient implementations.
-
-This simple solution for property transfer, however, raises a "technical obstacle" (line 581): the indexed data types do not match the types of standard algebraic properties.
-Existential wrapping is a simple, generic solution to this obstacle.
-
-I renamed the section from "Transporting Properties" to "Transporting Properties from Specification to Implementations"
-
-> Line 86: The clause "including the use of existential quantifiers" is confusing, as it in no way makes clear what the existential quantifiers are used for. I'd suggest dropping the clause entirely.
-
-I've appended "(as in Figure 1) to that sentence".
-
-> Line 500: The types of the primitive language constructs are elided throughout the paper.
-Therefore, it is a bit surprising to see that the variant of star which, ostensibly, abstracts away from lists, still references the list primitives foldr and All, as well as the monoid properties.
-
-Before generalizing, the list plays two different algebraic roles: monoid and foldable.
-Either can be generalized independently.
-
-> Line 608: Why are the existentially-wrapped versions specifically based on the inductive variant?
-Is there anything specific that makes these better or more amenable?
-The text so far led me to believe the coinductive variant was better and more performant.
-
-It was an arbitrary choice. Both implementations (inductive and coinductive) are wrapped alike. We will say so in the text.
-
-> Line 638: Is the pattern matching in the type valid Agda syntax? Again, I don't think these liftings really add anything to the paper.
-
-Yes. All such declarations and definitions in the paper (and those elided from the paper) are fully type- (and thus correctness-)checked when the paper is typeset.
-
-* * * * * * * * * * * * * * * * * * * *
-
-From the final review:
-<blockquote>
-The reviewers want to thank the author for this interesting submission and are happy to accept it with the understanding that the author will make all of the following changes:
-
-1.
-Clarify what's new in the context of proof mechanization or type theory, what's new in the context of parsing, and what's new in the context of formal languages & automata.
-In particular, some contributions do not seem to be novel in the last context.
-The author did not respond to Reviewer C's comment about the work of Lombardy and Sakarovitch, but their work seems to be similar to the derivative operations in the paper and the author should pay attention to it (in addition to other related works).
-
-2.
-Explain the Agda syntax more for the readers who are not familiar with Agda.
-For example, add concrete examples showing how operators in Fig.
-1 apply to concrete regular expressions and words, including the fully expanded types and values (as witnesses of the membership).
-The current draft has only 16 pages so space should not be an issue.
-
-3. Clarify the difference between data and codata: both generally and in terms of difference it makes for main results of this paper.
-
-4.
-Consider explaining how to recover the ordinary setting where a language is a set of words and language union is idempotent.
-For example, explain how the current work can be used to prove the classical theorem for the ordinary setting that the set of derivatives of a regular expression is finite modulo associativity, commutativity and idempotence of union (although the derivative defined here might require more).
-
-5. Fix typos and other issues that have already been discussed in the reviews and the response.
-</blockquote>
-
-*My response:*
-
-1.  I added the following paragraph to the beginning of section 7 (Related Work):
-
-    > The shift from languages as sets of strings to type-level predicates (i.e., proof relevance or ``parsing'') is akin to weighted automata \citep{Schutzenberger1961, DrosteKuske2019} and more generally to semiring-based parsing \citep{Chomsky1959CFL, Goodman1998PIO, Goodman1999SP, Liu2004}, noting that types (``\APT{Set}'' in Agda) form a (commutative) semiring (up to isomorphism).
-    > In particular, \citet{Lombardy2005} investigated Brzozowski-style derivatives is this more general setting, formalizing and generalizing the work of \citet{Antimirov1996}, who decomposed derivatives into simpler components (``partial derivatives'') that sum to the full derivative and lead to efficient recognizers in the form of nondeterministic automata.
-
-    Also an initial sentence to the (now) third paragraph of that section:
-
-    > The main contributions of this paper are to formalization and mechanization of language recognition and parsing.
-
-2.  I added a new figure 2 with examples of languages and membership proofs, along with a new footnote 2 giving some detailed explanations.
-
-3.  I added a new footnote 10.
-
-4.  Appended to the final paragraph of section 7:
-
-    > As mentioned at the end of \secref{Decidability}, one can simplify the development above somewhat by replacing proof isomorphism with the coarser notion of logical equivalence.
-    > Doing so restores union idempotence (modulo equivalence), losing proof relevance and thus replacing parsing by recognition.
-    > One might then also prove that the set of derivatives of a regular language is finite (again, modulo equivalence).
-
-5.  Done!

### 2021-05-31 paper.tex
processed review comments & requirements

@@
 \ifacm
-\documentclass[acmsmall,screen,timestamp,anonymous,review]{acmart}
+\documentclass[acmsmall,screen,timestamp]{acmart}  % ,anonymous,review
 \else
@@
 
-\item The types {\AB A \AD ⊎ \AB B} and {\AB A \ARe × \AB B} represent disjunction and conjunction respectively as non-dependent sum and products.
+\item For types (propositions) \AB A and \AB B, the types {\AB A \AD ⊎ \AB B} and {\AB A \ARe × \AB B} represent disjunction and conjunction respectively as non-dependent sum and products.
 
@@
 Likewise, a proof of {w \AF ∈ \AB{P} \AF ⋆ \AB{Q}}\out{ (language concatenation)} includes the choice of strings \AB{u} and \AB{v}, a proof that {\AB{w} \AD ≡ \AB{u} \AF ⊙ \AB{v}}, and proofs that {\AB{u} \AF ∈ \AB{P}} and {\AB{v} \AF ∈ \AB{Q}}.
+\rnc\source{Examples}
+\figrefdef{language-examples}{Language examples}{\agda{examples}} contains some examples of languages (left) and corresponding membership proofs (right).%
+\footnote{Some explanatory comments:
+\begin{itemize}
 
-The use of type-level predicates makes for simple, direct specification, including the use of existential quantification.
+\item Agda variable names can contain most characters other than spaces and parentheses, e.g., ``\AB{a∪b☆}''.
+
+\item The alphabet type (\AB A in \figref{Lang-ops}) used for these examples is \AF{Char}, i.e., characters.
+
+\item Single-item lists are notated with square brackets.
+
+\item Variables defined but not used can be replaced by the anonymous pattern ``\AK{\_}'' (as in proofs on the right).
+
+\item The \AF{inj₁} and \AF{inj₂} functions are left and right injections into a sum type, corresponding to the left and right introduction forms for logical disjunction.
+
+%% \item The \AF{inj₁} and \AF{inj₂} functions are left and right injections/introductions into a sum/disjunction type.
+
+\item Pairing\out{ (indicated by the infix comma operator)} is right-associative.
+      Since the second and third examples are existential propositions, their inhabiting proofs are dependent pairs.
+      Each comprises a division into substrings and a proof that the substrings concatenate to the given test string (the language's argument), along with language membership proofs for each substring.
+
+\item Equality proofs in these examples are by normalization and hence given by \AIC{refl}.
+\vspace{-2ex}% Avoid bogus blank footnote line spilling to next page
+\end{itemize}}
+
+The use of type-level predicates makes for simple, direct specification, including the use of existential quantification (as in \figref{Lang-ops}).
 As described in \secref{Properties}, it is also fairly easy to prove algebraic properties of predicates\out{, including the semimodule and semiring laws for both predicate semirings}.
@@
   Properties are stated first at the level of predicates, making them easy to understand and prove, thanks to elimination of operational details.
-  These properties are then transported to the computable parsing implementations (\secref{Transporting Properties}) with no additional proofs needed.
+  These properties are then transported to the computable parsing implementations (\secref{Transporting Properties from Specification to Implementations}) with no additional proofs needed.
 \end{itemize}
@@
 
-Since {\AB{u} \AF ⊙ \AIC{[]} \AD ≡ \AB{u}}, it follows that\footnote{Agda variable names can contain most characters other than spaces and parentheses, e.g., ``\AB{ν∘𝒟}'' as declared here.}\footnote{For functions \AB{f} and \AB{g}, {\AB{f} \AF{≗} \AB{g}} is extensional equality, i.e., {\AS ∀ \AB x \AS → \AB{f} \AB{x} \AD{≡} \AB{g} \AB{x}}.}
+Since {\AB{u} \AF ⊙ \AIC{[]} \AD ≡ \AB{u}}, it follows that\footnote{Recall that Agda variable names can contain most characters other than spaces and parentheses, e.g., ``\AB{ν∘𝒟}'' as declared here.}\footnote{For functions \AB{f} and \AB{g}, {\AB{f} \AF{≗} \AB{g}} is extensional equality, i.e., {\AS ∀ \AB x \AS → \AB{f} \AB{x} \AD{≡} \AB{g} \AB{x}}.}
 \agda{ν∘𝒟}
@@
 \agda{foldl}
+\vspace{-5ex}
 }
@@
 
-\item \AF{inj₁} and \AF{inj₂} are left and right injections into a sum type.
+\item Recall that \AF{inj₁} and \AF{inj₂} are left and right injections into a sum type.
 
@@
 \item \AF{\_✶‽} reflects the fact that the empty list exists for every element type (even \AF ⊥).
-
 \end{itemize}
-
+\vspace{-2ex}
 }
@@
 \agda{Decidable}
-With these definitions, we can formulate the problem of decidable language recognition: given a language \AB{P}, construct a term of type {\AF{Decidable} \AB{P}}.
+With these definitions, we can succinctly formulate the problem of decidable language recognition: given a language \AB{P}, construct a term of type {\AF{Decidable} \AB{P}}.
 To accomplish this transformation, we will look for decidable building blocks that mirror the predicate vocabulary defined in \secref{Specifying Languages}.
@@
 The lemmas in \figref{nu-delta-lemmas} tell us how to decompose languages defined in terms of the vocabulary from \figref{Lang-ops}, while the definitions in \figref{compositional-dec} tell us how compute inhabitation of the resulting types, resulting in decidable parsing.
-These lemmas and definitions cannot be applied \emph{automatically} in their present form, however, because languages are functions, as are \AF{ν} and \AF{δ}, and so are not subject to pattern-matching.
+These lemmas and definitions cannot be applied \emph{automatically} in their present form, however, because languages are functions, as are \AF{ν} and \AF{δ}, and so are not subject to structural inspection.
 An automatic solution would need some form of \emph{reflection} of language-constructing operations or of \AF{ν} and \AF{δ} as inspectable data.
@@
 
-\secreftwo{Symbolic Differentiation}{Automatic Differentiation} applies the symbolic and automatic strategies respectively to languages.
+\secreftwo{Symbolic Differentiation}{Automatic Differentiation} apply the symbolic and automatic strategies respectively to languages.
 Both algorithms are corollaries of the lemmas in \figref{nu-delta-lemmas} and are automatically proved correct by type checking.
@@
 The ``symbolic differentiation'' style reflects language-building vocabulary (\figref{Lang-ops}) into an inductive data type (of modestly extended regular expressions), while \AF{ν} and \AF{δ} become functions defined by pattern-matching on that data type.
-For convenience, we can use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the module prefix ``{◇.}\hspace{0.05em}''.
-The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\api} along with a function {\AF{⟦\_⟧}} that converts a symbolic language representation to a computable parser.
+For convenience, use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the module prefix ``{◇.}\hspace{0.05em}''.
+The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\api} along with a trivial semantic function {\AF{⟦\_⟧}} (mapping any representation of type {\AF{Lang} \AB P} to the corresponding type-level language \AB{P}) and a less trivial, computable function  {\AF{⟦\_⟧‽}} that converts a symbolic language representation to a computable parser.
+Note that the \AF{\_◂\_} operation from \secref{Decidability} has been reified as a new constructor alongside the more conventional language operators.
+The reason \AF{\_◂\_} must be part of the inductive representation is the same as the other constructors, namely so that the core lemmas (\figref{nu-delta-lemmas}) translate into an implementation in terms of that representation.
 
-Decidable parsing relies only on \AF{ν} and \AF{δ} (via {\AF{⟦\_⟧}}), which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\out{\notefoot{Describe this derivation.}}
+Decidable parsing relies only on \AF{ν} and \AF{δ} (via {\AF{⟦\_⟧‽}}), which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\out{\notefoot{Describe this derivation.}}
 These clauses are meant to be read in ``column-major'' order, i.e., each column is a complete definition.
 Correctness is guaranteed by the types of \AF{ν} and \AF{δ} and so is proved automatically by type-checking \figref{symbolic-defs}.
-(We could thus use \emph{any} definitions of \AF{ν} and \AF{δ} that type- and termination-check, although not many definitions would do so.)
+(Recall from \secref{Decidability} that correctness is defined as constructing a term of type {\AF{Decidable} \AF{P}} for a given language \AB{P}, as satisfied by the type of {\AF{⟦\_⟧‽}}.)
+We could thus use \emph{any} definitions of \AF{ν} and \AF{δ} that type- and termination-check, although not many definitions would do so.
 
@@
 Now consider the dual strategy of ``automatic differentiation''.
-This time, reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type of tries\footnote{The classic trie (``prefix tree'') data structure was introduced by \citet{Thue1912Gegenseitige} to represent sets of strings \citep[Section 6.3]{Knuth1998ACP3}, later generalized to arbitrary non-nested algebraic data types \citep{Connelly1995GenTrie}, and then from sets to functions \citep{Hinze2000GGT}.}, while redefining the language-building vocabulary as functions on that data type defined by \emph{copatterns} \citep{AbelPientka2016}.
+This time, reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type of tries\footnote{The classic trie (``prefix tree'') data structure was introduced by \citet{Thue1912Gegenseitige} to represent sets of strings \citep[Section 6.3]{Knuth1998ACP3}, later generalized to arbitrary non-nested algebraic data types \citep{Connelly1995GenTrie}, and then from sets to functions \citep{Hinze2000GGT}.}, while redefining the language-building vocabulary as functions on that data type defined by \emph{copatterns} \citep{AbelPientka2016}.\footnote{Inductive types (``data'') describe finitely large values and are processed recursively (as least fixed points) by pattern-matching clauses that decompose arguments.
+Dually, coinductive types (``codata'') describe infinitely large values and are processed recursively (as greatest fixed points) by copattern-matching clauses that compose results.
+Programs on inductive types are often proved by induction, while programs on coinductive types are often proved by coinduction\out{ and specifically bisumulation}.
+(As we will see in \secref{Transporting Properties from Specification to Implementations}, however, the simple relationship to the specification in \secref{Decomposing Languages} allows many trivial correctness proofs, needing neither introduction or coinduction.)
+See \citet{Gordon1995:coinduction} for a tutorial on theory and techniques of coinduction.
+
+Tries in general represent functions, with each trie datum corresponding to domain value.
+Even when each domain value is finitely large, there are often infinitely many of them (e.g., lists), so tries will naturally be infinite and thus more amenable to coinductive than inductive analysis.}
 (This program duality pattern has been noted and investigated by \citet{OstermannJabs2018}.)
@@
 
-The correctness proof is \emph{almost} accomplished by type- and termination-checking, but a technical problem arises.
+The proof of correctness (defined by {\AF{⟦\_⟧}} and {\AF{⟦\_⟧‽}}, defined as before) is \emph{almost} accomplished by type- and termination-checking, but a technical problem arises.
 The {\AF{δ} (\AB{p} \AF ⋆ \AB{q})} clause does not satisfy Agda's termination checker, which cannot see that the argument {\AF{δ} \AB{p} \AB{a}} in the recursive use of \AF{\_⋆\_} is in some sense smaller than \AB{p}.
@@
 \rnc\source{SizedAutomatic}
-This solution only requires giving \AF{Lang} (now tries) an index {\AB{i} \AK : \APT{Size}} corresponding to the maximum depth to which a trie can be searched, or equivalently the longest string that can be matched\out{ (via {\AF{⟦\_⟧}})}.
-In practice, we will work with arbitrarily deep tries, i.e., ones having index \APo{∞}, as in the type of {\AF{⟦\_⟧}}.
+This solution only requires giving \AF{Lang} (now tries) an index {\AB{i} \AK : \APT{Size}} corresponding to the maximum depth to which a trie can be searched, or equivalently the longest string that can be matched\out{ (via {\AF{⟦\_⟧‽}})}.
+In practice, we will work with arbitrarily deep tries, i.e., ones having index \APo{∞}, as in the type of {\AF{⟦\_⟧‽}}.
 The modified representation is shown in \figrefdef{sized-automatic-api}{Sized tries (coinductive)}{\api}.
@@
 
-To more clearly see the parallel with automatic differentiation (AD), note that AD implementations sample a function \emph{and} its derivative together to exploit the fact that these two computations typically share much common work.
+To see the parallel with automatic differentiation (AD) more clearly, note that AD implementations sample a function \emph{and} its derivative together to exploit the fact that these two computations typically share much common work.
 This work sharing also applies when differentiating languages.
@@
 \out{We will explore this generality in \secref{Beyond Predicates}.}
-When the domain monoid commutes (multiplicatively)---e.g., for functions of space and time rather than languages---the monoid semiring does as well.
+When the domain monoid commutes (multiplicatively)---e.g., for functions of space and time\out{ rather than languages}---the monoid semiring does as well.
 Applications include power series (univariate and multivariate) and image processing \citep{Elliott2019-convolution}.
@@
 
-\subsectionl{Transporting Properties}
+\subsectionl{Transporting Properties from Specification to Implementations}
 
@@
 \end{code}
-Referring to the type and operations of \figreftwo{Lang-ops}{symbolic-api} via the module prefixes ``{◇.}\hspace{0.05em}'' and ``{□.}\hspace{0.05em}'' respectively, define \agda{wrapped-Lang}
+Referring to the type and operations of \figreftwo{Lang-ops}{symbolic-api} (for the inductive language representation) via the module prefixes ``{◇.}\hspace{0.05em}'' and ``{□.}\hspace{0.05em}'' respectively, define \agda{wrapped-Lang}
 Then wrap union as follows:
 \agda{wrapped-union}
+Likewise for the coinductive language representation in \figref{sized-automatic-api}.
 The type-level language components of the arguments and result are given explicitly here but could instead be inferred by the compiler.
@@
 A crucial question is the choice of equivalence relation for the new properties.
-If we choose equivalence on these dependent pairs to be equivalence of \emph{first} projections (type \emph{indices}), then all algebraic properties of those indices will hold for the existentially wrapped versions as well.
+If we choose equivalence on these dependent pairs to be \emph{semantic}---i.e., equivalence of \emph{first} projections (type \emph{indices})---then all algebraic properties of those indices will hold for the existentially wrapped versions as well.
 This choice may at first seem like cheating, since the entire operational aspect of the representation is ignored.
 The correctness of that aspect, however, is guaranteed by its dependent typing, which anchors its meaning (no matter how cleverly implemented) to the non-operational type index.
-For the language implementations of \secref{From Languages to Parsing}, this anchoring is guaranteed by the type of \AF{⟦\_⟧} in \figreftwo{symbolic-api}{sized-automatic-api}.
+For the language implementations of \secref{From Languages to Parsing}, this anchoring is guaranteed by the types of \AF{⟦\_⟧} and \AF{⟦\_⟧‽} in \figreftwo{symbolic-api}{sized-automatic-api}.
 (Similarly, for the decidable type constructors in \figref{compositional-dec}, correctness is guaranteed by the types of the \AIC{yes} and \AIC{no} constructors in the definition of ‌\ARe{Dec} in \secref{Decidability}.)
@@
 
+The shift from languages as sets of strings to type-level predicates (i.e., proof relevance or ``parsing'') is akin to weighted automata \citep{Schutzenberger1961, DrosteKuske2019} and more generally to semiring-based parsing \citep{Chomsky1959CFL, Goodman1998PIO, Goodman1999SP, Liu2004}, noting that types (``\APT{Set}'' in Agda) form a (commutative) semiring (up to isomorphism).
+In particular, \citet{Lombardy2005} investigated Brzozowski-style derivatives is this more general setting, formalizing and generalizing the work of \citet{Antimirov1996}, who decomposed derivatives into simpler components (``partial derivatives'') that sum to the full derivative and lead to efficient recognizers in the form of nondeterministic automata.
+
 Most work on formal languages in functional programming has been in parsing, particularly via parsing combinators \citep{Hutton1996monadic, Leijen2001parsec, Swierstra2008combinator}.
@@
 
-There has also been some work using type theory to capture languages as type-level predicates much as in \secref{Specifying Languages}:
+The main contributions of this paper are to formalization and mechanization of language recognition and parsing.
+There has been some work using type theory to capture languages as type-level predicates much as in \secref{Specifying Languages}:
 \begin{itemize}
@@
 \citet{Brzozowski64} also made clear mention of the underlying meaning on sets of strings in his original work on regular expression derivatives.
-Moreover, none of these previous investigations appear to have addressed proof isomorphism (distinguishing multiple parsings/explanations), and correspondingly all use transformations based the less precise laws originally discovered by \citet{Brzozowski64}.
+Moreover, none of these previous investigations appear to have addressed proof isomorphism (distinguishing multiple parsings/explanations), and correspondingly all use transformations based on the less precise laws originally discovered by \citet{Brzozowski64}.
 Those laws were based on the assumption that union (``addition'' in both language semirings of \secref{Predicate Algebra}) is idempotent, which is not true for type-level language predicates related by proof isomorphism (rather than mere logical equivalence).
@@
 The difference is mainly visible in the \AF ν and \AF δ lemmas for {\AB P ⋆ \AB Q} and {\AB P ☆} in \figref{nu-delta-lemmas}.
+As mentioned at the end of \secref{Decidability}, one can simplify the development above somewhat by replacing proof isomorphism with the coarser notion of logical equivalence.
+Doing so restores union idempotence (modulo equivalence), losing proof relevance and thus replacing parsing by recognition.
+One might then also prove that the set of derivatives of a regular language is finite (again, modulo equivalence).
 

### 2021-06-02 response-to-reviews.md
chnages in respose to reviews

@@
 ***Note:** I say "we" instead of "I" in these responses, because the blind review process was still in effect when I wrote them.*
-
-> The authors’ response should address reviewer concerns, answer reviewer questions, and correct misunderstandings.
-Reviewers are not required to read past the 500 word limit, so please focus your response there.
-You may provide additional (e.g.
-point-by-point, per-reviewer) response beyond the limit if it may help the reviewers.
-
-***Oops. I read this clarification late in the game. To do: reorganize the following remarks accordingly.***
 
@@
 
-*My response:*
+My response:
 
@@
 
-    > The shift from languages as sets of strings to type-level predicates (i.e., proof relevance or ``parsing'') is akin to weighted automata \citep{Schutzenberger1961, DrosteKuske2019} and more generally to semiring-based parsing \citep{Chomsky1959CFL, Goodman1998PIO, Goodman1999SP, Liu2004}, noting that types (``\APT{Set}'' in Agda) form a (commutative) semiring (up to isomorphism).
-    > In particular, \citet{Lombardy2005} investigated Brzozowski-style derivatives is this more general setting, formalizing and generalizing the work of \citet{Antimirov1996}, who decomposed derivatives into simpler components (``partial derivatives'') that sum to the full derivative and lead to efficient recognizers in the form of nondeterministic automata.
+    > The shift from languages as sets of strings to type-level predicates (i.e., proof relevance or “parsing”) is akin to weighted automata [Schützenberger 1961; Droste and Kuske 2019] and more generally to semiring-based parsing [Chomsky and Schützenberger 1959; Goodman 1998, 1999; Liu 2004], noting that types (“Set” in Agda) form a (commutative) semiring (up to isomorphism).
+In particular, Lombardy and Sakarovitch [2005] investigated Brzozowski-style derivatives in this more general setting, formalizing and generalizing the work of Antimirov [1996], who decomposed derivatives into simpler components (“partial derivatives”) that sum to the full derivative and lead to efficient recognizers in the form of nondeterministic automata.
 
@@
 
-3.  I added a new footnote 10.
+3.  I added a new footnote 10:
+
+    > Inductive types (“data”) describe finitely large values and are processed recursively (as least fixed points) by pattern- matching clauses that decompose arguments.
+Dually, coinductive types (“codata”) describe infinitely large values and are processed corecursively (as greatest fixed points) by copattern-matching clauses that compose results.
+Programs on inductive types are often proved by induction, while programs on coinductive types are often proved by coinduction.
+(As we will see in Section 6.2, however, the simple relationship to the specification in Section 2 allows many trivial correctness proofs, needing neither induction or coinduction.) See Gordon [1995] for a tutorial on theory and techniques of coinduction.
+    > 
+    > Tries in general represent functions, with each trie datum corresponding to a domain value.
+Even when each domain value is finitely large, there are often infinitely many of them (e.g., lists), so tries will naturally be infinite and thus more amenable to coinductive than inductive analysis.
 
@@
 
-    > As mentioned at the end of \secref{Decidability}, one can simplify the development above somewhat by replacing proof isomorphism with the coarser notion of logical equivalence.
-    > Doing so restores union idempotence (modulo equivalence), losing proof relevance and thus replacing parsing by recognition.
-    > One might then also prove that the set of derivatives of a regular language is finite (again, modulo equivalence).
+    > As mentioned at the end of Section 3, one can simplify the development above somewhat by replacing proof isomorphism with the coarser notion of logical equivalence.
+Doing so restores union idempotence (modulo equivalence), losing proof relevance and thus re- placing parsing by recognition.
+One might then also prove that the set of derivatives of a regular language is finite (again, modulo equivalence).
 

### 2021-06-02 paper.tex
chnages in respose to reviews

@@
 For convenience, use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the module prefix ``{◇.}\hspace{0.05em}''.
-The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\api} along with a trivial semantic function {\AF{⟦\_⟧}} (mapping any representation of type {\AF{Lang} \AB P} to the corresponding type-level language \AB{P}) and a less trivial, computable function  {\AF{⟦\_⟧‽}} that converts a symbolic language representation to a computable parser.
+The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\api} along with a trivially defined semantic function {\AF{⟦\_⟧}} (mapping any representation of type {\AF{Lang} \AB P} to the corresponding type-level language \AB{P}) and a less trivial, computable function  {\AF{⟦\_⟧‽}} that converts a symbolic language representation to a computable parser.
 Note that the \AF{\_◂\_} operation from \secref{Decidability} has been reified as a new constructor alongside the more conventional language operators.
@@
 This time, reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type of tries\footnote{The classic trie (``prefix tree'') data structure was introduced by \citet{Thue1912Gegenseitige} to represent sets of strings \citep[Section 6.3]{Knuth1998ACP3}, later generalized to arbitrary non-nested algebraic data types \citep{Connelly1995GenTrie}, and then from sets to functions \citep{Hinze2000GGT}.}, while redefining the language-building vocabulary as functions on that data type defined by \emph{copatterns} \citep{AbelPientka2016}.\footnote{Inductive types (``data'') describe finitely large values and are processed recursively (as least fixed points) by pattern-matching clauses that decompose arguments.
-Dually, coinductive types (``codata'') describe infinitely large values and are processed recursively (as greatest fixed points) by copattern-matching clauses that compose results.
+Dually, coinductive types (``codata'') describe infinitely large values and are processed corecursively (as greatest fixed points) by copattern-matching clauses that compose results.
 Programs on inductive types are often proved by induction, while programs on coinductive types are often proved by coinduction\out{ and specifically bisumulation}.
-(As we will see in \secref{Transporting Properties from Specification to Implementations}, however, the simple relationship to the specification in \secref{Decomposing Languages} allows many trivial correctness proofs, needing neither introduction or coinduction.)
+(As we will see in \secref{Transporting Properties from Specification to Implementations}, however, the simple relationship to the specification in \secref{Decomposing Languages} allows many trivial correctness proofs, needing neither induction or coinduction.)
 See \citet{Gordon1995:coinduction} for a tutorial on theory and techniques of coinduction.
 
-Tries in general represent functions, with each trie datum corresponding to domain value.
+Tries in general represent functions, with each trie datum corresponding to a domain value.
 Even when each domain value is finitely large, there are often infinitely many of them (e.g., lists), so tries will naturally be infinite and thus more amenable to coinductive than inductive analysis.}
@@
 The shift from languages as sets of strings to type-level predicates (i.e., proof relevance or ``parsing'') is akin to weighted automata \citep{Schutzenberger1961, DrosteKuske2019} and more generally to semiring-based parsing \citep{Chomsky1959CFL, Goodman1998PIO, Goodman1999SP, Liu2004}, noting that types (``\APT{Set}'' in Agda) form a (commutative) semiring (up to isomorphism).
-In particular, \citet{Lombardy2005} investigated Brzozowski-style derivatives is this more general setting, formalizing and generalizing the work of \citet{Antimirov1996}, who decomposed derivatives into simpler components (``partial derivatives'') that sum to the full derivative and lead to efficient recognizers in the form of nondeterministic automata.
+In particular, \citet{Lombardy2005} investigated Brzozowski-style derivatives in this more general setting, formalizing and generalizing the work of \citet{Antimirov1996}, who decomposed derivatives into simpler components (``partial derivatives'') that sum to the full derivative and lead to efficient recognizers in the form of nondeterministic automata.
 

### 2021-06-03 paper.tex
misc near-final tweaks

@@
 \end{itemize}
-Although many proofs are elided below, the source code for this paper is freely available and contains fully verified Agda code with no postulates.%
-\ifacm\else
-%% \citep{Elliott2021-derivatives-repo}.
-%% (\repoURL).
-\footnote{\protect\repoURL}
-\fi{}
-All type signatures and definitions in the paper are generated by the Agda compiler.
+Although many proofs are elided below, the source code for this paper is freely available\footnote{\protect{\repoURL}} and contains fully verified Agda code with no postulates.
+The type signatures and definitions displayed in the paper are extracted from this source code by the Agda compiler.
 
@@
 The ``symbolic differentiation'' style reflects language-building vocabulary (\figref{Lang-ops}) into an inductive data type (of modestly extended regular expressions), while \AF{ν} and \AF{δ} become functions defined by pattern-matching on that data type.
-For convenience, use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the module prefix ``{◇.}\hspace{0.05em}''.
-The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\api} along with a trivially defined semantic function {\AF{⟦\_⟧}} (mapping any representation of type {\AF{Lang} \AB P} to the corresponding type-level language \AB{P}) and a less trivial, computable function  {\AF{⟦\_⟧‽}} that converts a symbolic language representation to a computable parser.
+For convenience, use the same names as in \secref{Specifying Languages} for these new counterparts, and refer to the original versions via the module prefix ``{◇.}\hspace{0.05em}''.
+The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\api} along with a computable function {\AF{⟦\_⟧‽}} that converts a symbolic language representation to a computable parser.
 Note that the \AF{\_◂\_} operation from \secref{Decidability} has been reified as a new constructor alongside the more conventional language operators.
@@
 
+The meaning of any {\AB r \AK{:} \AF{Lang} \AB P} is defined simply as \AB{P}, ignoring the representation \AB{r} itself\out{ (beyond its type)}:
+\agda{semantics}
+
 Decidable parsing relies only on \AF{ν} and \AF{δ} (via {\AF{⟦\_⟧‽}}), which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\out{\notefoot{Describe this derivation.}}
-These clauses are meant to be read in ``column-major'' order, i.e., each column is a complete definition.
+These clauses are meant to be read in column-major order, i.e., each column is a complete definition.
 Correctness is guaranteed by the types of \AF{ν} and \AF{δ} and so is proved automatically by type-checking \figref{symbolic-defs}.
@@
 Decidable recognition again relies only on \AF{ν} and \AF{δ}, which are defined for each language operation in \figrefdef{automatic-defs}{Automatic differentiation (row-major/copatterns)}{\agda{defs}} (though with a problem to be noted and fixed in the next paragraph).
-These clauses are meant to be read in ``row-major'' order, i.e., each row is a complete definition.
+These clauses are meant to be read in row-major order, i.e., each row is a complete definition.
 Note that the clauses in \figref{automatic-defs} are syntactically identical to those in \figref{symbolic-defs} but are organized dually.

### 2021-06-03 paper.tex
copattern reference

@@
 
-The meaning of any {\AB r \AK{:} \AF{Lang} \AB P} is defined simply as \AB{P}, ignoring the representation \AB{r} itself\out{ (beyond its type)}:
+The \emph{meaning} of any {\AB r \AK{:} \AF{Lang} \AB P} is defined simply as \AB{P}, ignoring the representation \AB{r} itself\out{ (beyond its type)}:
 \agda{semantics}
@@
 Decidable parsing relies only on \AF{ν} and \AF{δ} (via {\AF{⟦\_⟧‽}}), which are defined in \figrefdef{symbolic-defs}{Symbolic differentiation (column-major/patterns)}{\agda{defs}}, as systematically derived from the lemmas of \figref{nu-delta-lemmas}.\out{\notefoot{Describe this derivation.}}
-These clauses are meant to be read in column-major order, i.e., each column is a complete definition.
+These clauses are meant to be read in column-major order, i.e., each column is a complete definition (two definitions of ten pattern-matching clauses each).
 Correctness is guaranteed by the types of \AF{ν} and \AF{δ} and so is proved automatically by type-checking \figref{symbolic-defs}.
@@
 Now consider the dual strategy of ``automatic differentiation''.
-This time, reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type of tries\footnote{The classic trie (``prefix tree'') data structure was introduced by \citet{Thue1912Gegenseitige} to represent sets of strings \citep[Section 6.3]{Knuth1998ACP3}, later generalized to arbitrary non-nested algebraic data types \citep{Connelly1995GenTrie}, and then from sets to functions \citep{Hinze2000GGT}.}, while redefining the language-building vocabulary as functions on that data type defined by \emph{copatterns} \citep{AbelPientka2016}.\footnote{Inductive types (``data'') describe finitely large values and are processed recursively (as least fixed points) by pattern-matching clauses that decompose arguments.
+This time, reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type of tries\footnote{The classic trie (``prefix tree'') data structure was introduced by \citet{Thue1912Gegenseitige} to represent sets of strings \citep[Section 6.3]{Knuth1998ACP3}, later generalized to arbitrary non-nested algebraic data types \citep{Connelly1995GenTrie}, and then from sets to functions \citep{Hinze2000GGT}.}, while redefining the language-building vocabulary as functions on that data type defined by \emph{copatterns} \citep{AbelEtAl12013, AbelPientka2016}.\footnote{Inductive types (``data'') describe finitely large values and are processed recursively (as least fixed points) by pattern-matching clauses that decompose arguments.
 Dually, coinductive types (``codata'') describe infinitely large values and are processed corecursively (as greatest fixed points) by copattern-matching clauses that compose results.
@@
 
-Tries in general represent functions, with each trie datum corresponding to a domain value.
+Tries in general represent functions, with each trie datum position corresponding to a domain value.
 Even when each domain value is finitely large, there are often infinitely many of them (e.g., lists), so tries will naturally be infinite and thus more amenable to coinductive than inductive analysis.}
@@
 Decidable recognition again relies only on \AF{ν} and \AF{δ}, which are defined for each language operation in \figrefdef{automatic-defs}{Automatic differentiation (row-major/copatterns)}{\agda{defs}} (though with a problem to be noted and fixed in the next paragraph).
-These clauses are meant to be read in row-major order, i.e., each row is a complete definition.
+These clauses are meant to be read in row-major order, i.e., each row is a complete definition (ten definitions of two copattern-matching clauses each).
 Note that the clauses in \figref{automatic-defs} are syntactically identical to those in \figref{symbolic-defs} but are organized dually.

### 2021-07-02 paper.tex
misc

@@
 \author{Conal Elliott}
+\affiliation[obeypunctuation=true]{%
+%% \institution{Independent}
+\country{USA}}
+\email{conal@conal.net}
 
@@
 \end{itemize}
-Although many proofs are elided below, the source code for this paper is freely available\footnote{\protect{\repoURL}} and contains fully verified Agda code with no postulates.
+Although many proofs are elided below, the source code for this paper is freely available and contains fully verified Agda code with no postulates \citep{Elliott2021-language-derivatives-repo}.
 The type signatures and definitions displayed in the paper are extracted from this source code by the Agda compiler.
@@
 
-Conveniently, booleans and types form commutative semirings with all necessary proofs already in the standard library.\footnote{The equivalence relation used for types is isomorphism rather than equality.
+Conveniently, booleans and types form commutative semirings with all necessary proofs already in Agda's standard library.\footnote{The equivalence relation used for types is isomorphism rather than equality.
 The boolean semiring is idempotent, while the type semiring is not (in order to distinguish multiple parsings in ambiguous languages).
@@
 The two needed distributive properties hold for \emph{any} domain-transforming operation, whether associative or not.
-The proofs of these properties, while reasonably straightforward, are beyond the scope of this paper but are available in the paper's Agda source code.\out{\notefoot{There will be short and long forms of this paper.
-Perhaps the long form will contain narrated parts of the monoid semiring proof.}}
+The formal statements and proofs of these properties are included in this paper's Agda source code \citep{Elliott2021-language-derivatives-repo}.
 
@@
 Algebraic abstractions and their properties are simply typed.
-For instance, each instance of the \ARe{Monoid} abstraction involves a single type \AB τ with a binary operation {\AB{\_∙\_} \AK : \AB τ \AK → \AB τ \AK → \AB τ} and identity value {\AB ε \AK : \AB τ}.
-This recipe accommodates the language operations in \figreftwo{Lang-ops}{lang-via-pred} and their generalizations in \figref{predicates} but not the dependently typed, \emph{indexed} versions of languages and their operations in \figreftwo{symbolic-api}{sized-automatic-api}, nor decidable types and their vocabulary in \figref{compositional-dec}.
+For example, each instance of the \ARe{Monoid} abstraction involves a single type \AB τ with a binary operation {\AB{\_∙\_} \AK : \AB τ \AK → \AB τ \AK → \AB τ} and identity value {\AB ε \AK : \AB τ}.
+This recipe accommodates the language operations in \figreftwo{Lang-ops}{lang-via-pred} and their generalizations in \figref{predicates}, but not the dependently typed, \emph{indexed} versions of languages and their operations in \figreftwo{symbolic-api}{sized-automatic-api}, nor decidable types and their vocabulary in \figref{compositional-dec}.
 

### 2021-07-03 paper.tex
classifications, keywords, copyright

@@
 
+\setcopyright{rightsretained}
+\acmJournal{PACMPL}
+\acmYear{2021} \acmVolume{5} \acmNumber{ICFP} \acmArticle{78} \acmMonth{8} \acmPrice{}\acmDOI{10.1145/3473583}
+
 \begin{document}
@@
 \end{abstract}
+
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10003752.10003790.10011740</concept_id>
+<concept_desc>Theory of computation~Type theory</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept>
+<concept_id>10003752.10003766.10003776</concept_id>
+<concept_desc>Theory of computation~Regular languages</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10003752.10010124.10010138.10010142</concept_id>
+<concept_desc>Theory of computation~Program verification</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10003752.10010124.10010138.10010145</concept_id>
+<concept_desc>Theory of computation~Parsing</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Theory of computation~Type theory}
+\ccsdesc[500]{Theory of computation~Regular languages}
+\ccsdesc[500]{Theory of computation~Program verification}
+\ccsdesc[300]{Theory of computation~Parsing}
+
+\keywords{program calculation; language derivatives; tries}
 

### 2021-07-03 paper.tex
move explanatory footnote 2 so as not to break across pages
footnote placement; page breaks; proof

@@
 \rnc\source{Examples}
-\figrefdef{language-examples}{Language examples}{\agda{examples}} contains some examples of languages (left) and corresponding membership proofs (right).%
-\footnote{Some explanatory comments:
-\begin{itemize}
+\figrefdef{language-examples}{Language examples\protect\footnotemark}{\agda{examples}} contains some examples of languages (left) and corresponding membership proofs (right).
 
-\item Agda variable names can contain most characters other than spaces and parentheses, e.g., ``\AB{a∪b☆}''.
-
-\item The alphabet type (\AB A in \figref{Lang-ops}) used for these examples is \AF{Char}, i.e., characters.
-
-\item Single-item lists are notated with square brackets.
-
-\item Variables defined but not used can be replaced by the anonymous pattern ``\AK{\_}'' (as in proofs on the right).
-
-\item The \AF{inj₁} and \AF{inj₂} functions are left and right injections into a sum type, corresponding to the left and right introduction forms for logical disjunction.
-
-%% \item The \AF{inj₁} and \AF{inj₂} functions are left and right injections/introductions into a sum/disjunction type.
-
-\item Pairing\out{ (indicated by the infix comma operator)} is right-associative.
-      Since the second and third examples are existential propositions, their inhabiting proofs are dependent pairs.
-      Each comprises a division into substrings and a proof that the substrings concatenate to the given test string (the language's argument), along with language membership proofs for each substring.
-
-\item Equality proofs in these examples are by normalization and hence given by \AIC{refl}.
-\vspace{-2ex}% Avoid bogus blank footnote line spilling to next page
-\end{itemize}}
+%% I split the footnote mark & text and delayed the latter so the text would
+%% go on the same page as the figure and stay on one page.
 
@@
 The type signatures and definitions displayed in the paper are extracted from this source code by the Agda compiler.
+
+\footnotetext{Some explanatory comments:
+\begin{itemize}
+
+\item Agda variable names can contain most characters other than spaces and parentheses, e.g., ``\AB{a∪b☆}''.
+
+\item The alphabet type (\AB A in \figref{Lang-ops}) used for these examples is \AF{Char}, i.e., characters.
+
+\item Single-item lists are notated with square brackets.
+
+\item Variables defined but not used can be replaced by the anonymous pattern ``\AK{\_}'' (as in proofs on the right).
+
+\item The \AF{inj₁} and \AF{inj₂} functions are left and right injections into a sum type, corresponding to the left and right introduction forms for logical disjunction.
+
+%% \item The \AF{inj₁} and \AF{inj₂} functions are left and right injections/introductions into a sum/disjunction type.
+
+\item Pairing\out{ (indicated by the infix comma operator)} is right-associative.
+      Since the second and third examples are existential propositions, their inhabiting proofs are dependent pairs.
+      Each comprises a division into substrings and a proof that the substrings concatenate to the given test string (the language's argument), along with language membership proofs for each substring.
+
+\item Equality proofs in these examples are by normalization and hence given by \AIC{refl}.
+\vspace{-2ex}% Avoid bogus blank footnote line spilling to next page
+\end{itemize}}
 
@@
 \agda{Dec}
-For compositionality, we will use a few operations that lift decidability of types to decidability of constructions on those types, as shown in \figrefdef{compositional-dec}{Compositional decidability}{\agda{compositional-dec}}.\footnote{%
+For compositionality, we will use a few operations that lift decidability of types to decidability of constructions on those types, as shown in \figrefdef{compositional-dec}{Compositional decidability\protect\footnotemark}{\agda{compositional-dec}}.
+Moreover, decidability lifts naturally from types to predicates:
+\agda{Decidable}
+
+With these definitions, we can succinctly formulate the problem of decidable language recognition: \emph{given a language \AB{P}, construct a term of type {\AF{Decidable} \AB{P}}.}
+To accomplish this transformation, we will look for decidable building blocks that mirror the predicate vocabulary defined in \secref{Specifying Languages}.
+
+\footnotetext{%
 A few explanatory remarks:
@@
 }
-Moreover, decidability lifts naturally from types to predicates:
-\agda{Decidable}
-With these definitions, we can succinctly formulate the problem of decidable language recognition: given a language \AB{P}, construct a term of type {\AF{Decidable} \AB{P}}.
-To accomplish this transformation, we will look for decidable building blocks that mirror the predicate vocabulary defined in \secref{Specifying Languages}.
 
@@
 In practice, we will work with arbitrarily deep tries, i.e., ones having index \APo{∞}, as in the type of {\AF{⟦\_⟧‽}}.
+
 The modified representation is shown in \figrefdef{sized-automatic-api}{Sized tries (coinductive)}{\api}.
@@
 This time the compiler successfully proves \emph{total} correctness (including termination).
+
+\vspace{3ex}
 
@@
 \agda{𝒟′}
-The potential work-sharing version:
+\begin{samepage}
+The potential work-sharing version exploits the close relationship between \AB{f} and {\AF{𝒟} \AB{f}}:
 \agda{ʻ𝒟}
-Equivalence follows from lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} of \secref{Decomposing Languages}:
+\end{samepage}
+Equivalence follows from a simple inductive argument, thanks to lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} of \secref{Decomposing Languages}:
 \agda{𝒟′≡ʻ𝒟}
@@
 
+\begin{samepage}
 \sectionl{Properties}
@@
 \subsectionl{Predicate Algebra}
+\end{samepage}
 
@@
 
+\vspace{2.5ex}
 \subsectionl{Transporting Properties from Specification to Implementations}

### 2021-08-16 paper.tex
zip
misc tweaks for talk

@@
 \ifacm
-\documentclass[acmsmall,screen,timestamp]{acmart}  % ,anonymous,review
+\documentclass[acmsmall,screen]{acmart}
 \else
@@
 \fi
-\authorsaddresses{}
 
@@
 
-\RequirePackage{newunicodechar, stmaryrd, setspace, comment, scalerel}
+\RequirePackage{newunicodechar, stmaryrd, setspace, comment, scalerel, doi}
 
@@
 \input{macros}
+
+%% Use oadoi.org instead of doi.org. Doesn't seem to take.
+\makeatletter
+\def\@formatdoi#1{\url{https://oadoi.org/#1}}
+\makeatother
 
@@
 
+%% \setcopyright{rightsretained}
+%% \acmJournal{PACMPL}
+%% \acmYear{2021} \acmVolume{5} \acmNumber{ICFP} \acmArticle{78} \acmMonth{8} \acmPrice{}\acmDOI{10.1145/3473583}
+
+%%% The following is specific to ICFP '21 and the paper
+%%% 'Symbolic and Automatic Differentiation of Languages'
+%%% by Conal Elliott.
+%%%
 \setcopyright{rightsretained}
+\acmPrice{}
+\acmDOI{10.1145/3473583}
+\acmYear{2021}
+\copyrightyear{2021}
+\acmSubmissionID{icfp21main-p87-p}
 \acmJournal{PACMPL}
-\acmYear{2021} \acmVolume{5} \acmNumber{ICFP} \acmArticle{78} \acmMonth{8} \acmPrice{}\acmDOI{10.1145/3473583}
+\acmVolume{5}
+\acmNumber{ICFP}
+\acmArticle{78}
+\acmMonth{8}
 
@@
 </concept>
-<concept>
 <concept>
@@
 \end{itemize}
-Although many proofs are elided below, the source code for this paper is freely available and contains fully verified Agda code with no postulates \citep{Elliott2021-language-derivatives-repo}.
-The type signatures and definitions displayed in the paper are extracted from this source code by the Agda compiler.
-
 \footnotetext{Some explanatory comments:
@@
 \end{itemize}}
+Although many proofs are elided below, the source code for this paper is freely available and contains fully verified Agda code with no postulates \citep{Elliott2021-language-derivatives-repo}.
+The type signatures and definitions displayed in the paper are extracted from this source code by the Agda compiler.
 
@@
 
+%% The publisher wants URLs in the url field, while I prefer them in the title.
+\ifacm
+\bibliography{acm-bib}
+\else
 \bibliography{bib}
+\fi
 

### 2021-08-16 final-review-responses.md
talk and final responses

--- 2021-06-02 response-to-reviews.md
@@
-# Paper #87 review responses
-
-***Note:** I say "we" instead of "I" in these responses, because the blind review process was still in effect when I wrote them.*
-
-## Review A
-
-> My major concern is novelty, but unfortunately I am not familiar with the prior work.
-Given that I was a bit surprised by the absence of prior work, and that there were already so many papers on related topics, my rating is B.
-I might advocate for this paper more enthusiastically once novelty is confirmed.
-
-We were surprised as well not to find formalized & mechanized connections between the simple, non-operational semantics of language specification and the dual decidable representations of regular expressions and tries.
-Note also that proof relevance yields *parsing* (explanations) rather than mere recognition.
-We are not aware of formalized (or even rigorous) accounts of *parsing* (as opposed to mere recognition) based on non-operational specification of languages and derivatives.
-
-> I recognize that a note on the versions of Agda and its standard library would help the reader.
-
-Will do. Thank you for the suggestion!
-
-## Review B
-
-> It seems that in Fig. 1 "a language over an alphabet A" is not A^* but A^* -> Set l. This is at least nonstandard --- a standard definition should be A*. Explain the reason.
-
-`A^*` is the set of *all* strings over an alphabet `A`, i.e., the "universal language".
-Other languages are proper subsets, which are neatly formalized in type theory as `A^* -> Set l`, i.e., a function mapping each string `w` to the type of proofs that `w` is in the language.
-
-> By the way, can you use Prop instead of Set l? I suspect the former is more intuitive, especially for those who are not experts on type theory.
-
-`Set` appears to be much more widely used in the Agda ecosystem than `Prop` and so using `Set` allows reusing many more existing definitions and proofs, including type isomorphisms.
-More fundamentally, `Prop` is proof-irrelevant, so we would only get language recognition.
-In contrast, using `Set` gives proof-relevant recognition, i.e., parsing.
-
-> In Fig. 1, while I believe c :: A, I could not figure out the type of scalar, s.
-
-In the definition of scalar multiplication (`(s · P) w = s × P w`), `s` appears as an argument to the type product operator and so is a type (i.e., has type `Set ℓ`).
-Logically, it's a proposition being conjoined with the property of the string `w` being in the language `P`.
-This operator makes for more elegant and more general formulations of derivatives of concatenated and starred languages than Brzozowski's definitions.
-The additional generality comes from proof relevance ("parsing").
-It's also fundamentally important to the algebraic setting of semimodules (generalized vector spaces), which is a natural fit to proof-relevant languages.
-
-> This paper carefully distinguished three kinds of equalities, namely propositional equality, isomorphism, and extensional isomorphism.
-However, I did not understand why that equality should be used for each occurrence.
-For example, why in Fig.
-1 is propositional equality necessary?
-
-Figure 1 uses both propositional equality and extensional isomorphism.
-We could not have used (propositional) equality for all of the lemmas, since this strong condition does not hold in all cases.
-We could have used the weaker isomorphism condition for all of the lemmas, but the equalities are more specific and thus more informative.
-Moreover, isomorphisms require some notational overhead to apply, as shown where they are invoked in figures 5 and 7.
-In fact, there might be some notational overhead even for propositional equalities, but these ones are actually *definitional* (meaning proved by normalization, as noted on line 194), making them especially convenient.
-
-The reason for using both isomorphism and extensional isomorphism is simply that the `ν` lemmas are about types/propositions (nullability) while the `δ` lemmas are about predicates (general, proof-relevant language membership).
-
-*Correction:* My answer was about Figure 3.
-As for Figure 1, the uses of propositional equality are for lists (not types or type-level predicates), to which the other mentioned equality notions do not apply.
-
-> Section 2 uses A* -> B instead of A* -> Set l.
-Perhaps, the reason is that all the definitions can be polymorphic on the result, B.
-However, this difference at least makes the meaning difficult to understand.
-
-Indeed, this generalization is important because it points out that the central notion of derivatives is much more general than its usual setting of languages (even beyond generalizing from `Set` to arbitrary semirings), and some definitions and properties hold in the more general setting.
-
-> p5. "isomorphic types or predicates are equivalently decidable":
-> I could not understand the reason.
-
-The reason (proof) is outlined in the next sentence ("One direction of the isomorphism proves \AB{B} from \AB{A}, while the other proves {\AF ¬ B} from {\AF ¬ A}") and formalized in the definitions/proofs of the `_▹_` and `_▸_` signatures/lemmas stated in between.
-As with the rest of the paper, we mostly state the lemmas & theorems (signatures), leaving the interested reader to view proofs as desired in the paper/project's Agda source code.
-
-> Figs. 4 and 5 are nearly identical to Figs. 6 and 7. While I could understand that it is because of the duality, I still wonder what is the essential differences.
-
-The essential difference is that the representation data constructors and the representation examiners trade places (the duality).
-More specifically
-
-*   For regular expressions, language building blocks (`∅`, `_∪_`, `𝟙`, `_*_`, etc) are data constructors, while `ν` and `δ` are functions that examine the data.
-*   For tries, `ν` and `δ` are codata constructors, while language building blocks are functions that examine the codata.
-
-This duality is visually highlighted in the code arrangement in figures 4 and 6 and in the syntax coloring in figures 5 and 7.
-(Line 419: "The compiler-generated syntax coloring differs to reflect the changed interpretation of the definitions.")
-
-> Usually, an inductive definition gives a datatype, whereas a coinductive one does a function (or stream).
-However, the automatic differentiation requires the concrete construction process of the function even for the coinductive case.
-Another possible difference is the potential of the infiniteness of the coinductive definition.
-However, for this particular case, I am not aware of the usefulness of infiniteness.
-
-The same pattern occurs in this paper.
-Streams are a special case of string tries (for a singleton alphabet), being infinite in the specific and more general case.
-The pattern also generalizes beyond string tries to tries over arbitrary regular algebraic data types.
-
-By "concrete construction process", do you mean that there are still (corecursively defined) examination functions?
-If so, this situation is also customary in our experience when working with coinductive structures.
-
-> On page 9.
-it is argued that a difference is a work-sharing between the function and its derivative.
-I agree with this point but do not think it to be critical.
-Existing parsing methods based on the derivatives do such work sharing.
-Moreover, automatic differentiation alone cannot share repeated calculations of derivatives when we arrive at the same state more than once.
-
-We thought this work sharing issue worth pointing out, because it makes an enormous difference for symbolic vs automatic differentiation in the usual calculus sense, and indeed is the main argument usually given in the literature for preferring the latter over the former.
-We are not aware of a clear or rigorous discussion of the issue in derivative-based parsing work (though there are many optimizations described informally and without semantic basis).
-It is not clear what the practical implications are in the setting of languages, but it seems worth mentioning to stimulate closer examination.
-If the performance difference is much less than with differential calculus, then it's important to understand why.
-For instance, maybe there is a dual technique for regular expressions that could be examined clearly for useful practical and theoretical insight.
-On the other hand, if the difference is the same or greater, then it's important to exploit the potential with better justification and perhaps implementation.
-
-We certainly agree with your remark about the potential for improved work sharing when a "state" is reached in more than one way!
-
-## Review C
-
-> The specification of language membership (language operations) is non-operational and it seems to be one of the key ideas that the authors are advocating, i.e., to have as simple as possible specifications with respect to which other operations (or implementations) can be proved correct.
-
-Yes, exactly.
-
-> The result of this approach is that the provided language membership deciders (defined using differentiation on inductive or coinductive representations of the language) are correct.
-In other words, the function that decides language membership is a theorem saying that if there is a representation for a language P, then language membership in P is decidable, i.e., for any w, the function either computes a witness of P w (the type of language membership proofs of w in P) or constructs a proof that no such exists.
-
-Indeed. Well-summarized. Thank you.
-
-> Overall, I think this is a nice, clean, well-written and compact paper.
-Working things out so that clean and simple definitions are also "magically" correct according to Agda is often a non-trivial task.
-The idea that specifications should be as simple as possible is a very reasonable one.
-Therefore, I think these ideas deserve to be communicated.
-
-Thank you!
-The clean, simple, automatically verified nature is exactly what we were wanting to highlight.
-
-> One issue I have is that reading the paper might give the impression that the choice of type theory (instead of set theory) or type-level languages is necessary for achieving non-operational specification of language membership or more informative language membership than just true/false.
->The paper does demonstrate that type theory can be convenient for achieving this.
->
-> For example, the concatenation of P and Q (subsets of A*) is a certain subset of A* 
->
-> P * Q = { w \in A* | exists u, v. w = uv \and u \in P \and v \in Q }
->
-> and the characteristic function of this set corresponds to what is given in Figure 1.
-This could be done in Agda with decidable languages (A* -> 2) and thus also achieve non-operational specification (values of A* -> 2 as type indices).
-(Assumption: a word w can be decomposed in a finite number of ways.)
-
-We wrote a comparison with set-theoretical and decidable specifications and discarded it for better focus.
-Yes, one can start with decidable languages and exploit the property that every finite list can be unappended in finitely many ways.
-Doing so would go against the principle of simple, elegant, non-operational *specifications*, since it relies on cleverness involving and motivated by decidability of language *recognition* rather than the simpler nature of language *definition*.
-
-> line 18: I think it would be helpful to later mention what properties would have required induction/coinduction?
-
-Thanks. We might refer to Abel [2016], which proves a great number of algebraic properties on tries, all of them needing coinduction/bisimulation.
-The key to eliminating these much more complex proofs is inheriting properties from the *indices* (languages specified as type-level predicates), which are simple and neither inductive nor coinductive.
-
-> Contribution 1, line 93: Is this meant to say that capturing the semantic essence of Brzozowski's derivatives is something new?
-The semantic operation was defined before the syntactic one.
-Brzozowski references three papers where (under different names) derivatives of sequences are used and then defines the corresponding operation on syntax.
-
-*Formalizing* (stating and proving in a formal system) the semantic essence is new, as far as we know.
-Generalizing this essence from languages to functions from lists might be new as well.
-Moreover, most recent work we know of seems to identify derivatives with regular expressions, leading readers away from the more natural and general notion in practice and its alternative realizations, such as tries.
-(Abel [2016] is an exception but lacks a formal semantic specification separate from the implementation itself.)
-
-See also the paragraph starting on line 717, including the mention of Brzozowski on line 720.
-I imagine you will agree that it is easier to make clear comparisons after the technical substance is presented (in conclusions) than before (in introduction); and yet readers want at least a general sense of comparison before the technical substance.
-
-> Contribution 2, line 99: The duality is between regular expressions and a certain subset of tries (those constructed using the language building operations).
-The two things named Lang in Figures 4 and 6 are not equally expressive.
-
-We agree.
-This comparison can be expressed more clearly, including the fact that regular expressions *and their derivatives* also naturally generalize to a much broader class of languages, as Might and Darais [2010] and follow-up work has noted (though without much specification or rigor).
-
-> line 103: Is "proof isomorphism" (throughout the paper) the same as "type isomorphism" on line 188?
-
-Yes.
-"Type isomorphism" is more technically specific (suiting the technical meat of the paper), while its use as "proof isomorphism" seems more suitable for the introduction and the contrast with "logical equivalence" mentioned in line 103.
-
-> Contribution 3, line 103: Is this statement considering previous work on formalisations or formal languages in general?
-As soon as something more general is considered instead of nullability (true/false empty word property), the lemmas (or definitions) for concatenation and closure start to look like those in Figure 2.
-
-The statement refers to work we are aware of in formal languages as well as its formalization. Have we missed something important?
-
-> line 179: What is meant exactly by "liberation"?
-I think it is fairly standard to consider left and right quotients (wrt.
-a word) of languages (sets of strings, no grammars) as monoid actions and thus reducing language membership to nullability also in a step-by-step manner.
-
-This sentence is in the context of the immediately preceding sentence, which refers to Brzozowski's paper ("Derivatives of *Regular Expressions*"---emphasis mine) and its extensions by Might and Darais [2010] etc.
-We are trying to encourage a shift from recent near-identification of derivatives (for parsing) as about a particular grammatical/inductive language representation back to the underlying semantic basis you mention (though also tilted toward proof relevance for parsing and other applications), so that the formal development has a simpler formal starting point (and neither inductive nor coinductive) and alternatives to grammatical representation are easier to notice.
-
-> line 283: I think it would be good to also say what is the correctness condition (in addition to the fact that it is guaranteed by types).
-
-The condition is stated on line 228 and is captured in the type of `⟦_⟧` on line 310, namely that `⟦_⟧` is both computable and converts a language into a decidable recognizer/parser.
-We will add a reminder at line 283.
-Thanks.
-
-> line 308: This left triangle looks like a map operation. Maybe a comment about why this has to be part of the syntax.
-
-The left triangle *reifies* a map operation.
-The reason it must be part of the inductive representation is the same as the other constructors, namely so that the core lemmas (figure 2) translate into an implementation.
-We will add a comment.
-
-> line 333: Maybe a comment about why f works on the left hand side of the triangle with both p and with p differentiated.
-
-Thanks.
-The reason is that both `p` and `δ p a` are languages/predicates.
-
-> Figures 4 and 5: there is no definition of the language (A* -> Set) denoted by an expression and thus no formal statement that it indeed denotes the language in its type.
-
-Thank you.
-This missing semantic function maps every `p : Lang P` to `P`, regardless of the specifics of `p` (which is why the algebraic properties transfer effortlessly from type-level languages without having to consider the structure of `p`).
-We agree that it's worth stating explicitly and will do so.
-
-> line 623: By a choice of an equivalence relation we are quotienting the (encapsulated) representation. It would be good to give a bit more intuition here (which quotient is this).
-
-Indeed.
-This comment is closely related to your remark above about the denotation of the decidable language representations and has the same answer.
-Equivalence of representations is *semantic* equivalence, which is to say equivalence of the indices (type-level languages) that capture the semantics.
-We will add this clarification.
-
-> line 723: What are transformations?
-
-The rules of differentiation, as transformations of symbolic/grammatical representations.
-
-> line 724: "based *on* the"?
-
-Fixed. Thanks!
-
-## Review D
-
-> Section 6.2 is quite dense, and its motivation is not entirely clear to me.
-Is its purpose to show that existentials can be used to force the dependently-typed definitions developed throughout the paper into the simply-typed algebraic structures?
-If so, I believe a single paragraph would to, and would perhaps be less confusing.
-Furthermore, the section title is confusing: transport properties to what?
-We haven't defined the existentially quantifier variants yet, so it's not yet obvious that we have anywhere to transport properties to.
-
-The purpose is to show that the properties of non-decidable, type-level languages transfer effortlessly to the decidable implementations.
-The importance of this fact is that one gets both (a) easy proofs of useful correctness properties (without complications due to efficiency and even decidability) *and* (b) efficient implementations.
-
-This simple solution for property transfer, however, raises a "technical obstacle" (line 581): the indexed data types do not match the types of standard algebraic properties.
-Existential wrapping is a simple, generic solution to this obstacle.
-
-I renamed the section from "Transporting Properties" to "Transporting Properties from Specification to Implementations"
-
-> Line 86: The clause "including the use of existential quantifiers" is confusing, as it in no way makes clear what the existential quantifiers are used for. I'd suggest dropping the clause entirely.
-
-I've appended "(as in Figure 1) to that sentence".
-
-> Line 500: The types of the primitive language constructs are elided throughout the paper.
-Therefore, it is a bit surprising to see that the variant of star which, ostensibly, abstracts away from lists, still references the list primitives foldr and All, as well as the monoid properties.
-
-Before generalizing, the list plays two different algebraic roles: monoid and foldable.
-Either can be generalized independently.
-
-> Line 608: Why are the existentially-wrapped versions specifically based on the inductive variant?
-Is there anything specific that makes these better or more amenable?
-The text so far led me to believe the coinductive variant was better and more performant.
-
-It was an arbitrary choice. Both implementations (inductive and coinductive) are wrapped alike. We will say so in the text.
-
-> Line 638: Is the pattern matching in the type valid Agda syntax? Again, I don't think these liftings really add anything to the paper.
-
-Yes. All such declarations and definitions in the paper (and those elided from the paper) are fully type- (and thus correctness-)checked when the paper is typeset.
-
-* * * * * * * * * * * * * * * * * * * *
+# ICFP 2021 Paper #87 final review responses
 

### 2021-10-23 paper.tex
"bisumulation" --> "bisimulation"

@@
 Dually, coinductive types (``codata'') describe infinitely large values and are processed corecursively (as greatest fixed points) by copattern-matching clauses that compose results.
-Programs on inductive types are often proved by induction, while programs on coinductive types are often proved by coinduction\out{ and specifically bisumulation}.
+Programs on inductive types are often proved by induction, while programs on coinductive types are often proved by coinduction\out{ and specifically bisimulation}.
 (As we will see in \secref{Transporting Properties from Specification to Implementations}, however, the simple relationship to the specification in \secref{Decomposing Languages} allows many trivial correctness proofs, needing neither induction or coinduction.)
